### 2023-06-29 paper/example-programming.tex
Add paper strcuture

% -*- coding: utf-8; -*-
% vim: set fileencoding=utf-8 :
\documentclass[english,submission]{programming}
%% First parameter: the language is 'english'.
%% Second parameter: use 'submission' for initial submission, remove it for camera-ready (see 5.1)

\usepackage[backend=biber]{biblatex}
\addbibresource{example.bib}

%
% Packages and Commands specific to article (see 3)
%
% These ones  are used in the guide, replace with your own.
% 
\usepackage{multicol}
\lstdefinelanguage[programming]{TeX}[AlLaTeX]{TeX}{%
  deletetexcs={title,author,bibliography},%
  deletekeywords={tabular},
  morekeywords={abstract},%
  moretexcs={chapter},%
  moretexcs=[2]{title,author,subtitle,keywords,maketitle,titlerunning,authorinfo,affiliation,authorrunning,paperdetails,acks,email},
  moretexcs=[3]{addbibresource,printbibliography,bibliography},%
}%
\lstset{%
  language={[programming]TeX},%
  keywordstyle=\firamedium,
  stringstyle=\color{RosyBrown},%
  texcsstyle=*{\color{Purple}\mdseries},%
  texcsstyle=*[2]{\color{Blue1}},%
  texcsstyle=*[3]{\color{ForestGreen}},%
  commentstyle={\color{FireBrick}},%
}

\newcommand*{\CTAN}[1]{\href{http://ctan.org/tex-archive/#1}{\nolinkurl{CTAN:#1}}}
%%

%%%%%%%%%%%%%%%%%%
%% These data MUST be filled for your submission. (see 5.3)
\paperdetails{
  %% perspective options are: art, sciencetheoretical, scienceempirical, engineering.
  %% Choose exactly the one that best describes this work. (see 2.1)
  perspective=art,
  %% State one or more areas, separated by a comma. (see 2.2)
  %% Please see list of areas in http://programming-journal.org/cfp/
  %% The list is open-ended, so use other areas if yours is/are not listed.
  area={Social Coding, General-purpose programming},
  %% You may choose the license for your paper (see 3.)
  %% License options include: cc-by (default), cc-by-nc
  % license=cc-by,
}
%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%
%% These data are provided by the editors. May be left out on submission.
%\paperdetails{
%  submitted=2016-08-10,
%  published=2016-10-11,
%  year=2016,
%  volume=1,
%  issue=1,
%  articlenumber=1,
%}
%%%%%%%%%%%%%%%%%%

\begin{document}

\title{Live probes for free}
%\subtitle{Preparing Articles for Programming}% optional
%\titlerunning{Preparing Articles for Programming} %optional, in case that the title is too long; the running title should fit into the top page column

\author[a]{Tobias Pape}
\authorinfo{is the author of this {LaTeX} class. Contact him at
  \email{tobias.pape@hpi.uni-potsdam.de}.}
\affiliation[a]{Hasso Plattner Institute, University of Potsdam, Germany}
\author{Cristina V. Lopes}
\authorinfo{is associate editor for the first two issues of The Art, Science,
  and Engineering of Programming. Contact her at \email{lopes@ics.uci.edu}.}
\affiliation{University of California, Irvine, USA}
\author[a]{Robert Hirschfeld}
\authorinfo{is chair of the AOSA steering committee. The Art, Science,
  and Engineering of Programming is published by AOSA. Contact Robert at \email{hirschfeld@hpi.uni-potsdam.de}.}

% \authorrunning{T. Pape, C. Lopes, R. Hirschfeld} % Optional, for long author lists

\keywords{programming journal, paper formatting, submission preparation} % please provide 1--5 keywords

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Please go to https://dl.acm.org/ccs/ccs.cfm and generate your Classification
% System [view CCS TeX Code] stanz and copy _all of it_ to this place.
%% From HERE
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10002944.10011122.10003459</concept_id>
<concept_desc>General and reference~Computing standards, RFCs and guidelines</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10010405.10010476.10010477</concept_id>
<concept_desc>Applied computing~Publishing</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[300]{General and reference~Computing standards, RFCs and guidelines}
\ccsdesc[500]{Applied computing~Publishing}

% To HERE
%%%%%%%%%%%%%%%%%%%%%%%

\maketitle

% Please always include the abstract.
% The abstract MUST be written according to the directives stated in 
% http://programming-journal.org/submission/
% Failure to adhere to the abstract directives may result in the paper
% being returned to the authors.
\begin{abstract}
  %What is the broad context of the work? What is the importance of the general research area?
  \emph{Context}
  % What problem or question does the paper address? How has this problem or question been addressed by others (if at all)?
  \emph{Inquiry}
  % What was done that unveiled new knowledge?
  \emph{Approach}
  % What new facts were uncovered? If the research was not results oriented, what new capabilities are enabled by the work?
  \emph{Knowledge}
  % What argument, feasibility proof, artifacts, or results and evaluation support this work?
  \emph{Grounding}
  % Why does this work matter?
  \emph{Importance}

\end{abstract}
\section{Introduction}
\label{sec:introduction}
\section{Problem Overview}
\label{sec:problem-overview}
\section{Stack Recording}
\label{sec:stack-recording}
\section{Live Probes in Java with JDI}
\label{sec:live-probes-java}
\section{Generalizing Live Probes with Debugger Adapter Protocol}
\label{sec:generalizing-live-probes}
\section{Evaluation}
\label{sec:evaluation}
\section{Related Work}
\label{sec:related-work}
\section{Conclusion}
\label{sec:conclusion}

\printbibliography

\end{document}

% Local Variables:
% TeX-engine: luatex
% End:

### 2023-06-29 LiveFromDAP/src/livefromdap/agent/CLiveAgent.py

@@ -2,7 +2,7 @@ import os
 import subprocess
 import time
 
-from livefromdap.utils.Stackframe import Stackframe, Stacktrace
+from livefromdap.utils.StackRecord import Stackframe, StackRecord
 from .BaseLiveAgent import BaseLiveAgent
 
 class CLiveAgent(BaseLiveAgent):
@@ -168,7 +168,7 @@ class CLiveAgent(BaseLiveAgent):
         frame_id = self.get_stackframes(thread_id=self.thread_id)[0]["id"]
         self.evaluate(command, frame_id=frame_id)
         self.wait("event", event="stopped")
-        stacktrace = Stacktrace()
+        stacktrace = StackRecord()
         while True:
             stop, line, variables = self.get_local_variables()
             stackframe = Stackframe(line, variables)

### 2023-06-29 LiveFromDAP/src/livefromdap/utils/StackRecord.py

@@ -24,7 +24,7 @@ class Stackframe():
             if v["name"] == name:
                 return v["value"]
 
-class Stacktrace():
+class StackRecord():
     stackframes : list
     last_stackframe : Stackframe
     

### 2023-06-29 LiveFromDAP/src/prettyprinter/CPrettyPrinter.py

@@ -1,6 +1,6 @@
 from pycparser import c_ast, parse_file
 
-from livefromdap.utils.Stackframe import Stacktrace
+from livefromdap.utils.StackRecord import StackRecord
 
 
 class CPrettyPrinter(c_ast.NodeVisitor):
@@ -12,7 +12,7 @@ class CPrettyPrinter(c_ast.NodeVisitor):
         self.changed_source = False
         # Get the number of lines in the file
     
-    def pretty_print(self, stacktrace : Stacktrace, return_value=None):
+    def pretty_print(self, stacktrace : StackRecord, return_value=None):
         self.return_value = return_value
         self.stacktrace = stacktrace
         self.ast = parse_file(self.target_path, use_cpp=True)

### 2023-07-06 paper/example-programming.tex
paper draft 1

@@
 %%%%%%%%%%%%%%%%%%
+\definecolor{codegray}{gray}{0.9}
+\newcommand{\code}[1]{\colorbox{codegray}{\texttt{#1}}}
 
@@
   \emph{Context}
+  In his presentation "Inventing on Principles", Bret Victor demonstrates a live code editor: by specifying input values for a function, we can observe in real time the values taken by the variables during execution, as the code is written.
+  This information is often obtained using a language designed for live programming or by instrumentation of a specific runtime.
   % What problem or question does the paper address? How has this problem or question been addressed by others (if at all)?
   \emph{Inquiry}
+  \dots
   % What was done that unveiled new knowledge?
   \emph{Approach}
+  In this paper we propose to exploit the capabilities of debuggers to obtain the data needed to design a live code editor.
   % What new facts were uncovered? If the research was not results oriented, what new capabilities are enabled by the work?
   \emph{Knowledge}
+  \dots
   % What argument, feasibility proof, artifacts, or results and evaluation support this work?
   \emph{Grounding}
+  \dots
   % Why does this work matter?
   \emph{Importance}
+  \dots
 
@@
 \label{sec:introduction}
+
 \section{Problem Overview}
 \label{sec:problem-overview}
-\section{Stack Recording}
+
+\section{Using the Debugger ?}
 \label{sec:stack-recording}
+
+\subsection{Stack Recording}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.8\linewidth]{img/stackrecord.png}
+  \caption{Stack Recording example}
+  \label{fig:stack-recording}
+\end{figure}
+
+In order to generate the data needed to probe a variable, we need to be able to retrieve the state of the variable during execution. 
+This information is contained in the stackframe at the time of execution. 
+However, we need to add spatial and temporal information to this: we need to associate each stackframe with the location in the code at which it was retrieved, and we also need to know the order in which these stackframes were retrieved.
+
+In this paper, we introduce a structure for representing this data: a \textit{stack recording} \ref*{fig:stack-recording}. 
+A stack trace represents the different stackframes in the form of a chain, where each stackframe has a reference to the corresponding line of source code. 
+This representation allows us to maintain a link between the spatial location (the reference to the source code) and the temporal location (the order of these stackframes) of the execution.
+This representation has several advantages: it is easy to construct from debugger information, and it applies to most programming languages.
+
+\subsection{Keep Alive Agent}
+
+\begin{figure}[h]
+  \centering
+  %\includegraphics[width=0.8\linewidth]{img/keepalive-agent.png}
+  \caption{Keep Alive Agent}
+  \label{fig:keepalive-agent}
+\end{figure}
+
+A live environment must be able to react to two different events: a change in the code or a change in the test/input data. 
+If the code changes, we need to re-execute the code, and in the case of a compiled language, we need to compile the new code first. 
+If the data changes, we need to be able to re-execute the programme with the new data. 
+In addition, it is necessary to keep compilation and execution times low enough to maintain an interactive experience with the user.
+
+To meet these constraints, we propose to use the debugger on an intermediate program, a \textit{keep-alive agent}. 
+This program keeps the debugger alive between executions and code changes to reduce initialisation times. 
+The general operation of a keep-alive agent is shown in Figure \ref*{fig:keepalive-agent} :
+\begin{itemize}
+  \item At the start of the session, the debugger is started on the agent. Once initialised, the agent is paused.
+  \item The target code is loaded into the agent and a breakpoint is set at the input of the target method.
+  \item Execution of the target method triggers the breakpoint at the input of the method, and execution continues step by step to build the \textit{stack recording}.
+\end{itemize}
+
 \section{Live Probes in Java with JDI}
 \label{sec:live-probes-java}
+
+We initially developed a Java backend using JDI, a debugging interface for Java, to implement the concepts discussed in the previous section.
+The keep-alive agent includes a method for loading classes into the JVM.
+
+When executing the target method, the arguments are created in the client JVM and then passed to the debugger JVM using the \code{mirrorOf} and \code{newInstance} methods from the reflection API.
+The target method is invoked using \code{invokeMethod}.
+To handle events from the JDI when breakpoints are set, a dedicated thread is used to prevent deadlocks.
+
+If modifications are needed in the code of the target method, we employ the \code{redefineClasses} method. 
+This method allows changing the content of a class loaded in the JVM using an array of byte codes. 
+However, it has a limitation: it only works if the class signature remains unchanged. 
+If the class structure is modified, restarting the debugger is necessary.
+
 \section{Generalizing Live Probes with Debugger Adapter Protocol}

### 2023-07-06 LiveFromDAP/src/livefromdap/target/binary_search_alt.c
Add new example and compile script

@@ -1,5 +1,0 @@
-int binary_search(int arr[], int length, int target) {
-    int left = 0;
-    int right = length - 1;
-    return 40;
-}

### 2023-07-06 LiveFromDAP/src/livefromdap/target/c/bubblesort.c

@@ -1,0 +1,12 @@
+void bubblesort(int arr[], int size){
+    int i, j, temp;
+    for (i = 0; i < size; i++){
+        for (j = 0; j < size - i - 1; j++){
+            if (arr[j] > arr[j + 1]){
+                temp = arr[j];
+                arr[j] = arr[j+1];
+                arr[j+1] = temp;
+            }
+        }
+    }
+}

### 2023-07-06 LiveFromDAP/src/livefromdap/target/c/fibonnaci.c

@@ -1,0 +1,6 @@
+int fibonnaci(int n) {
+    if (n <= 1) {
+        return n;
+    }
+    return fibonnaci(n - 1) + fibonnaci(n - 2);
+}

### 2023-07-06 LiveFromDAP/src/livefromdap/target/c/prime.c

@@ -1,0 +1,19 @@
+int is_prime(int x){
+    for (int i = 2; i <= x / 2; i++) {
+        if (x % i == 0) {
+            return 0;
+        }
+    }
+    return 1;
+}
+
+int prime_in_interval(int x, int y){
+    // Return the number of prime numbers in the interval [x, y]
+    int count = 0;
+    for (int i = x; i <= y; i++) {
+        if (is_prime(i)) {
+            count++;
+        }
+    }
+    return count;
+}

### 2023-07-06 LiveFromDAP/src/livefromdap/target/java/BubbleSort.java

@@ -1,0 +1,19 @@
+public class BubbleSort {
+    public static void bubbleSort(char[] array) throws InterruptedException {
+        int n = array.length;
+        boolean swapped = true;
+        while (swapped) {
+            swapped = false;
+            for (int i = 1; i < n; i++) {
+                char a = array[i - 1];
+                char b = array[i];
+                if (a > b) {
+                    array[i - 1] = b;
+                    array[i] = a;
+                    swapped = true;
+                }
+            }
+        }
+    }
+    
+}

### 2023-07-06 LiveFromDAP/src/livefromdap/target/java/Fibonnaci.java

@@ -1,0 +1,8 @@
+public class Fibonnaci {
+    public static int fibonnaci(int n) throws InterruptedException {
+        if (n <= 1) {
+            return n;
+        }
+        return fibonnaci(n - 1) + fibonnaci(n - 2);
+    }
+}

### 2023-07-06 LiveFromDAP/src/livefromdap/target/java/Prime.java

@@ -1,0 +1,24 @@
+public class Prime {
+    public static boolean isPrime(int n) throws InterruptedException {
+        if (n <= 1) {
+            return false;
+        }
+        for (int i = 2; i < n; i++) {
+            Thread.sleep(1000);
+            if (n % i == 0) {
+                return false;
+            }
+        }
+        return true;
+    }
+
+    public static int primeInInterval(int a, int b) throws InterruptedException {
+        int count = 0;
+        for (int i = a; i <= b; i++) {
+            if (isPrime(i)) {
+                count++;
+            }
+        }
+        return count;
+    }
+}

### 2023-07-06 LiveFromDAP/src/livefromdap/target/python/bubblesort.py

@@ -1,0 +1,7 @@
+def bubblesort(a):
+    n = len(a)
+    for i in range(n):
+        for j in range(n - i - 1):
+            if a[j] > a[j + 1]:
+                a[j], a[j + 1] = a[j + 1], a[j]
+    return a

### 2023-07-06 LiveFromDAP/src/livefromdap/target/python/fibonnaci.py

@@ -1,0 +1,7 @@
+def fibonnaci(n):
+    if n == 0:
+        return 0
+    elif n == 1:
+        return 1
+    else:
+        fibonnaci(n-1) + fibonnaci(n-2)

### 2023-07-06 LiveFromDAP/src/livefromdap/target/python/prime.py

@@ -1,0 +1,26 @@
+def is_prime(n):
+    """
+    Returns True if n is prime, False otherwise.
+    """
+    if n <= 1:
+        return False
+    elif n <= 3:
+        return True
+    elif n%2 == 0 or n%3 == 0:
+        return False
+    i = 5
+    while i*i <= n:
+        if n%i == 0 or n%(i+2) == 0:
+            return False
+        i += 6
+    return True
+
+def prime_in_interval(x,y):
+    """
+    Returns a list of prime numbers in the interval [x,y].
+    """
+    primes = []
+    for i in range(x,y+1):
+        if is_prime(i):
+            primes.append(i)
+    return primes

### 2023-07-06 LiveFromDAP/tests/test_CLiveAgent.py

@@ -10,8 +10,8 @@ agent.start_server()
 t3 = time.time()
 agent.initialize()
 t4 = time.time()
-source_path = os.path.abspath("src/livefromdap/target/binary_search.c")
-compiled_path = os.path.abspath("src/livefromdap/target/binary_search.so")
+source_path = os.path.abspath("src/livefromdap/target/c/binary_search.c")
+compiled_path = os.path.abspath("src/livefromdap/target/c/binary_search.so")
 agent.load_code(compiled_path)
 t5 = time.time()
 return_value, _ = agent.execute(source_path, "binary_search", ["{1,2,3,4,5,6}", "6", "9"])

### 2023-07-06 LiveFromDAP/tests/test_CPrettyPrinter.py

@@ -1,22 +1,0 @@
-from prettyprinter.CPrettyPrinter import CPrettyPrinter
-from livefromdap.agent.CLiveAgent import CLiveAgent
-
-def test_CPrettyPrinter():
-    agent = CLiveAgent(
-        runner_path="src/livefromdap/runner/runner.c",
-        target_path="src/livefromdap/target/binary_search.c",
-        target_method="binary_search",
-        runner_path_exec="src/livefromdap/runner/runner",
-        target_path_exec="src/livefromdap/target/binary_search.so",
-        debug=False)
-    agent.load_code()
-    return_value, stacktrace = agent.execute(["{1,2,3,4,5,6}", "6", "9"])
-    assert return_value['value'] == "-1"
-    agent.stop()
-    printer = CPrettyPrinter("src/livefromdap/target/binary_search.c","binary_search")
-    print("== Pretty print ==")
-    print(printer.pretty_print(stacktrace, return_value=return_value['value']))
-    
-
-if __name__ == '__main__':
-    test_CPrettyPrinter()

### 2023-07-06 LiveFromDAP/tests/test_JavaLiveAgent.py

@@ -9,7 +9,7 @@ agent.start_server()
 t3 = time.time()
 agent.initialize()
 t4 = time.time()
-agent.load_code("/home/jbdod/CWI/LiveProbes/LiveFromDAP/src/livefromdap/target", "BinarySearch")
+agent.load_code("/home/jbdod/CWI/LiveProbes/LiveFromDAP/src/livefromdap/target/java", "BinarySearch")
 t5 = time.time()
 agent.execute("BinarySearch", "binarySearch", ["new char[]{'a', 'b'}", "'a'"])
 t6 = time.time()

### 2023-07-06 LiveFromDAP/tests/test_PythonLiveAgent.py

@@ -8,7 +8,7 @@ agent.start_server()
 t3 = time.time()
 agent.initialize()
 t4 = time.time()
-agent.load_code("/home/jbdod/CWI/LiveProbes/LiveFromDAP/src/livefromdap/target/binary_search.py")
+agent.load_code("/home/jbdod/CWI/LiveProbes/LiveFromDAP/src/livefromdap/target/python/binary_search.py")
 t5 = time.time()
 result, _ = agent.execute("binary_search", ["[1, 2, 3, 4, 5]", "3"])
 assert int(result) == 2

### 2023-07-10 paper/example-programming.tex
Add DAP paragraph + Keep alive agent figure

@@
   \centering
-  %\includegraphics[width=0.8\linewidth]{img/keepalive-agent.png}
+  \includegraphics[width=0.8\linewidth]{img/keepalive_agent.png}
   \caption{Keep Alive Agent}
@@
 \label{sec:generalizing-live-probes}
+The Debug Adapter Protocol (DAP), developed by Microsoft, is a standard method of communicating with a programming language debugger.
+It is compatible with various editors, including VS Code, and provides a unified interface for all programming languages.
+
+By exploiting this protocol, we have created a new language parametric backend, which we have implemented for the C, Python and Java programming languages.
+These languages are chosen to cover both compiled and interpreted languages.
+This backend offers an interface common to all three languages, which includes methods for starting and initialising the debugging server, loading and reloading code in the debugger, and executing a method while performing stack recording.
+
+These methods allow us to carry out stack recording independently of the chosen language and facilitate the future implementation of live programming interfaces.
+
+However, it is essential to develop a keep-alive agent specific to each language and to adapt these different methods to the debugging server.
+These methods depend on both the implementation of the debugging server and the keep-alive agent.
+Initialization in the DAP protocol is left to the discretion of the debugging server developer, which means that the initialization parameters must be configured for each language.
+
+How the code is loaded also depends on the language.
+For interpreted languages such as Python, this presents no problem because the code can be interpreted while the agent is running.
+For Java, classpaths and classes can be added at runtime by redefining the JVM's ClassLoader in the keep-alive agent.
+For the C language, shared libraries are used to import new code and can be loaded from the agent.
+
+Finally, execution for Java and Python is similar.
+Unlike the backend, which uses JDI, calling methods directly from the debugger does not trigger breakpoints.
+To remedy this, execution must be initiated from the agent and not by a debugger command.
+To this end, the Python and Java agents have fields for referencing a method and its arguments; when this information is entered, the agent launches execution.
+
 \section{Evaluation}

### 2023-07-10 LiveFromDAP/src/livefromdap/agent/PythonLiveAgent.py
Correct python fibonnaci example + add assert on compiled file

@@ -19,7 +19,7 @@ class PythonLiveAgent(BaseLiveAgent):
             ["python", self.debugpy_adapter_path],
             stdin=subprocess.PIPE,
             stdout=subprocess.PIPE,
-            stderr=subprocess.PIPE
+            stderr=subprocess.PIPE,
         )
         self.io = JsonIOStream.from_process(self.server)
     
@@ -29,10 +29,8 @@ class PythonLiveAgent(BaseLiveAgent):
 
     def stop_server(self):
         """Kill the subprocess"""
-        print("Killing server, in pid", self.server.pid)
         self.server.kill()
         if getattr(self, "server", None) is not None:
-            print("Killing debuggee, in pid", self.debugee.pid)
             self.debugee.kill()
     
     def initialize(self):

### 2023-07-10 LiveFromDAP/src/livefromdap/target/python/fibonnaci.py

@@ -4,4 +4,4 @@ def fibonnaci(n):
     elif n == 1:
         return 1
     else:
-        fibonnaci(n-1) + fibonnaci(n-2)
+        return fibonnaci(n-1) + fibonnaci(n-2)

### 2023-07-10 LiveFromDAP/tests/test_CLiveAgent.py

@@ -9,6 +9,9 @@ def test_c_binary_search():
 
     source_path = os.path.abspath("src/livefromdap/target/c/binary_search.c")
     compiled_path = os.path.abspath("src/livefromdap/target/c/binary_search.so")
+    #Assert that the file exists
+    assert os.path.exists(source_path), "Source file does not exist"
+    assert os.path.exists(compiled_path), "Compiled file does not exist, please compile it first with the Makefile"
 
     agent.load_code(compiled_path)
 
@@ -27,6 +30,8 @@ def test_c_bubblesort():
 
     source_path = os.path.abspath("src/livefromdap/target/c/bubblesort.c")
     compiled_path = os.path.abspath("src/livefromdap/target/c/bubblesort.so")
+    assert os.path.exists(source_path), "Source file does not exist"
+    assert os.path.exists(compiled_path), "Compiled file does not exist, please compile it first with the Makefile"
 
     agent.load_code(compiled_path)
 
@@ -43,6 +48,8 @@ def test_c_fibonnaci():
 
     source_path = os.path.abspath("src/livefromdap/target/c/fibonnaci.c")
     compiled_path = os.path.abspath("src/livefromdap/target/c/fibonnaci.so")
+    assert os.path.exists(source_path), "Source file does not exist"
+    assert os.path.exists(compiled_path), "Compiled file does not exist, please compile it first with the Makefile"
 
     agent.load_code(compiled_path)
 
@@ -61,6 +68,8 @@ def test_c_prime():
 
     source_path = os.path.abspath("src/livefromdap/target/c/prime.c")
     compiled_path = os.path.abspath("src/livefromdap/target/c/prime.so")
+    assert os.path.exists(source_path), "Source file does not exist"
+    assert os.path.exists(compiled_path), "Compiled file does not exist, please compile it first with the Makefile"
 
     agent.load_code(compiled_path)
 

### 2023-07-10 LiveFromDAP/tests/test_JavaLiveAgent.py

@@ -1,3 +1,4 @@
+import os
 from livefromdap.agent.JavaLiveAgent import JavaLiveAgent
 
 def test_java_binary_search():
@@ -5,7 +6,16 @@ def test_java_binary_search():
 
     agent.start_server()
     agent.initialize()
-    agent.load_code("/home/jbdod/CWI/LiveProbes/LiveFromDAP/src/livefromdap/target/java", "BinarySearch")
+
+    class_path = os.path.abspath("src/livefromdap/target/java")
+    class_name = "BinarySearch"
+    source_path = os.join(class_path, f"{class_name}.java")
+    compiled_path = os.join(class_path, f"{class_name}.class")
+    assert os.path.exists(source_path), "Source file does not exist"
+    assert os.path.exists(compiled_path), "Compiled file does not exist, please compile it first with the Makefile"
+
+    agent.load_code(class_path, class_name)
+    
 
     return_value, _ = agent.execute("BinarySearch", "binarySearch", ["new char[]{'a', 'b'}", "'a'"])
     assert int(return_value) == 0
@@ -20,7 +30,15 @@ def test_java_bubblesort():
 
     agent.start_server()
     agent.initialize()
-    agent.load_code("/home/jbdod/CWI/LiveProbes/LiveFromDAP/src/livefromdap/target/java", "BubbleSort")
+
+    class_path = os.path.abspath("src/livefromdap/target/java")
+    class_name = "BubbleSort"
+    source_path = os.join(class_path, f"{class_name}.java")
+    compiled_path = os.join(class_path, f"{class_name}.class")
+    assert os.path.exists(source_path), "Source file does not exist"
+    assert os.path.exists(compiled_path), "Compiled file does not exist, please compile it first with the Makefile"
+
+    agent.load_code(class_path, class_name)
 
     return_value, _ = agent.execute("BubbleSort", "bubbleSort", ["new int[]{1,2,3,4,5,6}"])
     assert return_value is None
@@ -35,7 +53,14 @@ def test_java_fibonnaci():
 
     agent.start_server()
     agent.initialize()
-    agent.load_code("/home/jbdod/CWI/LiveProbes/LiveFromDAP/src/livefromdap/target/java", "Fibonnaci")
+    class_path = os.path.abspath("src/livefromdap/target/java")
+    class_name = "Fibonnaci"
+    source_path = os.join(class_path, f"{class_name}.java")
+    compiled_path = os.join(class_path, f"{class_name}.class")
[…]

### 2023-07-10 LiveFromDAP/tests/test_PythonLiveAgent.py

@@ -33,13 +33,12 @@ def test_python_fibonnaci():
     agent.start_server()
     agent.initialize()
     agent.load_code("/home/jbdod/CWI/LiveProbes/LiveFromDAP/src/livefromdap/target/python/fibonnaci.py")
-
+    print("Loaded code")
     result, _ = agent.execute("fibonnaci", ["5"])
-    print(result, type(result))
-    assert int(result) == "5"
+    assert int(result) == 5
 
     result, _ = agent.execute("fibonnaci", ["8"])
-    assert int(result) == "21"
+    assert int(result) == 21
 
     agent.stop_server()
 
@@ -58,5 +57,5 @@ def test_python_prime():
     agent.stop_server()
 
 print("Testing PythonLiveAgent")
-test_python_binary_search()
+test_python_fibonnaci()
 

### 2023-07-11 paper/example-programming.tex
Start to add ref

@@
 \label{sec:evaluation}
+\subsection{Demo : A Small C Live Programming Environment}
+\label{sec:demo-small-c}
+\subsection{Performance}
+\label{sec:performance}
 \section{Related Work}
 \label{sec:related-work}
+
+Example-Based Live Programming for Everyone\cite{10.1145/3426428.3426919} : Use GraalVM/Truffle to get live information from the code of multiple languages.
+
+Example Centric Programming\cite{10.1145/1052883.1052894} : Use BeanShell(custom JVM) to get live information. Prototype for Java in Eclipse.
+
+Usable Live Programming\cite{10.1145/2509578.2509585} : New language for live programming(Ying Yang) with incremental compilation. Live programming environment for this language.
+Use source location to relate execution and code(=> almost like stack recording that link stackframe and code location)
+
+Scalable Omniscient Debugging\cite{10.1145/1297105.1297067} : Omniscient debugging with a lot of data. Use a lot of memory to store all the data.
+In this paper they record almost everything, we only record the stackframe.
+
 \section{Conclusion}

### 2023-07-11 LiveFromDAP/src/livefromdap/__init__.py

@@ -1,2 +1,3 @@
 from .agent.CLiveAgent import CLiveAgent
 from .agent.PythonLiveAgent import PythonLiveAgent
+from .agent.JavaLiveAgent import JavaLiveAgent

### 2023-07-11 LiveFromDAP/src/livefromdap/agent/JavaLiveAgent.py

@@ -767,7 +767,7 @@ class JavaLiveAgent(BaseLiveAgent):
     def stop_server(self):
         """Stop the target program"""
         self.ls_server.kill()
-        if getattr(self, "server", None) is not None:
+        if getattr(self, "debugee", None) is not None:
             self.debugee.kill()
         self.server.close()
     

### 2023-07-17 paper/example-programming.tex
Add perf 1

@@
 \addbibresource{example.bib}
+\usepackage{tikz}
+\usepackage{listings}
 
@@
 
-\author[a]{Tobias Pape}
-\authorinfo{is the author of this {LaTeX} class. Contact him at
-  \email{tobias.pape@hpi.uni-potsdam.de}.}
-\affiliation[a]{Hasso Plattner Institute, University of Potsdam, Germany}
-\author{Cristina V. Lopes}
-\authorinfo{is associate editor for the first two issues of The Art, Science,
-  and Engineering of Programming. Contact her at \email{lopes@ics.uci.edu}.}
-\affiliation{University of California, Irvine, USA}
-\author[a]{Robert Hirschfeld}
-\authorinfo{is chair of the AOSA steering committee. The Art, Science,
-  and Engineering of Programming is published by AOSA. Contact Robert at \email{hirschfeld@hpi.uni-potsdam.de}.}
-
-% \authorrunning{T. Pape, C. Lopes, R. Hirschfeld} % Optional, for long author lists
+\author[a]{Jean-Baptiste Döderlein}
+\affiliation[a]{ENS Rennes, Bruz, France}
+\author[b]{Tijs van der Storm}
+\affiliation[b]{CWI, Amsterdam, The Netherlands}
+\author[b]{Riemer van Rozen}
 
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{img/stackrecord.png}
-  \caption{Stack Recording example}
+  \begin{minipage}{0.2\textwidth}
+    \begin{lstlisting}[language=C]
+//foo(3)
+int foo(int n) {
+  int i = 0;
+  while (i < n) {
+    i++;
+  }
+  return i;
+}
+    \end{lstlisting}
+  \end{minipage}
+  \hfill
+  \begin{minipage}{0.7\textwidth}
+    \begin{tikzpicture}[x=0.75pt,y=0.75pt,yscale=-1.1,xscale=0.95]
+
+%Shape: Rectangle [id:dp48530454252997635] 
+\draw   (11,13.3) -- (68.6,13.3) -- (68.6,34.5) -- (11,34.5) -- cycle ;
+%Shape: Rectangle [id:dp4741172814816692] 
+\draw   (11,34.5) -- (68.6,34.5) -- (68.6,89.7) -- (11,89.7) -- cycle ;
+%Shape: Rectangle [id:dp39333640528563685] 
+\draw   (83.8,13.3) -- (141.4,13.3) -- (141.4,34.5) -- (83.8,34.5) -- cycle ;
+%Shape: Rectangle [id:dp852338699133017] 
+\draw   (83.8,34.5) -- (141.4,34.5) -- (141.4,89.7) -- (83.8,89.7) -- cycle ;
+%Shape: Rectangle [id:dp7978026394194542] 
+\draw   (155.6,13.3) -- (213.2,13.3) -- (213.2,34.5) -- (155.6,34.5) -- cycle ;
+%Shape: Rectangle [id:dp7423976709217543] 
+\draw   (155.6,34.5) -- (213.2,34.5) -- (213.2,89.7) -- (155.6,89.7) -- cycle ;
+%Shape: Rectangle [id:dp7923849318907303] 
+\draw   (225.8,13.7) -- (283.4,13.7) -- (283.4,34.9) -- (225.8,34.9) -- cycle ;
+%Shape: Rectangle [id:dp8325747263531937] 
+\draw   (225.8,34.9) -- (283.4,34.9) -- (283.4,90.1) -- (225.8,90.1) -- cycle ;
+%Shape: Rectangle [id:dp3066063334990279] 
+\draw   (343,14.1) -- (400.6,14.1) -- (400.6,35.3) -- (343,35.3) -- cycle ;
+%Shape: Rectangle [id:dp9425547748588214] 
+\draw   (343,35.3) -- (400.6,35.3) -- (400.6,90.5) -- (343,90.5) -- cycle ;
+%Straight Lines [id:da5537198175803346] 
+\draw    (69.4,56.1) -- (80,56.1) ;
+\draw [shift={(83,56.1)}, rotate = 180] [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.08]  [draw opacity=0] (7.14,-3.43) -- (0,0) -- (7.14,3.43) -- cycle    ;
+%Straight Lines [id:da9992180079914922] 
+\draw    (141.4,56.1) -- (152,56.1) ;
+\draw [shift={(155,56.1)}, rotate = 180] [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.08]  [draw opacity=0] (7.14,-3.43) -- (0,0) -- (7.14,3.43) -- cycle    ;
+%Straight Lines [id:da24526356738515775] 
+\draw    (212.4,56.1) -- (223,56.1) ;
+\draw [shift={(226,56.1)}, rotate = 180] [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.08]  [draw opacity=0] (7.14,-3.43) -- (0,0) -- (7.14,3.43) -- cycle    ;
+%Straight Lines [id:da7346663349503276] 
+\draw [line width=1.5]    (284.4,56.1) -- (337.6,56.47) ;
+\draw [shift={(341.6,56.5)}, rotate = 180.4] [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.08]  [draw opacity=0] (9.29,-4.46) -- (0,0) -- (9.29,4.46) -- cycle    ;
+
+% Text Node
+\draw (15.2,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 2};
+% Text Node
+\draw (22.4,39) node [anchor=north west][inner sep=0.75pt]   [align=left] {n : 3};
+% Text Node
+\draw (88,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 3};
+% Text Node
+\draw (95.2,39) node [anchor=north west][inner sep=0.75pt]   [align=left] {n : 3};
+% Text Node
+\draw (94.8,53.2) node [anchor=north west][inner sep=0.75pt]   [align=left] {i \ : 0};
+% Text Node
+\draw (159.8,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 4};
+% Text Node
+\draw (167,39) node [anchor=north west][inner sep=0.75pt]   [align=left] {n : 3};
+% Text Node
+\draw (166.6,53.2) node [anchor=north west][inner sep=0.75pt]   [align=left] {i \ : 0};
+% Text Node
+\draw (230,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 3};
+% Text Node
+\draw (237.2,39) node [anchor=north west][inner sep=0.75pt]   [align=left] {n : 3};
+% Text Node
+\draw (236.8,53.6) node [anchor=north west][inner sep=0.75pt]   [align=left] {i \ : 1};
+% Text Node
+\draw (347.2,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 6};
+% Text Node
+\draw (354.4,39) node [anchor=north west][inner sep=0.75pt]   [align=left] {n : 3};
+% Text Node
+\draw (354,54) node [anchor=north west][inner sep=0.75pt]   [align=left] {i \ : 2};
+    \end{tikzpicture}
+  \end{minipage}
+  \caption{Stack Recording Example}
   \label{fig:stack-recording}
@@
 This information is contained in the stackframe at the time of execution. 
-However, we need to add spatial and temporal information to this: we need to associate each stackframe with the location in the code at which it was retrieved, and we also need to know the order in which these stackframes were retrieved.
+However, we need to add spatial and temporal information to this: we need to associate each stackframe state with the location in the code at which it was retrieved, and we also need to know the order in which these stackframes were retrieved.
 
-In this paper, we introduce a structure for representing this data: a \textit{stack recording} \ref*{fig:stack-recording}. 
-A stack trace represents the different stackframes in the form of a chain, where each stackframe has a reference to the corresponding line of source code. 
+In this paper, we introduce a structure for representing this data: a \textit{stack recording}. 
+A stack record represents the different states of the stack frame during the execution of a method.
+The stack record is presented as a chain of recorded stack frames, to which information has been added about the source code's location and height in the stack.
 This representation allows us to maintain a link between the spatial location (the reference to the source code) and the temporal location (the order of these stackframes) of the execution.
 This representation has several advantages: it is easy to construct from debugger information, and it applies to most programming languages.
-
+Figure \ref*{fig:stack-recording} shows an example of a stack recording.
 \subsection{Keep Alive Agent}
@@
 \label{sec:performance}
+
+In the context of a live programming environment, it is essential to have short response times after user interactions. 
+We therefore measured the performance of our approach in response to various questions:
+
+\begin{itemize}
+  \item What is the average performance during a live programming session?
+  \item What is the compilation time as a function of programme size?
+  \item How long does it take to load the code into the debugger, depending on the size of the programme?
+  \item How long does it need to run the program, depending on the number of steps to make the stack recording ?
+\end{itemize}
+
+\subsubsection{What is the average performance during a live programming session?}
+
+To measure the performance of our approach in the general context of a live programming session, we measured 4 points :
+
+\begin{itemize}
+  \item Agent initialisation time, which occurs once at the start of the session.
+  \item The compilation time, if there is a compilation stage.
+  \item The time taken to load the code into the debugger.
+  \item  Program execution time.
+\end{itemize}
+
+%In the 3 languages, we have implemented time measurement for the binary_search method (see appendix) with the array \[1,2,3,4,5,6\] looking for 9.
+%The execution was repeated 100 times for each language.
+
+\begin{table}[h]
+  \centering
+  \begin{tabular}{@{} r c c c c c c c c c @{}}
+  \toprule
+  & \multicolumn{1}{c}{One time} & \multicolumn{3}{c}{Each Iteration} \\
+
+  Language & Initialization & Compile & Load Code & Execute \\ \midrule
+  C & 1.24 & 0.067 & 0.0098 & 0.193 \\
+  Java & 5.92 & 0.57 & 0.015 & 1.28 \\
+  Python & 0.639 & 0 & 0.144 & 3.26 \\
+  \bottomrule
+  \end{tabular}
+  \caption{Average time in seconds for each step of the live programming session.}
+  \label{table:average-time}
+\end{table}
+
 \section{Related Work}

### 2023-07-17 LiveFromDAP/performance/general_performance.py

@@ -1,0 +1,141 @@
+import os
+import subprocess
+import time
+from livefromdap import CLiveAgent,PythonLiveAgent,JavaLiveAgent
+
+def execute_command(command):
+    process = subprocess.Popen(command, shell=True)
+    process.wait()
+    return process.returncode
+
+def compile_c():
+    execute_command("gcc -g -fPIC -shared -o src/livefromdap/target/c/binary_search.so src/livefromdap/target/c/binary_search.c")
+
+def compile_java():
+    execute_command("javac -g -d src/livefromdap/target/java src/livefromdap/target/java/BinarySearch.java")
+
+# C
+c_source_path = os.path.abspath("src/livefromdap/target/c/binary_search.c")
+c_compiled_path = os.path.abspath("src/livefromdap/target/c/binary_search.so")
+c_method = "binary_search"
+c_args = ["{1,2,3,4,5,6}", "6", "9"]
+
+# Python
+python_source_path = os.path.abspath("src/livefromdap/target/python/binary_search.py")
+python_method = "binary_search"
+python_args = ["[1,2,3,4,5,6]", "9"]
+
+# Java
+java_class_path = os.path.abspath("src/livefromdap/target/java")
+java_class_name = "BinarySearch"
+java_method = "binarySearch"
+java_args = ["new char[]{'a', 'b','c','d','e','f'}", "'i'"]
+
+times = {
+    "c": {
+        "initialize": 0,
+        "compile": 0,
+        "load_code": 0,
+        "execute": 0
+    },
+    "python": {
+        "initialize": 0,
+        "compile": 0,
+        "load_code": 0,
+        "execute": 0
+    },
+    "java": {
+        "initialize": 0,
+        "compile": 0,
+        "load_code": 0,
+        "execute": 0
+    }
+}
+
+# C
+print(" == C == ")
+t1 = time.time()
+agent = CLiveAgent(debug=False)
+agent.start_server()
+agent.initialize()
+t2 = time.time()
+times["c"]["initialize"] = t2 - t1
+for i in range(100):
+    print("Iteration: ", i, end="\r")
+    t1 = time.time()
+    compile_c()
+    t2 = time.time()
+    times["c"]["compile"] += t2 - t1
+    t1 = time.time()
+    agent.load_code(c_compiled_path)
+    t2 = time.time()
+    times["c"]["load_code"] += t2 - t1
+    t1 = time.time()
[…]

### 2023-07-17 LiveFromDAP/src/livefromdap/agent/BaseLiveAgent.py

@@ -59,6 +59,9 @@ class BaseLiveAgent(BaseLiveAgentInterface):
     def _handleRunInTerminal(self, output : dict):
         """Handle the runInTerminal request from DAP"""
         if output["type"] == "request" and output["command"] == "runInTerminal":
+            # if not exists, create the tmp folder
+            if not os.path.exists("tmp"):
+                os.makedirs("tmp")
 
             debuggee = subprocess.Popen(
                 output["arguments"]["args"],

### 2023-07-18 paper/example-programming.tex
Add performance plot

@@
   \item What is the average performance during a live programming session?
-  \item What is the compilation time as a function of programme size?
+  \item What is the compilation time as a function of programme size ?
   \item How long does it take to load the code into the debugger, depending on the size of the programme?
@@
 
+\subsubsection{What is the compilation time as a function of programme size ?}
+\subsubsection{How long does it take to load the code into the debugger, depending on the size of the programme?}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.8\linewidth]{img/load_code.png}
+  \caption{Load code time in seconds depending on the size of the program.}
+  \label{fig:load-code}
+\end{figure}
+
+\subsubsection{How long does it need to run the program, depending on the number of steps to make the stack recording ?}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.8\linewidth]{img/execute.png}
+  \caption{Execution time in seconds depending on the number of steps.}
+  \label{fig:execute}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.8\linewidth]{img/execute_ps.png}
+  \caption{Execution time in seconds per step.}
+  \label{fig:execute_tps}
+\end{figure}
+
 \section{Related Work}

### 2023-07-18 LiveFromDAP/performance/execute_performance.py

@@ -1,0 +1,139 @@
+import os
+import subprocess
+import time
+from livefromdap import CLiveAgent,PythonLiveAgent,JavaLiveAgent
+
+def execute_command(command):
+    process = subprocess.Popen(command, shell=True)
+    process.wait()
+    return process.returncode
+
+
+# if tmp folder does not exist, create it
+if not os.path.exists("tmp"):
+    os.makedirs("tmp")
+
+# C
+c_code_template = """
+int useless_method(int x) {{
+    int d = 0;
+    for (int i = 0; i < x; i++) {{
+        d += i;
+    }}
+    return d;
+}}
+"""
+def compile_c():
+    execute_command("gcc -g -fPIC -shared -o tmp/useless.so tmp/useless.c")
+
+with open(os.path.join("tmp", "useless.c"), "w") as f:
+    f.write(c_code_template.format())
+compile_c()
+
+c_source_path = os.path.abspath("tmp/useless.c")
+c_compiled_path = os.path.abspath("tmp/useless.so")
+c_method = "useless_method"
+
+
+# Python
+python_code_template = """
+def useless_method(x):
+    d = 0
+    for i in range(x):
+        d += i
+    return d
+"""
+
+with open(os.path.join("tmp", "useless.py"), "w") as f:
+    f.write(python_code_template.format())
+
+python_source_path = os.path.abspath("tmp/useless.py")
+python_method = "useless_method"
+
+# Java
+java_code_template = """
+public class UselessClass {{
+    public static int uselessMethod(int x) {{
+        int d = 0;
+        for (int i = 0; i < x; i++) {{
+            d += i;
+        }}
+        return d;
+    }}   
+}}
+"""
+def compile_java():
+    execute_command("javac -g -d tmp tmp/UselessClass.java")
+
+with open(os.path.join("tmp", "UselessClass.java"), "w") as f:
+    f.write(java_code_template.format())
+compile_java()
+java_class_path = os.path.abspath("tmp")
+java_class_name = "UselessClass"
+java_method = "uselessMethod"
+
+
+
+times = {
+    "c": {},
+    "python": {},
+    "java": {}
+}
+
+# C
+print(" == C == ")
+agent = CLiveAgent(debug=False)
+agent.start_server()
+agent.initialize()
+agent.load_code(c_compiled_path)
+agent.execute(c_source_path, c_method, ["3"]) # warm up
+for i in range(50):
[…]

### 2023-07-18 LiveFromDAP/performance/load_code_performance.py

@@ -1,0 +1,133 @@
+import os
+import subprocess
+import time
+from livefromdap import CLiveAgent,PythonLiveAgent,JavaLiveAgent
+
+def execute_command(command):
+    process = subprocess.Popen(command, shell=True)
+    process.wait()
+    return process.returncode
+
+
+# if tmp folder does not exist, create it
+if not os.path.exists("tmp"):
+    os.makedirs("tmp")
+
+# C
+c_code_template = """
+int useless_method(int x) {{
+    int l = x;{}
+    return x;
+}}
+"""
+def compile_c():
+    execute_command("gcc -g -fPIC -shared -o tmp/useless.so tmp/useless.c")
+def update_c_file(line):
+    with open(os.path.join("tmp", "useless.c"), "w") as f:
+        f.write(c_code_template.format("\nl++;"*line))
+    compile_c()
+
+update_c_file(0)
+c_source_path = os.path.abspath("tmp/useless.c")
+c_compiled_path = os.path.abspath("tmp/useless.so")
+c_method = "useless_method"
+
+
+# Python
+python_code_template = """
+def useless_method(x):
+    res = x{}
+    return res
+"""
+def update_python_file(line):
+    with open(os.path.join("tmp", "useless.py"), "w") as f:
+        f.write(python_code_template.format("\n    res += 1"*line))
+
+update_python_file(0)
+python_source_path = os.path.abspath("tmp/useless.py")
+python_method = "useless_method"
+
+# Java
+java_code_template = """
+public class UselessClass {{
+    public static int uselessMethod(int x) {{
+        int a = x;{}
+        return a;
+    }}   
+}}
+"""
+def compile_java():
+    execute_command("javac -g -d tmp tmp/UselessClass.java")
+
+def update_java_file(line):
+    with open(os.path.join("tmp", "UselessClass.java"), "w") as f:
+        f.write(java_code_template.format("\na++; "*line))
+    compile_java()
+
+update_java_file(0)
+java_class_path = os.path.abspath("tmp")
+java_class_name = "UselessClass"
+java_method = "uselessMethod"
+
+
+
+times = {
+    "c": [],
+    "python": [],
+    "java": []
+}
+
+# C
+print(" == C == ")
+agent = CLiveAgent(debug=False)
+agent.start_server()
+agent.initialize()
+for i in range(100):
+    print("Iteration: ", i, end="\r")
+    t1 = time.time()
[…]

### 2023-07-18 LiveFromDAP/src/webdemo/main.py

@@ -143,6 +143,13 @@ def stack(language):
     sessions[session_id] = Session(session_id, socketio, language, raw=True)
     return render_template('stackexplorer.html', language=language, session_id=session_id)
 
+@app.route('/stack2/<language>')
+def stackt(language):
+    # create a unique session id
+    session_id = str(uuid.uuid4())
+    sessions[session_id] = Session(session_id, socketio, language, raw=True)
+    return render_template('stackexplorer2.html', language=language, session_id=session_id)
+
 @socketio.on('disconnect')
 def on_disconnect(*args, **kwargs):
     if request.sid in sessions_to_sid:

### 2023-07-18 LiveFromDAP/src/webdemo/static/stack2.js

@@ -1,0 +1,85 @@
+var carousel = document.querySelectorAll('.carousel')[0];
+
+var carousel_instance = M.Carousel.init(carousel, {
+    fullWidth: true,
+    indicators: true,
+    noWrap: true,
+    onCycleTo: onCycle
+});
+
+function onCycle(data) {
+    if (carousel_instance == undefined) {
+        return;
+    }
+    let current_stack = getStack(carousel_instance.center);
+    editor.getAllMarks().forEach(function(mark) { mark.clear(); });
+    //highlight the current line
+    editor.markText({line: current_stack["pos"]["line"]-1, ch: current_stack["pos"]["column"]-1}, {line: current_stack["pos"]["line"]-1, ch: 1000}, {className: "highlight"});
+}
+
+
+
+
+var curentStackRecord = {
+    "stacktrace": []
+};
+
+function getStack(id){
+    return curentStackRecord["stacktrace"][id];
+}
+
+function displayStack() {
+    /* Need to create for each stack a slide in the carousel liek this
+    <div class="carousel-item red" href="#">
+        <h2>Stack #? line ?</h2>
+        <p>
+            <ul>
+                <li>x = 1</li>
+                <li>y = 2</li>
+            </ul>   
+        </p>
+    </div>
+    */
+    if (carousel_instance != undefined){
+        carousel_instance.destroy();
+    }
+    carousel.innerHTML = "";
+    for(var i = 0; i < curentStackRecord["stacktrace"].length; i++) {
+        var stack = getStack(i);
+        var slide = document.createElement("div");
+        slide.classList.add("carousel-item");
+        slide.setAttribute("href", "#");
+        var h2 = document.createElement("h2");
+        h2.innerHTML = "Stack #" + i + " line " + stack["pos"]["line"];
+        var p = document.createElement("p");
+        var ul = document.createElement("ul");
+        for(var j = 0; j < stack["variables"].length; j++) {
+            var li = document.createElement("li");
+            li.innerHTML = stack["variables"][j]["name"] + " = " + stack["variables"][j]["value"];
+            ul.appendChild(li);
+        }
+        p.appendChild(ul);
+        slide.appendChild(h2);
+        slide.appendChild(p);
+        carousel.appendChild(slide);
+    }
+    carousel_instance = M.Carousel.init(carousel, {
+        fullWidth: true,
+        indicators: true,
+        noWrap: true,
+        onCycleTo: onCycle
+    }); 
+}
+
[…]

### 2023-07-18 paper/example-programming.tex
Bug fix and demo pics

@@
 
-We initially developed a Java backend using JDI, a debugging interface for Java, to implement the concepts discussed in the previous section.
+We developed a Java backend using JDI, a debugging interface for Java, to implement the concepts discussed in the previous section. 
 The keep-alive agent includes a method for loading classes into the JVM.
@@
 
-However, it is essential to develop a keep-alive agent specific to each language and to adapt these different methods to the debugging server.
+The implementation for each language includes a keep-alive agent and code to communicate with the debugger and keep-alive agent to provide the interface functionality.
 These methods depend on both the implementation of the debugging server and the keep-alive agent. 
-Initialization in the DAP protocol is left to the discretion of the debugging server developer, which means that the initialization parameters must be configured for each language.
+The debug server initialisation parameters are specific to each language.
 
-How the code is loaded also depends on the language.
-For interpreted languages such as Python, this presents no problem because the code can be interpreted while the agent is running.
-For Java, classpaths and classes can be added at runtime by redefining the JVM's ClassLoader in the keep-alive agent.
-For the C language, shared libraries are used to import new code and can be loaded from the agent.
+The way in which the code is loaded also depends on the language. 
+For interpreted languages such as Python, the code can be interpreted at runtime and then loaded into the debugger's memory. 
+For Java, we have extended the ClassLoader to add and modify classpaths and classes at runtime in the keep-alive agent. 
+For C, the code is loaded into shared libraries that can be added and reloaded during execution.
 
-Finally, execution for Java and Python is similar.
-Unlike the backend, which uses JDI, calling methods directly from the debugger does not trigger breakpoints.
+Method execution for Java and Python is different to that used in C. 
+For Java and Python, calling methods directly from the debugger does not trigger breakpoints for these languages. 
 To remedy this, execution must be initiated from the agent and not by a debugger command. 
-To this end, the Python and Java agents have fields for referencing a method and its arguments; when this information is entered, the agent launches execution.
-
+To do this, the Python and Java agents have fields for referencing a method and its arguments; when this information is entered, the agent starts execution.
+For C, as is the case in the backend using JDI, the method call is made from the debugger console.
 \section{Evaluation}
@@
 \label{sec:demo-small-c}
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{img/demo/c.png}
+  \caption{Demo of the live programming environment for C.}
+  \label{fig:demo}
+\end{figure}
+
 \subsection{Performance}

### 2023-07-18 LiveFromDAP/src/livefromdap/agent/CLiveAgent.py

@@ -163,7 +163,7 @@ class CLiveAgent(BaseLiveAgent):
         recorded_stackframe = Stackframe(line, column, height, variables)
         stackrecording.add_stackframe(recorded_stackframe)
     
-    def execute(self, source_file, method, args):
+    def execute(self, source_file, method, args, max_steps=300):
         self.set_function_breakpoint([method])
         command = f"-exec call {method}({','.join(args)})"
         frame_id = self.get_stackframes(thread_id=self.main_thread_id)[0]["id"]
@@ -172,6 +172,7 @@ class CLiveAgent(BaseLiveAgent):
         end_lines = FunctionEndFinder(os.path.abspath(source_file),method).end_line
         stackrecording = StackRecording()
         self.initial_height = None
+        i = 0
         while True:
             stackframes = self.get_stackframes(thread_id=self.main_thread_id)
             if self.initial_height is None:
@@ -185,10 +186,16 @@ class CLiveAgent(BaseLiveAgent):
                 if len(variables) == 0:
                     return_value = None
                 else:
-                    return_value = variables[0]
+                    return_value = variables[0]["value"]
                 break
 
             self.add_variable(frame_id, stackframes, stackrecording)
+            i += 1
+            if i > max_steps:
+                # we need to pop the current frame
+                self.restart_server()
+                self.initialize()
+                return "Interrupted", stackrecording
             if stackframes[0]["line"] in end_lines:
                 self.evaluate("-exec fin", frame_id=frame_id)
                 self.wait("event", event="stopped")

### 2023-07-18 LiveFromDAP/src/livefromdap/agent/JavaLiveAgent.py

@@ -221,10 +221,10 @@ class JavaLiveAgent(BaseLiveAgent):
             self.add_classpath(class_path)
         frame_id = self.get_stackframes(self.thread_id)[0]["id"]
         self.evaluate(f"runner.loadClass(\"{class_name}\")", frame_id)
+        target_file = os.path.join(class_path, class_name + ".java")
         if class_name not in self.loaded_classes:
-            target_file = os.path.join(class_path, class_name + ".java")
-            self.lsp_add_document(target_file)
             self.loaded_classes[class_name] = os.path.abspath(os.path.join(class_path, class_name + ".java"))
+        self.lsp_add_document(target_file)
 
     def load_method(self, method_name):
         frame_id = self.get_stackframes(self.thread_id)[0]["id"]
@@ -280,7 +280,7 @@ class JavaLiveAgent(BaseLiveAgent):
 
         return (not self.method_loaded in scope_name), line_number, column, height, variables
 
-    def execute(self, clazz, method, args):
+    def execute(self, clazz, method, args, max_steps=100):
         """Execute a method with the given arguments"""
         if not clazz in self.loaded_classes.keys(): # error, class not loaded
             raise Exception(f"Class {clazz} not loaded")
@@ -303,10 +303,17 @@ class JavaLiveAgent(BaseLiveAgent):
         # We can now start the stack recording
         stacktrace = StackRecording()
         self.initial_height = None
+        i = 0
         while True:
             stop, line, column, height, variables = self.get_local_variables()
             stackframe = Stackframe(line, column, height, variables)
             stacktrace.add_stackframe(stackframe)
+            i += 1
+            if i > max_steps:
+                # we need to pop the current frame
+                self.restart_server()
+                self.initialize()
+                return "Interrupted", stacktrace
             if stop:
                 self.next_breakpoint()
                 self.wait("event", event="stopped")

### 2023-07-18 LiveFromDAP/src/livefromdap/agent/JavaLiveAgentPerf.py

@@ -1,73 +1,0 @@
-import os
-import subprocess
-import time
-from debugpy.common.messaging import JsonIOStream
-from debugpy.common import sockets
-
-from livefromdap.utils.StackRecording import Stackframe, StackRecording
-from .JavaLiveAgent import JavaLiveAgent
-
-class JavaLiveAgentPerf(JavaLiveAgent):
-    def __init__(self, *args, **kwargs):
-        super().__init__(*args, **kwargs)
-    
-    def get_local_variables(self):
-        stacktrace = self.get_stackframes(thread_id=self.thread_id, levels=1)
-        frame_id = stacktrace[0]["id"]
-        scope_name,line_number = stacktrace[0]["name"], stacktrace[0]["line"]
-        scope = self.get_scopes(frame_id)[0]
-        variables = self.get_variables(scope["variablesReference"])
-        return (not self.method_loaded in scope_name), line_number, variables
-
-
-    def execute(self, clazz, method, args):
-        """Execute a method with the given arguments"""
-        print(f"Executing {clazz}.{method} with args {args}")
-        t1 = time.time()
-        if not clazz in self.loaded_classes.keys(): # error, class not loaded
-            raise Exception(f"Class {clazz} not loaded")
-        self.load_method(method)
-        # Setup function breakpoints
-        breakpoint = self.get_start_line(self.loaded_classes[clazz], clazz, method) + 1
-        self.set_breakpoint(self.loaded_classes[clazz], [breakpoint])
-        frame_id = self.get_stackframes(self.thread_id)[0]["id"]
-        t4 = time.time()
-        # We need to load the arguments into the target program
-        self.evaluate(f"runner.args = new Object[{len(args)}]", frame_id)
-        for i, arg in enumerate(args):
-            if arg.startswith('{') and arg.endswith('}'):
-                raise NotImplementedError("Array need to be created, for example replace {'a', 'b'} with new char[]{'a', 'b'}")
-            self.evaluate(f"runner.args[{i}] = {arg}", frame_id)
-        t5 = time.time()
-        # we now continue
-        self.next_breakpoint()
-        self.wait("event", "stopped")
-        t6 = time.time()
-        # We can now start the stack recording
-        stacktrace = StackRecording()
-        times = {
-            "init": (t4 - t1) + (t6 - t5),
-            "load_arguments": t5 - t4,
-            "get_variables": 0,
-            "execution_step": 0
-        }
-        while True:
-            t7 = time.time()
[…]

### 2023-07-18 LiveFromDAP/src/prettyprinter/CPrettyPrinter.py

@@ -50,10 +50,12 @@ class CPrettyPrinter(c_ast.NodeVisitor):
             # We need to get the correct stackframe
             # we need to add at the decl position in out something like:
             # params1= ...; params2= ...; params3= ...; -> return_value= ...
-            string_output = ""
+            string_output = f"{node.type.declname}("
             for param in params_names:
-                string_output += f"{param}={self.get_variable(node.type.coord.line+1,param)[0]}; "
-            string_output += f"-> return_value={self.return_value}"
+                string_output += f"{param}={self.get_variable(node.type.coord.line+1,param)[0]}, "
+            string_output = string_output[:-2]
+            string_output += ")"
+            string_output += f"->{self.return_value}"
             self.output[node.type.coord.line-1] = string_output
     
     def visit_FuncDef(self, node):
@@ -68,46 +70,38 @@ class CPrettyPrinter(c_ast.NodeVisitor):
         output_string = f"{node.name}= "
         for stackframe in self.stacktrace.get_stackframes_line(node.coord.line):
             if (value:=self.get_value_string(stackframe.successor, node.name)) is not None:
-                output_string += f"{value} | "
+                output_string += f"{value}, "
         self.output[node.coord.line-1] = output_string[:-2]
 
     def visit_Assignment(self, node):
         output_string = f"{node.lvalue.name}= "
         for stackframe in self.stacktrace.get_stackframes_line(node.coord.line):
             if (value:=self.get_value_string(stackframe.successor, node.lvalue.name)) is not None:
-                output_string += f"{value} | "
+                output_string += f"{value}, "
         if output_string != f"{node.lvalue.name}= ":
             self.output[node.coord.line-1] = output_string[:-2]
 
     def visit_While(self, node):
         cond = node.cond
         # check if left is a variable
+        output_while = ""
         if isinstance(cond.left, c_ast.ID):
             output_string = f"{cond.left.name}= "
             for stackframe in self.stacktrace.get_stackframes_line(node.coord.line):
                 if (value:=stackframe.get_variable(cond.left.name)) is not None:
-                    output_string += f"{value} | "
+                    output_string += f"{value}, "
             if output_string != f"{cond.left.name}= ":
-                self.output[node.coord.line-1] = output_string[:-2]
+                output_while += output_string[:-2]
+                
         # check if right is a variable
         if isinstance(cond.right, c_ast.ID):
             output_string = f"{cond.right.name}= "
[…]

### 2023-07-18 LiveFromDAP/src/prettyprinter/JavaPrettyPrinter.py

@@ -41,13 +41,14 @@ class JavaPrettyPrinter(JavaVisitor):
 
     def visit_MethodDeclaration(self, node):
         if self.method_name is None or node.name == self.method_name:
-            method_output = ""
+            method_output = f"{node.name}("
             first_stackframe = self.stacktrace.stackframes[0]
             for parameter in node.parameters:
                 value = first_stackframe.get_variable(parameter.name)
-                method_output += f"{parameter.name} = {value}, "
+                method_output += f"{parameter.name}={value}, "
             method_output = method_output[:-2]
-            method_output += " -> "
+            method_output += ")"
+            method_output += "->"
             if self.return_value is not None:
                 method_output += self.return_value
             self.output[node.position.line-1] = method_output

### 2023-07-18 LiveFromDAP/src/prettyprinter/PythonPrettyPrinter.py

@@ -77,9 +77,9 @@ class PythonPrettyPrinter(ast.NodeVisitor):
             if isinstance(var, ast.Name):
                 value = self.get_variable(var.lineno, var.id)
                 if len(value) != 0:
-                    while_output += f"{var.id} = {value}, "
+                    while_output += f"{var.id} = {value} | "
         if len(while_output) != 0:
-            while_output = while_output[:-2]
+            while_output = while_output[:-3]
             self.output[node.lineno-1] = while_output
         self.generic_visit(node)
                 

### 2023-07-18 LiveFromDAP/src/webdemo/AutoAgent.py

@@ -48,17 +48,6 @@ class AutoLiveAgent(ABC):
         """Restart the debuggee"""
         pass
 
-# @buzy decorator that check if the agent is buzy, and if not, execute the function in buzy mode
-def buzy(func):
-    def wrapper(self, *args, **kwargs):
-        if self.buzy:
-            return None
-        self.buzy = True
-        res = func(self, *args, **kwargs)
-        self.buzy = False
-        return res
-    return wrapper
-
 
 class AutoCLiveAgent(AutoLiveAgent):
     
@@ -76,7 +65,7 @@ class AutoCLiveAgent(AutoLiveAgent):
         self.source_path = os.path.abspath("src/webdemo/tmp/tmp.c")
         with open(self.source_path, "w") as f:
             f.write("")
-        self.compiled_path = os.path.abspath("src/webdemo/tmp/libtmp.so")
+        self.compiled_path = os.path.abspath("src/webdemo/tmp/tmp.so")
 
     def restart(self):
         self.agent.stop_server()
@@ -125,45 +114,35 @@ class AutoCLiveAgent(AutoLiveAgent):
         if self.raw:
             print("Sending raw output")
             return json.dumps({
-                "return_value": return_value["value"],
+                "return_value": return_value,
                 "stacktrace": stacktrace.to_json()
             })
         else:
             printer = CPrettyPrinter(self.source_path,method)
-            output = printer.pretty_print(stacktrace, return_value=return_value['value'])
+            output = printer.pretty_print(stacktrace, return_value=return_value)
             return output
 
     def execute(self, method, args):
-        thread = ThreadWithReturnValue(target=self.agent.execute, args=(self.source_path, method, args,))
-        thread.start()
-        # Wait for the thread to finish, with a timeout of 3 seconds
-        output = thread.join(10)
-        # If thread is still alive, it means it has timed out
-        if thread.is_alive():
-            raise TimeoutError
+        output = self.agent.execute(self.source_path, method, args)
+        if output[0] == "Interrupted":
+            self.agent.load_code(self.compiled_path)
         # Get the output of the thread
         # Save the json result in a file
-        return self.construct_result_json(method, output)
-    
-
-        
+        return self.construct_result_json(method, output)    
 
 class AutoJavaLiveAgent(AutoLiveAgent):
     
     def __init__(self, raw=False):
-        self.buzy = True
         self.raw = raw
         self.agent = JavaLiveAgent(debug=False)
[…]

### 2023-07-18 LiveFromDAP/src/webdemo/main.py

@@ -1,4 +1,5 @@
 from queue import Queue
+import re
 from threading import Thread
 import time
 import uuid
@@ -45,9 +46,11 @@ def extract_exec_request(code, language="python"):
             exec_request = line[len(prefix):].strip()
             if "(" in exec_request and exec_request.endswith(")"):
                 method = exec_request.split("(")[0]
-                args = exec_request.split("(")[1][:-1].split(",")
+                args_str = exec_request.split("(")[1][:-1]
+                # magix regex to split the args
+                args = re.split(r',(?![^\[\]\(\)\{\}]*[\]\)\}])', args_str)
                 if not "" in map(lambda x: x.strip(), args):
-                    return method, args
+                    return method, list(map(lambda x: x.strip(), args))
     return None
 
 class Session():
@@ -79,8 +82,10 @@ class Session():
             # If the request is None, it means the session is closed
             if request is None:
                 break
-            # If the request is not None, execute it
-            self.handle_request(request)
+            try:
+                self.handle_request(request)
+            except Exception as e:
+                print(e)
             # Notify the queue that the request is done
             self.queue.task_done()
 

### 2023-07-18 LiveFromDAP/src/webdemo/static/bret.js

@@ -4,7 +4,7 @@ var outputArea = document.getElementById('output');
 var output = CodeMirror.fromTextArea(outputArea, {
     lineNumbers: true,
     indentUnit: 4,
-    readOnly: true
+    readOnly: false
 });
 
 var oldText = output.getValue();

### 2023-07-19 paper/example-programming.tex
fix perf

@@
 %Straight Lines [id:da7346663349503276] 
-\draw [line width=1.5]    (284.4,56.1) -- (337.6,56.47) ;
+\draw [line width=1.25, dashed]    (284.4,56.1) -- (337.6,56.47) ;
 \draw [shift={(341.6,56.5)}, rotate = 180.4] [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.08]  [draw opacity=0] (9.29,-4.46) -- (0,0) -- (9.29,4.46) -- cycle    ;
@@
 % Text Node
-\draw (15.2,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 2};
+\draw (15.2,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 3};
 % Text Node
@@
 % Text Node
-\draw (88,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 3};
+\draw (88,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 4};
 % Text Node
@@
 % Text Node
-\draw (159.8,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 4};
+\draw (159.8,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 5};
 % Text Node
@@
 % Text Node
-\draw (230,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 3};
+\draw (230,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 4};
 % Text Node
@@
 % Text Node
-\draw (347.2,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 6};
+\draw (347.2,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 7};
 % Text Node
@@
   \item What is the average performance during a live programming session?
-  \item What is the compilation time as a function of programme size ?
   \item How long does it take to load the code into the debugger, depending on the size of the programme?
-  \item How long does it need to run the program, depending on the number of steps to make the stack recording ?
+  \item How long does it take to run the program, depending on the number of stack frames needed to make the stack recording?
 \end{itemize}
@@
 \subsubsection{What is the average performance during a live programming session?}
-
-To measure the performance of our approach in the general context of a live programming session, we measured 4 points :
-
-\begin{itemize}
-  \item Agent initialisation time, which occurs once at the start of the session.
-  \item The compilation time, if there is a compilation stage.
-  \item The time taken to load the code into the debugger.
-  \item Program execution time.
-\end{itemize}
-
-%In the 3 languages, we have implemented time measurement for the binary_search method (see appendix) with the array \[1,2,3,4,5,6\] looking for 9.
-%The execution was repeated 100 times for each language.
 
@@
 
-\subsubsection{What is the compilation time as a function of programme size ?}
+To evaluate the performance of our approach in the general context of a live programming session, we measured the times taken by the different stages of the interaction loop:
+
+\begin{itemize}
+  \item Agent initialisation time, which occurs once at the start of the session.
+  \item The compilation time, if there is a compilation stage.
+  \item The time taken to load the code into the debugger.
+  \item Program execution time.
+\end{itemize}
+
+For the 3 languages we have implemented, we have performed the measurements for the execution of a binary search function\ref*{fig:binary-search}.
+For each language, the agent was initialized, then the code was compiled, loaded into the debugger and executed (with the array [1,2,3,4,5,6] and target 9) 100 times.
+Each execution generate a stack recording of approximately 20 stack frames.
+The results are shown in table \ref*{table:average-time}.
+
 \subsubsection{How long does it take to load the code into the debugger, depending on the size of the programme?}
@@
 
+\newpage
+\section*{Appendix}
+\label{sec:appendix}
+
+\begin{figure}[h]
+  \centering
+    \begin{lstlisting}[language=C]
+int binary_search(int arr[], int length, int target) {
+  int left = 0;
+  int right = length - 1;
+  while (left <= right) {
+
+      int mid = (left + right) / 2;
+      if (arr[mid] == target) {
+          return mid;
+      } else if (arr[mid] < target) {
+          left = mid + 1;
+      } else {
+          right = mid - 1;
+      }
+  }
+  return 0;
+}
+    \end{lstlisting}
+
+    \begin{lstlisting}[language=Python]
+def binary_search(arr, target):
+    left = 0
+    right = len(arr) - 1
+
+    while left <= right:
+        mid = (left + right) // 2
+
+        if arr[mid] == target:
+            return mid
+        elif arr[mid] < target:
+            left = mid + 1
+        else:
+            right = mid - 1
+
+    return 0 
+    \end{lstlisting}
+
+    \begin{lstlisting}[language=Java]
+public class BinarySearch {
+    public static int binarySearch(int[] array, int key){
+        int low = 0;
+        int high = array.length - 1;
+        while (low <= high) {
+            int mid = (low + high) / 2;
+            int value = array[mid];
+            if (value < key) {
+                low = mid + 1;
+            } else if (value > key) {
+                high = mid - 1;
+            } else {
+                return mid;
+            }
+        }
+        return 0;
+    }
+}
+      \end{lstlisting}
+  \caption{Binary Search in C, Python and Java}
+  \label{fig:binary-search}
+\end{figure}
+
 \end{document}

### 2023-07-19 LiveFromDAP/performance/general_performance.py

@@ -8,9 +8,8 @@ def execute_command(command):
     process.wait()
     return process.returncode
 
-
-def compile_java():
-    execute_command("javac -g -d src/livefromdap/target/java src/livefromdap/target/java/BinarySearch.java")
+if not os.path.exists("tmp"):
+    os.makedirs("tmp")
 
 # C
 c_code_template = """

### 2023-07-19 LiveFromDAP/src/livefromdap/agent/BaseLiveAgent.py
Lint and docstring

@@ -1,7 +1,5 @@
 import subprocess
-from debugpy.common.messaging import JsonIOStream
 import os
-import shutil
 from abc import ABC, abstractmethod
 
 from livefromdap.utils.StackRecording import StackRecording
@@ -87,7 +85,12 @@ class BaseLiveAgent(BaseLiveAgentInterface):
         return False
     
     def set_breakpoint(self, path : str, lines : list):
-        """Set a breakpoint in the debuggee"""
+        """Set breakpoints in the debuggee for the given path and lines
+
+        Args:
+            path (str): path of the source file
+            lines (list): list of lines to set breakpoints
+        """
         breakpoint_request = {
             "seq": self.new_seq(),
             "type": "request",
@@ -109,7 +112,11 @@ class BaseLiveAgent(BaseLiveAgentInterface):
         self.io.write_json(breakpoint_request)
 
     def set_function_breakpoint(self, names : list):
-        """Set a breakpoint in the debuggee"""
+        """Set breakpoints in the debuggee for the given function names
+
+        Args:
+            names (list): list of function names to set breakpoints
+        """
         breakpoint_request = {
             "seq": self.new_seq(),
             "type": "request",
@@ -125,6 +132,8 @@ class BaseLiveAgent(BaseLiveAgentInterface):
         self.io.write_json(breakpoint_request)
 
     def configuration_done(self):
+        """Send the configurationDone request to the debuggee
+        """
         complete_request = {
             "seq": self.new_seq(),
             "type": "request",
@@ -132,7 +141,16 @@ class BaseLiveAgent(BaseLiveAgentInterface):
         }
         self.io.write_json(complete_request)
     
-    def get_stackframes(self, thread_id : int = 1, levels : int = 100):
+    def get_stackframes(self, thread_id : int = 1, levels : int = 100) -> list:
+        """Get the stackframes from the debuggee
+
+        Args:
+            thread_id (int, optional): Defaults to 1.
+            levels (int, optional): Number of stackframes to get. Defaults to 100.
+
+        Returns:
+            list: list of stackframes
+        """
         stackframe_request = {
             "seq": self.new_seq(),
             "type": "request",
@@ -148,6 +166,11 @@ class BaseLiveAgent(BaseLiveAgentInterface):
         return output["body"]["stackFrames"]
             
     def next_breakpoint(self, thread_id : int = 1):
+        """Send the next request to the debuggee
+
+        Args:
+            thread_id (int, optional): Defaults to 1.
[…]

### 2023-07-19 LiveFromDAP/src/livefromdap/agent/CLiveAgent.py

@@ -148,8 +148,8 @@ class CLiveAgent(BaseLiveAgent):
         frame_id = self.get_stackframes(thread_id=self.main_thread_id)[0]["id"]
         libname = os.path.basename(path)
         if self.current_loaded_shared_libraries is not None:
-            self.evaluate(f"-exec call close_lib()", frame_id=frame_id)
-        self.evaluate(f"-exec call load_lib(\"{path}\")", frame_id=frame_id)
+            self.evaluate(f"-exec call close_lib()", frame_id)
+        self.evaluate(f"-exec call load_lib(\"{path}\")", frame_id)
         self.current_loaded_shared_libraries = libname
             
 
@@ -167,7 +167,7 @@ class CLiveAgent(BaseLiveAgent):
         self.set_function_breakpoint([method])
         command = f"-exec call {method}({','.join(args)})"
         frame_id = self.get_stackframes(thread_id=self.main_thread_id)[0]["id"]
-        self.evaluate(command, frame_id=frame_id)
+        self.evaluate(command, frame_id)
         self.wait("event", event="stopped")
         end_lines = FunctionEndFinder(os.path.abspath(source_file),method).end_line
         stackrecording = StackRecording()
@@ -197,7 +197,7 @@ class CLiveAgent(BaseLiveAgent):
                 self.initialize()
                 return "Interrupted", stackrecording
             if stackframes[0]["line"] in end_lines:
-                self.evaluate("-exec fin", frame_id=frame_id)
+                self.evaluate("-exec fin", frame_id)
                 self.wait("event", event="stopped")
             else:
                 self.step(thread_id=self.main_thread_id)

### 2023-07-19 LiveFromDAP/src/livefromdap/agent/PythonLiveAgent.py

@@ -1,10 +1,12 @@
 import os
 import subprocess
 
-from livefromdap.utils.StackRecording import StackRecording, Stackframe
-from .BaseLiveAgent import BaseLiveAgent
 import debugpy
 from debugpy.common.messaging import JsonIOStream
+from livefromdap.utils.StackRecording import Stackframe, StackRecording
+
+from .BaseLiveAgent import BaseLiveAgent
+
 
 class PythonLiveAgent(BaseLiveAgent):
     """Communicate with the debugpy adapter to get stackframes of the execution of a method"""
@@ -106,7 +108,7 @@ class PythonLiveAgent(BaseLiveAgent):
     def load_code(self, path: str):
         stacktrace = self.get_stackframes()
         frameId = stacktrace[0]["id"]
-        self.evaluate(f"set_import('{os.path.abspath(path)}')", frame_id=frameId)
+        self.evaluate(f"set_import('{os.path.abspath(path)}')", frameId)
         self.next_breakpoint()
         self.wait("event", "stopped")
             
@@ -114,7 +116,7 @@ class PythonLiveAgent(BaseLiveAgent):
         self.set_function_breakpoint([method])
         stacktrace = self.get_stackframes()
         frameId = stacktrace[0]["id"]
-        self.evaluate(f"set_method('{method}',[{','.join(args)}])", frame_id=frameId)
+        self.evaluate(f"set_method('{method}',[{','.join(args)}])", frameId)
         # We need to run the debug agent loop until we are on a breakpoint in the target method
         stackrecording = StackRecording()
         while True:

### 2023-07-24 paper/paper.tex
Rename paper
New plots style + Scenario

--- 2023-07-19 paper/example-programming.tex
@@
 \usepackage[backend=biber]{biblatex}
-\addbibresource{example.bib}
+\addbibresource{reference.bib}
 \usepackage{tikz}
 \usepackage{listings}
+\usepackage{graphicx}
+\usepackage{subcaption} % For subfigures
 
@@
 This representation has several advantages: it is easy to construct from debugger information, and it applies to most programming languages.
-Figure \ref*{fig:stack-recording} shows an example of a stack recording.
+Figure \ref{fig:stack-recording} shows an example of a stack recording.
+On the left is a C function and on the right is the stack recording of the execution of \code{foo(3)}.
+Each rectangle represents the state of the stackframe during execution, with the source code location (here simplified to the line number).
 \subsection{Keep Alive Agent}
@@
 This program keeps the debugger alive between executions and code changes to reduce initialisation times. 
-The general operation of a keep-alive agent is shown in Figure \ref*{fig:keepalive-agent} :
+The general operation of a keep-alive agent is shown in Figure \ref{fig:keepalive-agent} :
+% TODO : make figure clean and add better description here
 \begin{itemize}
@@
 For C, as is the case in the backend using JDI, the method call is made from the debugger console.
+
 \section{Evaluation}
 \label{sec:evaluation}
-\subsection{Demo : A Small C Live Programming Environment}
+\subsection{Demo : A Minimal Live Programming Environment}
 \label{sec:demo-small-c}
-\begin{figure}[h]
+\begin{figure}[htbp]
   \centering
@@
 \end{figure}
+
+We have developed a dynamic programming environment that incorporates live probes through Java, C, and Python debug servers. 
+The user interface consists of a web page that interacts with a local server. 
+In Figure \ref{fig:demo}, we present a screenshot depicting a session with a binary\_search function in C. 
+On the left-hand side, there is a code editor (based on CodeMirror 5), while on the right, a panel showcases live probes.
+
+Whenever changes are made in the code editor, the code is sent to the server, which then checks for parseability and syntax errors. 
+Additionally, if the code contains a comment starting with "@", followed by a function call, the server attempts to create a stack recording of that function, along with the provided parameters. 
+This stack recording is subsequently processed through a parser to generate live probes, offering the following functionalities:
+
+\begin{itemize}
+  \item Displaying input arguments and return values for each function call.
+  \item Showing the value of each variable definition or assignment. In cases where the assignment is within a loop or called multiple times, all the different values are displayed.
+  \item Presenting the comparison values for each loop (for and while).
+\end{itemize}
 
@@
 \begin{itemize}
+  \item How long does it take to compile and to load the code into the debugger, depending on the size of the programme?
+  \item How long does it take to run the program, depending on the number of stack frames needed to make the stack recording?
   \item What is the average performance during a live programming session?
-  \item How long does it take to load the code into the debugger, depending on the size of the programme?
-  \item How long does it take to run the program, depending on the number of stack frames needed to make the stack recording?
 \end{itemize}
+
+\subsubsection{How long does it take to compile and to load the code into the debugger, depending on the size of the programme?}
+
+\begin{figure}[htbp]
+  \centering
+  \begin{subfigure}[b]{0.48\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/compile_code.png}
+      \caption{\centering Compile code time}
+      \label{subfig:compile}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.48\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/load_code.png}
+      \caption{\centering Load code time}
+      \label{subfig:load}
+  \end{subfigure}
+  \caption{Compile and load code time depending on the number of lines of code.}
+  \label{fig:compileload}
+\end{figure}
+
+In our evaluation, we assessed the time required to compile and load code into the debugger memory for Python, C, and Java. 
+To accomplish this, we compiled and loaded programs ranging from 5 to 100 lines of code. 
+The summarized results can be found in Figure \ref{fig:compileload}.
+
+In Sub-figure \ref{subfig:compile}, we present the compilation time for C and Java based on the number of lines of code. 
+The compilation process was executed from the command line using javac and gcc. 
+The compilation times remain nearly constant, regardless of the number of lines of code, with an average of 34.6 ms for Java and 8.5 ms for C.
+
+Moving to Sub-figure \ref{subfig:load}, we depict the loading time in the debugger for C, Java, and Python as a function of the number of lines of code. 
+The data shows that the loading time remains constant concerning the number of lines of code. 
+On average, Python takes 3 ms, Java takes 2.8 ms, and C takes 1 ms for loading.
+
+\subsubsection{How long does it need to run the program, depending on the number of steps to make the stack recording ?}
+
+\begin{figure}[htbp]
+  \centering
+  \begin{subfigure}[b]{0.45\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/execute.png}
+      \caption{\centering Total execution time}
+      \label{subfig:execute-time-total}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.45\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/execute_ps.png}
+      \caption{\centering Execution time per step}
+      \label{subfig:execute-time-ps}
+  \end{subfigure}
+  \caption{Execution time in seconds depending on the number of steps.}
+  \label{fig:execute-time}
+\end{figure}
+
+Subsequently, we proceeded to evaluate the execution time concerning the number of steps in the debugger. 
+Figure \ref{subfig:execute-time-total} presents the total time taken based on the number of steps during stack recording, while Figure \ref{subfig:execute-time-ps} demonstrates the time per step as a function of the number of steps.
+Notably, the execution time exhibits a linear relationship with the number of steps in the method.
 
@@
 
+\begin{figure}[htbp]
+  \centering
+  \begin{subfigure}[b]{0.3\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/scenario_bs_c.png}
+      \caption{\centering C}
+      \label{subfig:scenario-bs-c}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.3\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/scenario_bs_java.png}
+      \caption{\centering Java}
+      \label{subfig:scenario-bs-java}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.3\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{img/scenario_bs_python.png}
+    \caption{\centering Python}
+    \label{subfig:scenario-bs-py}
+  \end{subfigure}
+
+  \caption{Execution time and Stackrecording size for binary search scenario.}
+  \label{fig:scenario-bs}
+\end{figure}
+
 To evaluate the performance of our approach in the general context of a live programming session, we measured the times taken by the different stages of the interaction loop:
@@
 
-For the 3 languages we have implemented, we have performed the measurements for the execution of a binary search function\ref*{fig:binary-search}.
+For the 3 languages we have implemented, we have performed the measurements for the execution of a binary search function \ref{fig:binary-search}.
 For each language, the agent was initialized, then the code was compiled, loaded into the debugger and executed (with the array [1,2,3,4,5,6] and target 9) 100 times.
 Each execution generate a stack recording of approximately 20 stack frames.
-The results are shown in table \ref*{table:average-time}.
-
-\subsubsection{How long does it take to load the code into the debugger, depending on the size of the programme?}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=0.8\linewidth]{img/load_code.png}
-  \caption{Load code time in seconds depending on the size of the program.}
-  \label{fig:load-code}
-\end{figure}
-
-\subsubsection{How long does it need to run the program, depending on the number of steps to make the stack recording ?}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=0.8\linewidth]{img/execute.png}
-  \caption{Execution time in seconds depending on the number of steps.}
-  \label{fig:execute}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=0.8\linewidth]{img/execute_ps.png}
-  \caption{Execution time in seconds per step.}
-  \label{fig:execute_tps}
-\end{figure}
+The results are shown in table \ref{table:average-time}.
 

### 2023-07-24 LiveFromDAP/performance/execute_performance.py

@@ -52,8 +52,8 @@ python_method = "useless_method"
 
 # Java
 java_code_template = """
-public class UselessClass {{
-    public static int uselessMethod(int x) {{
+public class ExecutePerformance {{
+    public static int executePerformanceMethod(int x) {{
         int d = 0;
         for (int i = 0; i < x; i++) {{
             d += i;
@@ -63,14 +63,14 @@ public class UselessClass {{
 }}
 """
 def compile_java():
-    execute_command("javac -g -d tmp tmp/UselessClass.java")
+    execute_command("javac -g -d tmp tmp/ExecutePerformance.java")
 
-with open(os.path.join("tmp", "UselessClass.java"), "w") as f:
+with open(os.path.join("tmp", "ExecutePerformance.java"), "w") as f:
     f.write(java_code_template.format())
 compile_java()
 java_class_path = os.path.abspath("tmp")
-java_class_name = "UselessClass"
-java_method = "uselessMethod"
+java_class_name = "ExecutePerformance"
+java_method = "executePerformanceMethod"
 
 
 
@@ -86,11 +86,15 @@ agent = CLiveAgent(debug=False)
 agent.start_server()
 agent.initialize()
 agent.load_code(c_compiled_path)
-agent.execute(c_source_path, c_method, ["3"]) # warm up
+
+# warm up
+for _ in range(5):
+    agent.execute(c_source_path, c_method, ["3"], max_steps=1000000)
+
 for i in range(50):
     print("Iteration: ", i, end="\r")
     t1 = time.time()
-    _, st = agent.execute(c_source_path, c_method, [str(i+1)])
+    _, st = agent.execute(c_source_path, c_method, [str(i+1)], max_steps=1000000)
     t2 = time.time()
     times["c"][len(st.stackframes)] = t2 - t1
 agent.stop_server()
@@ -102,7 +106,11 @@ agent = PythonLiveAgent(debug=False)
 agent.start_server()
 agent.initialize()
 agent.load_code(python_source_path)
-agent.execute(python_method, ["3"], max_steps=1000000) # warm up
+
+# warm up
+for _ in range(5):
+    agent.execute(python_method, ["3"], max_steps=1000000)
+    
 for i in range(50):
     print("Iteration: ", i, end="\r")
     t1 = time.time()
@@ -120,11 +128,15 @@ agent = JavaLiveAgent(debug=False)
 agent.start_server()
 agent.initialize()
[…]

### 2023-07-24 LiveFromDAP/performance/general_performance.py

@@ -142,7 +142,7 @@ for i in range(100):
     t2 = time.time()
     times["c"]["load_code"] += t2 - t1
     t1 = time.time()
-    agent.execute(c_source_path, c_method, c_args)
+    agent.execute(c_source_path, c_method, c_args, max_steps=1000000)
     t2 = time.time()
     times["c"]["execute"] += t2 - t1
 agent.stop_server()
@@ -167,7 +167,7 @@ for i in range(100):
     t2 = time.time()
     times["python"]["load_code"] += t2 - t1
     t1 = time.time()
-    agent.execute(python_method, python_args)
+    agent.execute(python_method, python_args, max_steps=1000000)
     t2 = time.time()
     times["python"]["execute"] += t2 - t1
 agent.stop_server()
@@ -195,7 +195,7 @@ for i in range(100):
     t2 = time.time()
     times["java"]["load_code"] += t2 - t1
     t1 = time.time()
-    agent.execute(java_class_name, java_method, java_args)
+    agent.execute(java_class_name, java_method, java_args, max_steps=1000000)
     t2 = time.time()
     times["java"]["execute"] += t2 - t1
 agent.stop_server()
@@ -208,6 +208,6 @@ times["java"]["execute"] /= 100
 # convert to dataframe
 import pandas as pd
 df = pd.DataFrame(times)
-df.to_csv("performance/general_performance.csv")
+df.to_csv("general_performance.csv")
 print(df)
 

### 2023-07-24 LiveFromDAP/performance/load_code_performance.py

@@ -22,12 +22,7 @@ int useless_method(int x) {{
 """
 def compile_c():
     execute_command("gcc -g -fPIC -shared -o tmp/useless.so tmp/useless.c")
-def update_c_file(line):
-    with open(os.path.join("tmp", "useless.c"), "w") as f:
-        f.write(c_code_template.format("\nl++;"*line))
-    compile_c()
 
-update_c_file(0)
 c_source_path = os.path.abspath("tmp/useless.c")
 c_compiled_path = os.path.abspath("tmp/useless.so")
 c_method = "useless_method"
@@ -64,7 +59,6 @@ def update_java_file(line):
         f.write(java_code_template.format("\na++; "*line))
     compile_java()
 
-update_java_file(0)
 java_class_path = os.path.abspath("tmp")
 java_class_name = "UselessClass"
 java_method = "uselessMethod"
@@ -77,17 +71,35 @@ times = {
     "java": []
 }
 
+compile_times = {
+    "c": [],
+    "java": []
+}
+
 # C
 print(" == C == ")
 agent = CLiveAgent(debug=False)
 agent.start_server()
 agent.initialize()
+#warm up
+for _ in range(5):
+    with open(os.path.join("tmp", "useless.c"), "w") as f:
+        f.write(c_code_template.format(""))
+    compile_c()
+    agent.load_code(c_compiled_path)
+
 for i in range(100):
     print("Iteration: ", i, end="\r")
+    with open(os.path.join("tmp", "useless.c"), "w") as f:
+        f.write(c_code_template.format("\nl++;"*i))
+    t1 = time.time()
+    compile_c()
+    t2 = time.time()
+    compile_times["c"].append(t2 - t1)
+    
     t1 = time.time()
     agent.load_code(c_compiled_path)
     t2 = time.time()
-    update_c_file(i+1)
     times["c"].append(t2 - t1)
 agent.stop_server()
 print("")
@@ -97,12 +109,17 @@ print(" == Python == ")
 agent = PythonLiveAgent(debug=False)
 agent.start_server()
 agent.initialize()
+# Warm up
+for _ in range(5):
+    update_python_file(0)
+    agent.load_code(python_source_path)
+
 for i in range(100):
     print("Iteration: ", i, end="\r")
+    update_python_file(i)
     t1 = time.time()
     agent.load_code(python_source_path)
     t2 = time.time()
-    update_python_file(i+1)
     times["python"].append(t2 - t1)
     
[…]

### 2023-07-25 paper/paper.tex
Abstract, authors

@@
 
-\title{Live probes for free}
+\title{Language-Parametric Techniques for Live Probes}
 %\subtitle{Preparing Articles for Programming}% optional
@@
 \affiliation[a]{ENS Rennes, Bruz, France}
-\author[b]{Tijs van der Storm}
+\author[b]{Riemer van Rozen}
 \affiliation[b]{CWI, Amsterdam, The Netherlands}
-\author[b]{Riemer van Rozen}
+\author[b,c]{Tijs van der Storm}
+\affiliation[c]{University of Groningen, Groningen, The Netherlands}
 
@@
   %What is the broad context of the work? What is the importance of the general research area?
-  \emph{Context}
-  In his presentation "Inventing on Principles", Bret Victor demonstrates a live code editor: by specifying input values for a function, we can observe in real time the values taken by the variables during execution, as the code is written.
-  This information is often obtained using a language designed for live programming or by instrumentation of a specific runtime.
+  \paragraph{Context}
+  In the first part of his 2012 presentation ``Inventing on Principle''~\cite{InventingOnPrinciple}, Bret Victor gives a demo of live code editor for Javascript which shows the dynamic history of values of variables in real time.
+This form of live programming has become known as ``live probes''~\cite{GiladBlogMethodsLive,SeanMcDirmid,ExampleBasedProg}.
+Using live probes the programmer has a permanent and continuous insight into the dynamic evolution of function or method variables, thus improving feedback and developer experience.
   % What problem or question does the paper address? How has this problem or question been addressed by others (if at all)?
-  \emph{Inquiry}
-  \dots
+  \paragraph{Inquiry} Although Bret Victor shows a working prototype of live probes in the context of Javascript, he does not discuss strategies for implementing them.
+Later work~\cite{ExampleBasedProg} provides an implementation approach, but this requires a programming language to be implemented on top of the GraalVM runtime~\cite{GraalVM}.
+In this paper we investigate generic engineering strategies that can be applied in the context of many programming languages, without requiring dedicated compiler or runtime support.
+
   % What was done that unveiled new knowledge?
-  \emph{Approach}
-  In this paper we propose to exploit the capabilities of debuggers to obtain the data needed to design a live code editor.
+  \paragraph{Approach}
+  In this paper we explore how debugger protocols can be exploited to implement live probes.
+Methods or functions are compiled after every code change and executed inside the debugger.
+During execution the evolution of all local variables in the current stack frame are recorded and communicated back to the IDE for display to the user.
   % What new facts were uncovered? If the research was not results oriented, what new capabilities are enabled by the work?
-  \emph{Knowledge}
-  \dots
+  \paragraph{Knowledge}
+  It turns out that debugger protocols are rich enough for implementing live probes.
+Step-wise execution, hot code swapping and stack frame inspection provide the right granularity and sufficient information to realize live probes, without modifying compilers or language runtimes.
+Furthermore, it turns out that the recently proposed Debugger Adapter Protocol (DAP)~\cite{DAP} provides an even more generic approach of implementing live probes, but, in some cases, at the cost of a significant performance penalty.
+
   % What argument, feasibility proof, artifacts, or results and evaluation support this work?
-  \emph{Grounding}
-  \dots
+  \paragraph{Grounding}
+  We have implemented live probes using stack recording natively for Java through the Java Debugger Interface (JDI), and through the DAP for Java, Python, and C, all with minimum glue code.
+We evaluate the run-time performance of all four live probes prototypes, decomposed into: compile-after-change, hotswap, single step overhead, and stack recording overhead.
+Our results show that live probes on top of native debugger APIs are performant enough for interactive use.
+In the case of DAP, it depends on characteristics of the programming language.
+
   % Why does this work matter?
-  \emph{Importance}
-  \dots
+  \paragraph{Importance} Live programming improves the programmer experience by providing immediate feedback about a program's execution and eliminating disruptive edit-compile-restart sequences.
+Live probes are one way to shorten the programmer feedback loop at the level of functions and methods.
+Although live probes are not new, and have been implemented in (prototype) systems, our approach of building live probes on top of existing and generic debugger protocols opens up live probes for a host of mainstream programming language, with minimum effort.
 

### 2023-07-25 paper/paper.tex
Bret figure
bit of intro

@@
 \author[b]{Riemer van Rozen}
-\affiliation[b]{CWI, Amsterdam, The Netherlands}
+\affiliation[b]{Centrum Wiskunde \&\ Informatica (CWI), Amsterdam, The Netherlands}
 \author[b,c]{Tijs van der Storm}
@@
 Although live probes are not new, and have been implemented in (prototype) systems, our approach of building live probes on top of existing and generic debugger protocols opens up live probes for a host of mainstream programming language, with minimum effort.
+\end{abstract}
 
-\end{abstract}
 \section{Introduction}
@@
 
+Live programming~\cite{Tanimoto,UsableLive,JonathanEdwards,VanRozenEtal} promises a better programming experience by bringing the execution of a program closer to the abstractions of the source code.
+By erasing the distinction between a program on the one hand, and its execution on the other, programmers may enjoy a more fluid experience, with immediate dynamic feedback after every code change.
+
+Live programming has a rich history, both in academia and industrial research~\cite{???}, originating in the original Lisp systems and Smalltak which allowed dynamic replacement of code units (hot swapping) and inspection of program structures at run-time (sometimes referred to as ``programming in the debugger'').
+However, live programming became center stage after Bret Victor's influential talk ``Inventing on Principle'', which went viral in 2012.
+
+In that presentation Victor makes a compelling case that (paraphrasing) ``creators need to see what they are creating''.
+In the case of programming, this entails continuous inspection of the run-time behavior of a program, and immediate and seamless adaptation after a code change.
+While all the prototypes demonstrated by Victor are impressive, for the purpose of this paper we zoom in on one of them: a live editor for Javascript with a side pane showing the history of values of local function variables during an execution (see Figure~\ref{FIG:bret} below).
+Although not mentioned by Victor himself, this live programming features has become known as ``live probes''~\cite{SeanMcDirmid,ExampleBasedGraal}.
+
+The question left unanswered in Victor's presentation is: how to build such live probes?
+While there has been recent research showcasing an implementation approach~\cite{ExampleBasedGraal}, it depends on specific a runtime system (GraalVM), and cannot be easily transferred to other languages.
+So a more specific question is: how to  engineer live probes without dedicated compiler or runtime support?
+
+In this paper we present an approach to implementing live probes reusing standard debugger protocols and APIs.
+Most language support an API to facilitate the implementation of debuggers and other run-time tools (e.g., profilers, tracers, etc.).
+It turns out that such facilities are sufficiently powerful for implementing live probes.
+The key to the approach is to (re)execute a function or method inside the debugger, step by step, and at each step record a snapshot of the current stack-frame.
+After the method has finished executing, the resulting stack recording is communicated back to the IDE and visualized alongside the code, matching stack recording entries to the source code using source location information.
+
+We first detail how to implement live probes using a native debugger interface, the Java Debugger Interface (JDI).
+Then we generalize the approach to employ the recently proposed Debugger Adapter Protocol (DAP).
+This implementation is showcased with live probe implementations in Java (again), C, Python, and Javascript. 
+As a result, live probes can be supported in any programming language that provides an implementation of the DAP, with only a limited amount of configuration code.
+
+Since live programming (and hence live probes) is all about immediate feedback, it is important that the performance overhead of our implementation strategy does not impede practical use.
+We therefore provide a detail analysis and measurement of the performance overhead of (re)compilation, hot-swapping, step-wise execution, and stack recording.
+To assess the overall usability of the approach, we furthermore decompose a full editing scenario (derived from Victor's presentation) in $X$ logical steps, replay it on every live probe implementation, and measure the performance of each step.
+
+The contributions of this paper can be summarized as follows:
+\begin{itemize}
+  \item We analyze the requirements for implementing live probes (Section~\ref{sec:problem-overview})
+  \item We introduce \textit{stack recording} as a technique to obtain the required information through standard debugger protocols (Section~\ref{sec:stack-recording})
+  \item We implement live probes through stack recording on top of the Java Debugger Interface (JDI) (Section~\ref{sec:live-probes-java})
+  \item We then generalize the approach via the Debugger Adapter Protocol (DAP) (Section~\ref{sec:generalizing-live-probes})
+   \item We evaluate the feasibility of the approach by implementing live probes for Java, Python, C and Javascript on top the DAP, investigate the performance overhead of the individual steps, and reproduce an editing scenario to assess practicality (Section~\ref{sec:evaluation})
+\end{itemize}
+
 \section{Problem Overview}
 \label{sec:problem-overview}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{img/bret}
+  \caption{Screenshot of Bret Victor's 2012 talk ``Inventing on Principle'' (23rd minute)}
+  \label{FIG:bret}
+\end{figure}
 

### 2023-07-25 paper/paper.tex
requirements

--- 2023-07-24 paper/paper.tex
@@
 
+\begin{comment}
 \begin{table}[h]
@@
 
+To evaluate the performance of our approach in the general context of a live programming session, we measured the times taken by the different stages of the interaction loop:
+
+\begin{itemize}
+  \item Agent initialisation time, which occurs once at the start of the session.
+  \item The compilation time, if there is a compilation stage.
+  \item The time taken to load the code into the debugger.
+  \item Program execution time.
+\end{itemize}
+
+For the 3 languages we have implemented, we have performed the measurements for the execution of a binary search function \ref{fig:binary-search}.
+For each language, the agent was initialized, then the code was compiled, loaded into the debugger and executed (with the array [1,2,3,4,5,6] and target 9) 100 times.
+Each execution generate a stack recording of approximately 20 stack frames.
+The results are shown in table \ref{table:average-time}.
+\end{comment}
 \begin{figure}[htbp]
@@
 
-To evaluate the performance of our approach in the general context of a live programming session, we measured the times taken by the different stages of the interaction loop:
+To test the performance of our approach under real conditions, we recreated the live programming scenario presented in the video by bret victor. 
+The scenario describes the programming of a function performing a binary search on an array of characters.
+It consists of 18 steps, including 10 steps where the source code is modified and 8 steps where the parameters are changed.
+The scenario also includes the case where the tested method does not finish. As the code in Bret Victor's presentation is in Javascript, it has been adapted to the 3 languages studied in this paper.
 
-\begin{itemize}
-  \item Agent initialisation time, which occurs once at the start of the session.
-  \item The compilation time, if there is a compilation stage.
-  \item The time taken to load the code into the debugger.
-  \item Program execution time.
-\end{itemize}
+For this test we set the maximum size of the stack recording at 80 stackframes. When the recording exceeds this limit, the debugger is stopped and restarted.
 
-For the 3 languages we have implemented, we have performed the measurements for the execution of a binary search function \ref{fig:binary-search}.
-For each language, the agent was initialized, then the code was compiled, loaded into the debugger and executed (with the array [1,2,3,4,5,6] and target 9) 100 times.
-Each execution generate a stack recording of approximately 20 stack frames.
-The results are shown in table \ref{table:average-time}.
+The figure \ref{fig:scenario-bs} shows the time and number of stackframes recorded for each stage of the scenario in Python, C and Java. 
+When the initialisation is executed for Java and C, the time taken at each stage follows the same trend as the number of stack frames registered. 
+At step 16, we also observe the case of an infinite loop in the scenario.
 
@@
 
-\newpage
-\section*{Appendix}
-\label{sec:appendix}
-
-\begin{figure}[h]
-  \centering
-    \begin{lstlisting}[language=C]
-int binary_search(int arr[], int length, int target) {
-  int left = 0;
-  int right = length - 1;
-  while (left <= right) {
-
-      int mid = (left + right) / 2;
-      if (arr[mid] == target) {
-          return mid;
-      } else if (arr[mid] < target) {
-          left = mid + 1;
-      } else {
-          right = mid - 1;
-      }
-  }
-  return 0;
-}
-    \end{lstlisting}
-
-    \begin{lstlisting}[language=Python]
-def binary_search(arr, target):
-    left = 0
-    right = len(arr) - 1
-
-    while left <= right:
-        mid = (left + right) // 2
-
-        if arr[mid] == target:
-            return mid
-        elif arr[mid] < target:
-            left = mid + 1
-        else:
-            right = mid - 1
-
-    return 0 
-    \end{lstlisting}
-
-    \begin{lstlisting}[language=Java]
-public class BinarySearch {
-    public static int binarySearch(int[] array, int key){
-        int low = 0;
-        int high = array.length - 1;
-        while (low <= high) {
-            int mid = (low + high) / 2;
-            int value = array[mid];
-            if (value < key) {
-                low = mid + 1;
-            } else if (value > key) {
-                high = mid - 1;
-            } else {
-                return mid;
-            }
-        }
-        return 0;
-    }
-}
-      \end{lstlisting}
-  \caption{Binary Search in C, Python and Java}
-  \label{fig:binary-search}
-\end{figure}
-
 \end{document}

### 2023-07-25 paper/paper.tex
Overview

@@
   %% Choose exactly the one that best describes this work. (see 2.1)
-  perspective=art,
+  perspective=engineering,
   %% State one or more areas, separated by a comma. (see 2.2)
@@
   %% The list is open-ended, so use other areas if yours is/are not listed.
-  area={Social Coding, General-purpose programming},
+  area={Live programming, language engineering},
   %% You may choose the license for your paper (see 3.)
@@
   \paragraph{Grounding}
-  We have implemented live probes using stack recording natively for Java through the Java Debugger Interface (JDI), and through the DAP for Java, Python, and C, all with minimum glue code.
+  We have implemented live probes using stack recording natively for Java through the Java Debugger Interface (JDI), and through the DAP for Java, Python, C, and Javascript, all with minimum glue code.
 We evaluate the run-time performance of all four live probes prototypes, decomposed into: compile-after-change, hotswap, single step overhead, and stack recording overhead.
@@
 
-\section{Problem Overview}
+\section{Background and Overview}
 \label{sec:problem-overview}
 
-\begin{figure}
+The basic concept of live probes is illustrated in Figure~\ref{FIG:bret}, showing two panes. 
+The left pane shows an editor containing the source code of a possible implementation of binary search in Javascript. 
+The right pane shows the values of variables matched up to the source line they were last updated.
+The top two rows specify example input data that the function is executed on.
+The consecutive values of variables that are updated in a loop (e.g.
+\lstinline{low}, \lstinline{high}, \lstinline{mid} and \lstinline{value}) are displayed adjacently.
+So for \lstinline{low}, the history of values is 0, 3, and 5.
+Note that the update of \lstinline{high} in the else-if branch (apparently) never executed, so there is no (redundant) value display in the right pane.
+Finally, the last line shows the final return value of the function: apparently the \lstinline{key} was not found in the \lstinline[language=c]{array}.
+
+What makes this \textit{live} is that while the programmer is editing the program on the left (or the input arguments to the function), the value display on the right continuously updates so that it is apparent what is going on at run time.
+
+Dissecting this example leads to the following technical requirements for implementing live probes:
+\begin{itemize}
+  \item there should a way to specify the inputs to a method
+  \item after every change to the code, it needs to be recompiled and reexecuted
+  \item during execution the values of variables needs to be recorded and linked to source lines
+  \item the resulting history list of values needs to be communicated to the IDE
+\end{itemize}
+
+The basic intuition of our approach to meet these requirements is as follows: a separate process, called the \textit{Keep Alive Agent} (KAA) starts the debugger in the background and monitors the editor for changes to the source code.
+Whenever the programmer edits the code, the KAA attempts to recompile the code, and, if successful, hotswaps the current method into the debugger.
+It then queries the source code for annotations (either explicit annotations as in Java, or special comments) specifying the input arguments to the method.
+The KAA then sets a breakpoint on the first statement of the method, and invokes it.
+Since there is a breakpoint, the debugger halts.
+Using the common step-over facility of a typical debugger, the KAA then steps through the whole method until its execution completes.
+During this process however, the KAA takes snapshots of the current stack frame to record the values of local variables.
+These snapshots are stored in a data structure, called \textit{stack recording}, which supports displaying the values in the IDE, for instance, like in Figure~\ref{FIG:bret}, inline~\cite{LiveLit,ExampleBasedGraal}, or using some other UI affordance.
+
+\begin{figure}[t]
   \centering
@@
 
+LiveLiterals (Van der Storm, Live 2016)
+
 \section{Conclusion}

### 2023-07-25 paper/paper.tex
tweaks

@@
 The question left unanswered in Victor's presentation is: how to build such live probes?
-While there has been recent research showcasing an implementation approach~\cite{ExampleBasedGraal}, it depends on specific a runtime system (GraalVM), and cannot be easily transferred to other languages.
-So a more specific question is: how to  engineer live probes without dedicated compiler or runtime support?
+While there has been recent research showcasing implementation approaches~\cite{ExampleBasedGraal}, they depends on a specific runtime system (e.g., GraalVM), or language specific ``hacks''~\cite{LiveLit}, and hence cannot be easily transferred to other languages.
+So a more specific question is: how to  engineer live probes \textit{generically}, without dedicated compiler or runtime support?
 

### 2023-07-25 paper/paper.tex
Fixed some refs

@@
   In the first part of his 2012 presentation ``Inventing on Principle''~\cite{InventingOnPrinciple}, Bret Victor gives a demo of live code editor for Javascript which shows the dynamic history of values of variables in real time.
-This form of live programming has become known as ``live probes''~\cite{GiladBlogMethodsLive,SeanMcDirmid,ExampleBasedProg}.
+This form of live programming has become known as ``live probes''~\cite{MakingMethodsLive,UsableLiveProgramming,ExampleBasedGraalVM}.
 Using live probes the programmer has a permanent and continuous insight into the dynamic evolution of function or method variables, thus improving feedback and developer experience.
@@
   \paragraph{Inquiry} Although Bret Victor shows a working prototype of live probes in the context of Javascript, he does not discuss strategies for implementing them.
-Later work~\cite{ExampleBasedProg} provides an implementation approach, but this requires a programming language to be implemented on top of the GraalVM runtime~\cite{GraalVM}.
+Later work~\cite{ExampleBasedGraalVM} provides an implementation approach, but this requires a programming language to be implemented on top of the GraalVM runtime~\cite{GraalVM}.
 In this paper we investigate generic engineering strategies that can be applied in the context of many programming languages, without requiring dedicated compiler or runtime support.
@@
   \paragraph{Grounding}
-  We have implemented live probes using stack recording natively for Java through the Java Debugger Interface (JDI), and through the DAP for Java, Python, C, and Javascript, all with minimum glue code.
+  We have implemented live probes using stack recording natively for Java through the Java Debug Interface (JDI)~\cite{JDI}, and through the DAP for Java, Python, C, and Javascript, all with minimum glue code.
 We evaluate the run-time performance of all four live probes prototypes, decomposed into: compile-after-change, hotswap, single step overhead, and stack recording overhead.
@@
 
-Live programming~\cite{Tanimoto,UsableLive,JonathanEdwards,VanRozenEtal} promises a better programming experience by bringing the execution of a program closer to the abstractions of the source code.
+Live programming~\cite{TanimotoLevels,UsableLiveProgramming,ExampleCentric,Hancock03} promises a better programming experience by bringing the execution of a program closer to the abstractions of the source code.
 By erasing the distinction between a program on the one hand, and its execution on the other, programmers may enjoy a more fluid experience, with immediate dynamic feedback after every code change.
 
-Live programming has a rich history, both in academia and industrial research~\cite{???}, originating in the original Lisp systems and Smalltak which allowed dynamic replacement of code units (hot swapping) and inspection of program structures at run-time (sometimes referred to as ``programming in the debugger'').
+Live programming has a rich history, both in academia and industrial research\footnote{For brief and incomplete overview of some of the history of live programming, see: \url{https://liveprogramming.github.io/liveblog/2013/01/a-history-of-live-programming/}}, originating in the original Lisp systems (e.g.,~\cite{Sandewall78,Masinter81}) and Smalltalk~\cite{Goldberg80} which allowed dynamic replacement of code units (hot swapping) and inspection of program structures at run-time (sometimes referred to as ``programming in the debugger'').
 However, live programming became center stage after Bret Victor's influential talk ``Inventing on Principle'', which went viral in 2012.
@@
 While all the prototypes demonstrated by Victor are impressive, for the purpose of this paper we zoom in on one of them: a live editor for Javascript with a side pane showing the history of values of local function variables during an execution (see Figure~\ref{FIG:bret} below).
-Although not mentioned by Victor himself, this live programming features has become known as ``live probes''~\cite{SeanMcDirmid,ExampleBasedGraal}.
+Although not mentioned by Victor himself, this live programming features has become known as ``live probes''~\cite{UsableLiveProgramming,ExampleBasedGraalVM}.
 
 The question left unanswered in Victor's presentation is: how to build such live probes?
-While there has been recent research showcasing implementation approaches~\cite{ExampleBasedGraal}, they depends on a specific runtime system (e.g., GraalVM), or language specific ``hacks''~\cite{LiveLit}, and hence cannot be easily transferred to other languages.
+While there has been research showcasing implementation approaches~\cite{ExampleBasedGraalVM}, they depend on a specific runtime system (e.g., GraalVM), or employ language specific ``hacks''~\cite{LiveLiterals}, and hence cannot be easily transferred to other languages.
 So a more specific question is: how to  engineer live probes \textit{generically}, without dedicated compiler or runtime support?
@@
 
-Example-Based Live Programming for Everyone\cite{10.1145/3426428.3426919} : Use GraalVM/Truffle to get live information from the code of multiple languages.
+Example-Based Live Programming for Everyone\cite{ExampleBasedGraalVM} : Use GraalVM/Truffle to get live information from the code of multiple languages.
 
-Example Centric Programming\cite{10.1145/1052883.1052894} : Use BeanShell(custom JVM) to get live information. Prototype for Java in Eclipse.
+Example Centric Programming\cite{ExampleCentric} : Use BeanShell(custom JVM) to get live information. Prototype for Java in Eclipse.
 
-Usable Live Programming\cite{10.1145/2509578.2509585} : New language for live programming(Ying Yang) with incremental compilation. Live programming environment for this language.
+Usable Live Programming\cite{UsableLiveProgramming} : New language for live programming(Ying Yang) with incremental compilation. Live programming environment for this language.
 Use source location to relate execution and code(=> almost like stack recording that link stackframe and code location)
 
-Scalable Omniscient Debugging\cite{10.1145/1297105.1297067} : Omniscient debugging with a lot of data. Use a lot of memory to store all the data.
+Scalable Omniscient Debugging\cite{ScalableOmniscient} : Omniscient debugging with a lot of data. Use a lot of memory to store all the data.
 In this paper they record almost everything, we only record the stackframe.
 
-LiveLiterals (Van der Storm, Live 2016)
+LiveLiterals~\cite{LiveLiterals}
 

### 2023-07-25 paper/paper.tex
tweaks

@@
   %What is the broad context of the work? What is the importance of the general research area?
-  \paragraph{Context}
+  \textit{Context.}
   In the first part of his 2012 presentation ``Inventing on Principle''~\cite{InventingOnPrinciple}, Bret Victor gives a demo of live code editor for Javascript which shows the dynamic history of values of variables in real time.
@@
 Using live probes the programmer has a permanent and continuous insight into the dynamic evolution of function or method variables, thus improving feedback and developer experience.
+
   % What problem or question does the paper address? How has this problem or question been addressed by others (if at all)?
-  \paragraph{Inquiry} Although Bret Victor shows a working prototype of live probes in the context of Javascript, he does not discuss strategies for implementing them.
+  \textit{Inquiry.} Although Bret Victor shows a working prototype of live probes in the context of Javascript, he does not discuss strategies for implementing them.
 Later work~\cite{ExampleBasedGraalVM} provides an implementation approach, but this requires a programming language to be implemented on top of the GraalVM runtime~\cite{GraalVM}.
@@
   % What was done that unveiled new knowledge?
-  \paragraph{Approach}
+  \textit{Approach.}
   In this paper we explore how debugger protocols can be exploited to implement live probes.
@@
 During execution the evolution of all local variables in the current stack frame are recorded and communicated back to the IDE for display to the user.
+
   % What new facts were uncovered? If the research was not results oriented, what new capabilities are enabled by the work?
-  \paragraph{Knowledge}
+  \textit{Knowledge.}
   It turns out that debugger protocols are rich enough for implementing live probes.
@@
   % What argument, feasibility proof, artifacts, or results and evaluation support this work?
-  \paragraph{Grounding}
+  \textit{Grounding.}
   We have implemented live probes using stack recording natively for Java through the Java Debug Interface (JDI)~\cite{JDI}, and through the DAP for Java, Python, C, and Javascript, all with minimum glue code.
@@
   % Why does this work matter?
-  \paragraph{Importance} Live programming improves the programmer experience by providing immediate feedback about a program's execution and eliminating disruptive edit-compile-restart sequences.
+  \textit{Importance.} Live programming improves the programmer experience by providing immediate feedback about a program's execution and eliminating disruptive edit-compile-restart sequences.
 Live probes are one way to shorten the programmer feedback loop at the level of functions and methods.
@@
 
-Live programming has a rich history, both in academia and industrial research\footnote{For brief and incomplete overview of some of the history of live programming, see: \url{https://liveprogramming.github.io/liveblog/2013/01/a-history-of-live-programming/}}, originating in the original Lisp systems (e.g.,~\cite{Sandewall78,Masinter81}) and Smalltalk~\cite{Goldberg80} which allowed dynamic replacement of code units (hot swapping) and inspection of program structures at run-time (sometimes referred to as ``programming in the debugger'').
+Live programming has a rich history, both in academia and industrial research\footnote{For brief and incomplete overview of some of the history of live programming, see~\cite{LiveProgHist13}}, originating in the original Lisp systems (e.g.,~\cite{Sandewall78,Masinter81}) and Smalltalk~\cite{Goldberg80} which allowed dynamic replacement of code units (hot swapping) and inspection of program structures at run-time (sometimes referred to as ``programming in the debugger'').
 However, live programming became center stage after Bret Victor's influential talk ``Inventing on Principle'', which went viral in 2012.

### 2023-07-25 paper/paper.tex
add steps

--- 2023-07-25 paper/paper.tex
@@
   %% Choose exactly the one that best describes this work. (see 2.1)
-  perspective=engineering,
+  perspective=art,
   %% State one or more areas, separated by a comma. (see 2.2)
@@
   %% The list is open-ended, so use other areas if yours is/are not listed.
-  area={Live programming, language engineering},
+  area={Social Coding, General-purpose programming},
   %% You may choose the license for your paper (see 3.)
@@
   \paragraph{Grounding}
-  We have implemented live probes using stack recording natively for Java through the Java Debugger Interface (JDI), and through the DAP for Java, Python, C, and Javascript, all with minimum glue code.
+  We have implemented live probes using stack recording natively for Java through the Java Debugger Interface (JDI), and through the DAP for Java, Python, and C, all with minimum glue code.
 We evaluate the run-time performance of all four live probes prototypes, decomposed into: compile-after-change, hotswap, single step overhead, and stack recording overhead.
@@
 The question left unanswered in Victor's presentation is: how to build such live probes?
-While there has been recent research showcasing implementation approaches~\cite{ExampleBasedGraal}, they depends on a specific runtime system (e.g., GraalVM), or language specific ``hacks''~\cite{LiveLit}, and hence cannot be easily transferred to other languages.
-So a more specific question is: how to  engineer live probes \textit{generically}, without dedicated compiler or runtime support?
+While there has been recent research showcasing an implementation approach~\cite{ExampleBasedGraal}, it depends on specific a runtime system (GraalVM), and cannot be easily transferred to other languages.
+So a more specific question is: how to  engineer live probes without dedicated compiler or runtime support?
 
@@
 
-\section{Background and Overview}
+\section{Problem Overview}
 \label{sec:problem-overview}
 
-The basic concept of live probes is illustrated in Figure~\ref{FIG:bret}, showing two panes. 
-The left pane shows an editor containing the source code of a possible implementation of binary search in Javascript. 
-The right pane shows the values of variables matched up to the source line they were last updated.
-The top two rows specify example input data that the function is executed on.
-The consecutive values of variables that are updated in a loop (e.g.
-\lstinline{low}, \lstinline{high}, \lstinline{mid} and \lstinline{value}) are displayed adjacently.
-So for \lstinline{low}, the history of values is 0, 3, and 5.
-Note that the update of \lstinline{high} in the else-if branch (apparently) never executed, so there is no (redundant) value display in the right pane.
-Finally, the last line shows the final return value of the function: apparently the \lstinline{key} was not found in the \lstinline[language=c]{array}.
-
-What makes this \textit{live} is that while the programmer is editing the program on the left (or the input arguments to the function), the value display on the right continuously updates so that it is apparent what is going on at run time.
-
-Dissecting this example leads to the following technical requirements for implementing live probes:
-\begin{itemize}
-  \item there should a way to specify the inputs to a method
-  \item after every change to the code, it needs to be recompiled and reexecuted
-  \item during execution the values of variables needs to be recorded and linked to source lines
-  \item the resulting history list of values needs to be communicated to the IDE
-\end{itemize}
-
-The basic intuition of our approach to meet these requirements is as follows: a separate process, called the \textit{Keep Alive Agent} (KAA) starts the debugger in the background and monitors the editor for changes to the source code.
-Whenever the programmer edits the code, the KAA attempts to recompile the code, and, if successful, hotswaps the current method into the debugger.
-It then queries the source code for annotations (either explicit annotations as in Java, or special comments) specifying the input arguments to the method.
-The KAA then sets a breakpoint on the first statement of the method, and invokes it.
-Since there is a breakpoint, the debugger halts.
-Using the common step-over facility of a typical debugger, the KAA then steps through the whole method until its execution completes.
-During this process however, the KAA takes snapshots of the current stack frame to record the values of local variables.
-These snapshots are stored in a data structure, called \textit{stack recording}, which supports displaying the values in the IDE, for instance, like in Figure~\ref{FIG:bret}, inline~\cite{LiveLit,ExampleBasedGraal}, or using some other UI affordance.
-
-\begin{figure}[t]
+\begin{figure}
   \centering
@@
 The scenario describes the programming of a function performing a binary search on an array of characters.
-It consists of 18 steps, including 10 steps where the source code is modified and 8 steps where the parameters are changed.
-The scenario also includes the case where the tested method does not finish. As the code in Bret Victor's presentation is in Javascript, it has been adapted to the 3 languages studied in this paper.
+It consists of 19 steps, including 10 steps where the source code is modified and 9 steps where the parameters are changed.
+
+\begin{enumerate}
+  \item The function is defined and the parameters are defined to be an array of characters(\code{['a','b','c','d','e',f']}) and a target character(\code{'d'}).
+  \item A new variable \code{low} is defined.
+  \item A new variable \code{high} is defined.
+  \item A new variable \code{mid} is defined.
+  \item The variable \code{mid} is changed to be a integer.
+  \item A new variable \code{value} is defined.
+  \item A if and else snippets is added.
+  \item The target is changed to \code{'b'}.
+  \item The target is changed to \code{'c'}.
+  \item The target is changed to \code{'d'} and the code from the \code{mid} is refactored to be in a \code{while(true)} loop.
+  \item The target is changed to \code{'a'}.
+  \item The target is changed to \code{'b'}.
+  \item The target is changed to \code{'c'}.
+  \item The target is changed to \code{'d'}.
+  \item The target is changed to \code{'e'}.
+  \item The target is changed to \code{'f'}.
+  \item The target is changed to \code{'g'}.
+  \item The condition of the while loop is changed to \code{low <= high}.
+  \item A return \code{-1} is added at the end of the function.
+\end{enumerate}
+
+The scenario also includes the case where the tested method does not finish at step 17.
+As the code in Bret Victor's presentation is in Javascript, it has been adapted to the 3 languages studied in this paper.
 
@@
 
-LiveLiterals (Van der Storm, Live 2016)
-
 \section{Conclusion}

### 2023-07-25 paper/paper.tex
More ref fixes

@@
 During this process however, the KAA takes snapshots of the current stack frame to record the values of local variables.
-These snapshots are stored in a data structure, called \textit{stack recording}, which supports displaying the values in the IDE, for instance, like in Figure~\ref{FIG:bret}, inline~\cite{LiveLit,ExampleBasedGraal}, or using some other UI affordance.
+These snapshots are stored in a data structure, called \textit{stack recording}, which supports displaying the values in the IDE, for instance, like in Figure~\ref{FIG:bret}, inline~\cite{LiveLiterals,ExampleBasedGraalVM}, or using some other UI affordance.
 

### 2023-07-26 paper/paper.tex
another ref
temp notes

@@
 Live probes are one way to shorten the programmer feedback loop at the level of functions and methods.
-Although live probes are not new, and have been implemented in (prototype) systems, our approach of building live probes on top of existing and generic debugger protocols opens up live probes for a host of mainstream programming language, with minimum effort.
+Although live probes are not new, and have been implemented in (prototype) systems, our approach of building live probes on top of existing and generic debugger protocols opens up live probes for a host of mainstream programming languages, with minimum effort.
 \end{abstract}
@@
 
-The basic intuition of our approach to meet these requirements is as follows: a separate process, called the \textit{Keep Alive Agent} (KAA) starts the debugger in the background and monitors the editor for changes to the source code.
+\textbf{TODO: FIX THIS:}
+The basic intuition of our approach to meet these requirements is as follows: the editor starts a debugger, not on the user program, but on a driver program called the Keep Alive Agent (KAA).
+This program consists of a while-true-loop, halted on a breakpoint, and is designed to service live probe requests, without having to restart the debugger (that would be too slow).
+When the code changes
+
+a separate process, called the \textit{Keep Alive Agent} (KAA) starts the debugger in the background and monitors the editor for changes to the source code.
 Whenever the programmer edits the code, the KAA attempts to recompile the code, and, if successful, hotswaps the current method into the debugger.
@@
   \includegraphics[width=\textwidth]{img/bret}
-  \caption{Screenshot of Bret Victor's 2012 talk ``Inventing on Principle'' (23rd minute)}
+  \caption{Screenshot of Bret Victor's 2012 talk ``Inventing on Principle'' (23rd minute)~\cite{InventingOnPrinciple}}
   \label{FIG:bret}

### 2023-07-26 paper/paper.tex
fixed overview

@@
 \begin{itemize}
-  \item there should a way to specify the inputs to a method
-  \item after every change to the code, it needs to be recompiled and reexecuted
-  \item during execution the values of variables needs to be recorded and linked to source lines
-  \item the resulting history list of values needs to be communicated to the IDE
+  \item there should a way to specify the input arguments to a method (e.g., \lstinline{key} and \lstinline[language=c]{array});
+  \item after every change to the code, it needs to be recompiled and reexecuted;
+  \item during execution the values of variables needs to be recorded and linked to the source locations where they were updated.
+  \item the resulting history list of values needs to be communicated to the IDE for display to the user.
 \end{itemize}
 
-\textbf{TODO: FIX THIS:}
-The basic intuition of our approach to meet these requirements is as follows: the editor starts a debugger, not on the user program, but on a driver program called the Keep Alive Agent (KAA).
-This program consists of a while-true-loop, halted on a breakpoint, and is designed to service live probe requests, without having to restart the debugger (that would be too slow).
-When the code changes
+The basic intuition of our approach to meet these requirements is as follows: the editor starts a custom debugger (implemented against some debugger API), not on the user program, but on a driver program called the Keep Alive Agent (KAA).
+This program consists of a while-true-loop, halted on a breakpoint, and will be the verhicle for executing methods, without having to restart the debugger upon each request.
 
-a separate process, called the \textit{Keep Alive Agent} (KAA) starts the debugger in the background and monitors the editor for changes to the source code.
-Whenever the programmer edits the code, the KAA attempts to recompile the code, and, if successful, hotswaps the current method into the debugger.
-It then queries the source code for annotations (either explicit annotations as in Java, or special comments) specifying the input arguments to the method.
-The KAA then sets a breakpoint on the first statement of the method, and invokes it.
-Since there is a breakpoint, the debugger halts.
-Using the common step-over facility of a typical debugger, the KAA then steps through the whole method until its execution completes.
-During this process however, the KAA takes snapshots of the current stack frame to record the values of local variables.
-These snapshots are stored in a data structure, called \textit{stack recording}, which supports displaying the values in the IDE, for instance, like in Figure~\ref{FIG:bret}, inline~\cite{LiveLiterals,ExampleBasedGraalVM}, or using some other UI affordance.
+When the code changes, the editor communicates to the debugger to hotswap the code, and execute it with the input arguments extracted from the source code (based on some kind of annotation).
+The debugger, however, does not simply execute it, but simulates execution by issuing step-over commands until the method completes.
+To keep stack recording local to the current method, the debugger never steps \textit{into} a call to another method.
+
+In between steps, the debugger takes snapshots of the current stack frame to record the values of local variables.
+These snapshots are stored in a data structure, called \textit{stack recording}, which supports displaying the values in the IDE, for instance, like in Figure~\ref{FIG:bret}, inline~\cite{LiveLiterals,ExampleBasedGraalVM}, or using some dedicated UI affordance.
 

### 2023-07-26 paper/paper.tex
New figure and JDI paragraph

--- 2023-07-25 paper/paper.tex
@@
 \section{Using the Debugger ?}
+
+\subsection{Architecture}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.8\linewidth]{img/architecture.png}
+  \caption{General architecture (placeholder)}
+  \label{fig:architecture}
+\end{figure}
+
+A live environment must be able to react to two different events: a change in the code or a change in the test/input data. 
+If the code changes, we need to re-execute the code, and in the case of a compiled language, we need to compile the new code first. 
+If the data changes, we need to be able to re-execute the programme with the new data. 
+In addition, it is necessary to keep compilation and execution times low enough to maintain an interactive experience with the user.
+
+To meet these constraints, we propose the architecture shown in Figure \ref{fig:architecture} :
+\begin{itemize}
+  \item The IDE or editor communicates with a server. The IDE sends changes to the code and function parameters, and the server sends back the stack trace of the method.
+  \item The server initialises and communicates with the debugger to apply the changes received from the IDE, and generates a stack recording when the method is executed.
+  \item The debugger runs an intermediate program, a keep alive agent. This keeps the debugger alive to reduce initialisation times.
+\end{itemize}
+
 \label{sec:stack-recording}
@@
 Each rectangle represents the state of the stackframe during execution, with the source code location (here simplified to the line number).
-\subsection{Keep Alive Agent}
 
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{img/keepalive_agent.png}
-  \caption{Keep Alive Agent}
-  \label{fig:keepalive-agent}
+  \includegraphics[width=0.6\linewidth]{img/stackrecording_impl.png}
+  \caption{Record Stackrecording (placeholder)}
+  \label{fig:stack-recording-impl}
 \end{figure}
 
-A live environment must be able to react to two different events: a change in the code or a change in the test/input data. 
-If the code changes, we need to re-execute the code, and in the case of a compiled language, we need to compile the new code first. 
-If the data changes, we need to be able to re-execute the programme with the new data. 
-In addition, it is necessary to keep compilation and execution times low enough to maintain an interactive experience with the user.
-
-To meet these constraints, we propose to use the debugger on an intermediate program, a \textit{keep-alive agent}. 
-This program keeps the debugger alive between executions and code changes to reduce initialisation times. 
-The general operation of a keep-alive agent is shown in Figure \ref{fig:keepalive-agent} :
-% TODO : make figure clean and add better description here
-\begin{itemize}
-  \item At the start of the session, the debugger is started on the agent. Once initialised, the agent is paused.
-  \item The target code is loaded into the agent and a breakpoint is set at the input of the target method.
-  \item Execution of the target method triggers the breakpoint at the input of the method, and execution continues step by step to build the \textit{stack recording}.
-\end{itemize}
+Figure \ref{fig:stack-recording-impl} shows how a stack recording is recorded.
+To obtain the different states of the stackframe during the execution of a method, we place a breakpoint at the start of the function. 
+Once this has been reached, the method is executed step by step, using the debugger's step instruction. 
+Between each step, the state of the stackframe is recorded along with the source code's location information.
 
@@
 \label{sec:live-probes-java}
+% listing of a java class
+\begin{figure}[htbp]
+  \centering
+  \begin{lstlisting}[language=Java]
+public class Debugger{
+  Debugger(){
+    setBreakpointOnWhileTrue();
+    launchVM(KeepeAliveAgent);
+  }
+  public void loadClass(classPath, className){
+    addClassPathToDynamicClassLoader(classPath);
+    loadClass(className);
+  }
 
-We developed a Java backend using JDI, a debugging interface for Java, to implement the concepts discussed in the previous section. 
-The keep-alive agent includes a method for loading classes into the JVM.
+  public StackRecording execute(method, arguments){
+    createMirrorArguments(arguments);
+    setBreakpointAtMethod(method);
+    invokeMethod(method, arguments);
+    stackRecording = new StackRecording();
+    while(!isMethodFinished()){
+      stackRecording.add(getStackFrame());
+      stepDebuggee();
+    }
+    return stackRecording;
+  }
+}
+  \end{lstlisting}
+  \caption{JDI Debugger classes}
+  \label{fig:debugger-class}
+\end{figure}
 
-When executing the target method, the arguments are created in the client JVM and then passed to the debugger JVM using the \code{mirrorOf} and \code{newInstance} methods from the reflection API. 
+\begin{figure}[htbp]
+  \centering
+  \begin{lstlisting}[language=Java]
+public class KeepAliveAgent {
+  DynamicClassLoader dynamicClassLoader;
+
+  public void loadClass(String className){
+    dynamicClassLoader.loadClass(className);
+  }
+
+  public static void main(String[] args){
+    while (true) {}
+  }
+}
+  \end{lstlisting}
+  \caption{JDI Keep Alive Agent}
+  \label{fig:java-keep-alive-agent}
+\end{figure}
+
+We have developed a debugger capable of stack recording using JDI, a debugging interface for Java. 
+A simplified pseudo code of this class is presented in Figure \ref{fig:debugger-class}.
+The class includes a method to initialise the JVM (line 2), a method to search or reload a class in the VM (line 5) and a method to generate a stack recording of a method (line 10).
+
+For initialisation, a JVM containing the instrumentation provided by JDI is launched on the KeepAliveAgent class \ref{fig:java-keep-alive-agent} and a breakpoint is set on the infinite loop in the \code{main} function (line 9).
+This breakpoint is used to keep the debugger alive in a wait state, which is necessary to use the JDI primitives \code{mirrorOf}, \code{newInstance} and \code{invokeMethod}.
+
+To load a class, we use a dynamic ClassLoader which allows us to add class paths and load a class during execution. 
+When the code changes, the class is reloaded into the KeepAliveAgent. 
+This method has one limitation: it only works if the class signature has not changed, and only the contents of the methods have changed. 
+If the structure has changed, the KeepAliveAgent must be restarted. 
+This limitation can be overcome with certain JVM implementations\cite{KABANOV201151}.
+
+To execute the target method (line 10 in Debugger\ref{fig:debugger-class}), the arguments are created in the client JVM, then passed to the debugger JVM using the \code{mirrorOf} and \code{newInstance} methods of the reflection API (line 11).
+A breakpoint is then placed at the input of the target method (line 12).
 The target method is invoked using \code{invokeMethod}. 
-To handle events from the JDI when breakpoints are set, a dedicated thread is used to prevent deadlocks.
-
-If modifications are needed in the code of the target method, we employ the \code{redefineClasses} method. 
-This method allows changing the content of a class loaded in the JVM using an array of byte codes. 
-However, it has a limitation: it only works if the class signature remains unchanged. 
-If the class structure is modified, restarting the debugger is necessary.
+Stack recording is then performed using the algorithm described in the previous paragraph. 
+As long as the method is not finished, the debugger registers the stackframe and takes a step in the debuggee.
 

### 2023-07-26 paper/paper.tex

@@
 \begin{itemize}
-  \item there should a way to specify the inputs to a method
-  \item after every change to the code, it needs to be recompiled and reexecuted
-  \item during execution the values of variables needs to be recorded and linked to source lines
-  \item the resulting history list of values needs to be communicated to the IDE
+  \item there should a way to specify the input arguments to a method (e.g., \lstinline{key} and \lstinline[language=c]{array});
+  \item after every change to the code, it needs to be recompiled and reexecuted;
+  \item during execution the values of variables needs to be recorded and linked to the source locations where they were updated.
+  \item the resulting history list of values needs to be communicated to the IDE for display to the user.
 \end{itemize}
 
-The basic intuition of our approach to meet these requirements is as follows: a separate process, called the \textit{Keep Alive Agent} (KAA) starts the debugger in the background and monitors the editor for changes to the source code.
-Whenever the programmer edits the code, the KAA attempts to recompile the code, and, if successful, hotswaps the current method into the debugger.
-It then queries the source code for annotations (either explicit annotations as in Java, or special comments) specifying the input arguments to the method.
-The KAA then sets a breakpoint on the first statement of the method, and invokes it.
-Since there is a breakpoint, the debugger halts.
-Using the common step-over facility of a typical debugger, the KAA then steps through the whole method until its execution completes.
-During this process however, the KAA takes snapshots of the current stack frame to record the values of local variables.
-These snapshots are stored in a data structure, called \textit{stack recording}, which supports displaying the values in the IDE, for instance, like in Figure~\ref{FIG:bret}, inline~\cite{LiveLiterals,ExampleBasedGraalVM}, or using some other UI affordance.
+The basic intuition of our approach to meet these requirements is as follows: the editor starts a custom debugger (implemented against some debugger API), not on the user program, but on a driver program called the Keep Alive Agent (KAA).
+This program consists of a while-true-loop, halted on a breakpoint, and will be the verhicle for executing methods, without having to restart the debugger upon each request.
+
+When the code changes, the editor communicates to the debugger to hotswap the code, and execute it with the input arguments extracted from the source code (based on some kind of annotation).
+The debugger, however, does not simply execute it, but simulates execution by issuing step-over commands until the method completes.
+To keep stack recording local to the current method, the debugger never steps \textit{into} a call to another method.
+
+In between steps, the debugger takes snapshots of the current stack frame to record the values of local variables.
+These snapshots are stored in a data structure, called \textit{stack recording}, which supports displaying the values in the IDE, for instance, like in Figure~\ref{FIG:bret}, inline~\cite{LiveLiterals,ExampleBasedGraalVM}, or using some dedicated UI affordance.
 

### 2023-07-27 LiveFromDAP/src/livefromdap/__init__.py
Add Javascript Live Agent

@@ -1,3 +1,4 @@
 from .agent.CLiveAgent import CLiveAgent
 from .agent.PythonLiveAgent import PythonLiveAgent
 from .agent.JavaLiveAgent import JavaLiveAgent
+from .agent.JavascriptLiveAgent import JavascriptLiveAgent

### 2023-07-27 LiveFromDAP/src/livefromdap/agent/BaseLiveAgent.py

@@ -4,6 +4,11 @@ from abc import ABC, abstractmethod
 
 from livefromdap.utils.StackRecording import StackRecording
 
+class DebuggeeTerminatedError(Exception):
+    def __init__(self):
+        super().__init__("Debuggee terminated")
+    
+
 class BaseLiveAgentInterface(ABC):
     """Interface for the LiveAgent
     This class define all methods that a LiveAgent should implement"""
@@ -49,6 +54,12 @@ class BaseLiveAgent(BaseLiveAgentInterface):
     def __init__(self, *args, **kwargs):
         self.debug = kwargs.get("debug", False)
         self.seq = 0
+        
+    def __del__(self):
+        try:
+            self.stop_server()
+        except:
+            pass
 
     def new_seq(self):
         self.seq += 1
@@ -304,7 +315,6 @@ class BaseLiveAgent(BaseLiveAgentInterface):
                     if command is None or output["command"] == command:
                         return output
             if output["type"] == "event" and output["event"] == "terminated":
-                self.stop_server()
-                raise Exception("Debuggee terminated")
+                raise DebuggeeTerminatedError()
 
     

### 2023-07-27 LiveFromDAP/src/livefromdap/agent/JavascriptLiveAgent.py

@@ -1,0 +1,320 @@
+import os
+import subprocess
+from debugpy.common.messaging import JsonIOStream
+from debugpy.common import sockets
+
+from livefromdap.utils.StackRecording import Stackframe, StackRecording
+from .BaseLiveAgent import BaseLiveAgent, DebuggeeTerminatedError
+from tree_sitter import Language, Parser
+
+
+class JavaScriptPreprocessor:
+    """Take a javascript file and preprocess it to be used by the agent
+    This is done by adding module.exports = {function_name: function_name} to the end of the file
+    """
+    
+    def __init__(self, input_path, output_path):
+        self.input_path = input_path
+        self.output_path = output_path
+        self.lang = Language(os.path.abspath(os.path.join(os.path.dirname(__file__), "..", "bin","treesitter","javascript.so")), 'javascript')
+        self.parser = Parser()
+        self.parser.set_language(self.lang)
+        with open(self.input_path, "r") as f:
+            source = f.read()
+        self.tree = self.parser.parse(bytes(source, "utf8"))
+        self.extract_function()
+        self.add_module_exports()
+        
+    def extract_function(self):
+        self.functions = {}
+        query = self.lang.query(
+            """
+            (function_declaration
+                name: (identifier) @function.name
+                body: (statement_block) @function.body)
+            """
+        )
+        
+        captures = query.captures(self.tree.root_node)
+        # The capture is a list of tuples (node, capture)
+        # for each function we have to consecutive captures, the first one is the name, the second one is the body
+        for i in range(0, len(captures), 2):
+            fun_def = captures[i]
+            body = captures[i+1]
+            if fun_def[1] == "function.name" and body[1] == "function.body": 
+                self.functions[fun_def[0].text.decode("utf8")] = int(body[0].children[1].start_point[0])+1
+    
+    def add_module_exports(self):
+        with open(self.input_path, "r") as f:
+            source = f.read()
+            
+        query = self.lang.query(
+            """
+            (assignment_expression
+            left: (
+                member_expression
+                property: (property_identifier) @exports
+            ))
+            """
+        )
+        captures = query.captures(self.tree.root_node)
+        if any(c[0].text.decode("utf8") == "exports" for c in captures):
+            old = source.split("\n")
+            source = "\n".join(old[:captures[0][0].start_point[0]] + old[captures[0][0].end_point[0]+1:])
+            
+        source += "\nmodule.exports = {" + ",".join([name for name in self.functions.keys()]) + "}"
+        
[…]

### 2023-07-27 LiveFromDAP/src/livefromdap/runner/js_runner.js

@@ -1,0 +1,26 @@
+
+var target_function = null;
+var target_args = null;
+var last_path = null;
+
+var loadFile = function (filename) {
+    if (last_path != null) {
+        delete require.cache[require.resolve(last_path)]
+    }
+    last_path = filename;
+    let content = require(filename);
+    console.log(content);
+    // add content to global
+    for (let key in content) {
+        global[key] = content[key];
+    }
+}
+
+// test file /home/jbdod/CWI/LiveProbes/test.js
+while (true) {
+    if (target_function != null && target_args != null) {
+        let result = target_function.apply(null, target_args);
+        target_function = null;
+        target_args = null;
+    }  
+}

### 2023-07-27 LiveFromDAP/src/webdemo/AutoAgent.py
Adding v1 to web

@@ -3,8 +3,11 @@ import json
 import os
 from threading import Thread
 
+from tree_sitter import Language, Parser
+
 from livefromdap.agent.CLiveAgent import CLiveAgent
 from livefromdap.agent.JavaLiveAgent import JavaLiveAgent
+from livefromdap.agent.JavascriptLiveAgent import JavascriptLiveAgent
 from livefromdap.agent.PythonLiveAgent import PythonLiveAgent
 from prettyprinter.CPrettyPrinter import CPrettyPrinter
 from pycparser import c_parser, parse_file, c_generator
@@ -265,6 +268,83 @@ class AutoPythonLiveAgent(AutoLiveAgent):
 
     def execute(self, method, args):
         output = self.agent.execute(method, args)
+        if output[0] == "Interrupted":
+            self.agent.load_code(self.source_path)
+        # Get the output of the thread
+        # Save the json result in a file
+        return self.construct_result_json(method, output)
+    
+    
+class AutoJavascriptLiveAgent(AutoLiveAgent):
+    
+    def __init__(self, raw=False):
+        self.raw = raw
+        self.agent = JavascriptLiveAgent(debug=False)
+        self.agent.start_server()
+        self.agent.initialize()
+        self.source_path = os.path.abspath("src/webdemo/tmp/tmp.js")
+        with open(self.source_path, "w") as f:
+            f.write("")
+        self.lang = Language(os.path.abspath(os.path.join(os.path.dirname(__file__), "..", "livefromdap", "bin","treesitter","javascript.so")), 'javascript')
+        self.parser = Parser()
+        self.parser.set_language(self.lang)
+        self.previous_ast = None
+    
+    def restart(self):
+        self.agent.stop_server()
+        self.agent.start_server()
+        self.agent.initialize()
+
+
+    def check_if_parsable(self, code):
+        parsable = False
+        changed = False
+        try:
+            ast = self.parser.parse(bytes(code, "utf8"))
+            print("parsing ", ast.root_node.sexp())
+            # Query to find ERROR or MISSING nodes
+            query = self.lang.query("""
+            (ERROR) @error
+            """)
+            captures = query.captures(ast.root_node)
+            parsable = len(captures) == 0
+        except Exception as e:
+            return False, False
+        
+        if self.previous_ast is None:
+            self.previous_ast = ast
+            changed = True
+        elif self.previous_ast.get_changed_ranges(ast) != []:
+            self.previous_ast = ast
+            changed = True
+        return parsable, changed
+    
+    def update_code(self, code):
[…]

### 2023-07-27 LiveFromDAP/src/webdemo/main.py

@@ -5,7 +5,7 @@ import time
 import uuid
 from flask import Flask, render_template, request
 from flask_socketio import SocketIO, join_room, send
-from webdemo.AutoAgent import AutoCLiveAgent, AutoJavaLiveAgent, AutoPythonLiveAgent
+from webdemo.AutoAgent import AutoCLiveAgent, AutoJavaLiveAgent, AutoPythonLiveAgent, AutoJavascriptLiveAgent
 
 app = Flask(__name__)
 app.config['SECRET_KEY'] = 'secret!'
@@ -21,6 +21,8 @@ def create_agent(language, raw=False):
         return AutoJavaLiveAgent(raw=raw)
     elif language == "python":
         return AutoPythonLiveAgent(raw=raw)
+    elif language == "javascript":
+        return AutoJavascriptLiveAgent(raw=raw)
     else:
         raise NotImplementedError() # TODO implement other languages
 
@@ -30,6 +32,8 @@ def get_language_prefix(language):
     elif language == "java":
         return "//@"
     elif language == "c":
+        return "//@"
+    elif language == "javascript":
         return "//@"
     else:
         raise NotImplementedError()

### 2023-07-27 LiveFromDAP/performance/execute_performance.py
Add js to performance

@@ -2,6 +2,7 @@ import os
 import subprocess
 import time
 from livefromdap import CLiveAgent,PythonLiveAgent,JavaLiveAgent
+from livefromdap.agent.JavascriptLiveAgent import JavascriptLiveAgent
 
 def execute_command(command):
     process = subprocess.Popen(command, shell=True)
@@ -72,12 +73,29 @@ java_class_path = os.path.abspath("tmp")
 java_class_name = "ExecutePerformance"
 java_method = "executePerformanceMethod"
 
+# JS
 
+js_code_template = """
+function useless_method(x) {{
+    let d = 0;
+    for (let index = 0; index < x; index++) {{
+        d += index;
+    }}
+    return d;
+}}
+"""
+
+with open(os.path.join("tmp", "useless.js"), "w") as f:
+    f.write(js_code_template.format())
+
+js_source_path = os.path.abspath("tmp/useless.js")
+js_method = "useless_method"
 
 times = {
     "c": {},
     "python": {},
-    "java": {}
+    "java": {},
+    "js": {}
 }
 
 # C
@@ -139,6 +157,28 @@ for i in range(50):
     _, st = agent.execute(java_class_name, java_method, [str(i+1)], max_steps=1000000)
     t2 = time.time()
     times["java"][len(st.stackframes)] = t2 - t1
+
+agent.stop_server()
+print("")
+
+
+# JS
+print(" == Javascript == ")
+agent = JavascriptLiveAgent(debug=False)
+agent.start_server()
+agent.initialize()
+agent.load_code(js_source_path)
+
+# warm up
+for _ in range(3):
+    agent.execute(js_source_path, js_method, [1], max_steps=1000000)
+    
+for i in range(33):
+    print("Iteration: ", i, end="\r")
+    t1 = time.time()
+    _, st = agent.execute(js_source_path, js_method, [i+1], max_steps=1000000)
+    t2 = time.time()
+    times["js"][len(st.stackframes)] = t2 - t1
 
 agent.stop_server()
 print("")

### 2023-07-27 LiveFromDAP/performance/load_code_performance.py

@@ -2,6 +2,7 @@ import os
 import subprocess
 import time
 from livefromdap import CLiveAgent,PythonLiveAgent,JavaLiveAgent
+from livefromdap.agent.JavascriptLiveAgent import JavascriptLiveAgent
 
 def execute_command(command):
     process = subprocess.Popen(command, shell=True)
@@ -63,12 +64,30 @@ java_class_path = os.path.abspath("tmp")
 java_class_name = "UselessClass"
 java_method = "uselessMethod"
 
+# JS
+
+# Python
+js_code_template = """
+function useless_method(x){{
+    let res = x;{}
+    return res;
+}}
+"""
+def update_js_file(line):
+    with open(os.path.join("tmp", "useless.js"), "w") as f:
+        f.write(python_code_template.format("\n    res = res + 1;"*line))
+
+update_js_file(0)
+js_source_path = os.path.abspath("tmp/useless.js")
+js_method = "useless_method"
+
 
 
 times = {
     "c": [],
     "python": [],
-    "java": []
+    "java": [],
+    "js": []
 }
 
 compile_times = {
@@ -153,6 +172,28 @@ for i in range(100):
     agent.load_code(java_class_path, java_class_name)
     t2 = time.time()
     times["java"].append(t2 - t1)
+agent.stop_server()
+print("")
+
+
+# JS
+print(" == Javascript == ")
+agent = JavascriptLiveAgent(debug=False)
+agent.start_server()
+agent.initialize()
+# Warm up
+for _ in range(5):
+    update_js_file(0)
+    agent.load_code(js_source_path)
+
+for i in range(100):
+    print("Iteration: ", i, end="\r")
+    update_js_file(i)
+    t1 = time.time()
+    agent.load_code(js_source_path)
+    t2 = time.time()
+    times["js"].append(t2 - t1)
+    
 agent.stop_server()
 print("")
 

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/0.js

@@ -1,0 +1,4 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'd';undefined;true
+function binary_search(arr, target){
+    
+}

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/1.js

@@ -1,0 +1,4 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'd';undefined;true
+function binary_search(arr, target){
+    var low = 0;
+}

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/10.js

@@ -1,0 +1,19 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'a';0;false
+function binary_search(arr, target){
+    var low = 0;
+    var high = arr.length - 1;
+    while (true) {
+        var mid = Math.floor((low + high) / 2);
+        var value = arr[mid];
+
+        if (value < target) {
+            low = mid + 1;
+        }
+        else if (value > target) {
+            high = mid - 1;
+        }
+        else {
+            return mid;
+        }
+    }
+}

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/11.js

@@ -1,0 +1,19 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'b';1;false
+function binary_search(arr, target){
+    var low = 0;
+    var high = arr.length - 1;
+    while (true) {
+        var mid = Math.floor((low + high) / 2);
+        var value = arr[mid];
+
+        if (value < target) {
+            low = mid + 1;
+        }
+        else if (value > target) {
+            high = mid - 1;
+        }
+        else {
+            return mid;
+        }
+    }
+}

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/12.js

@@ -1,0 +1,19 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'c';2;false
+function binary_search(arr, target){
+    var low = 0;
+    var high = arr.length - 1;
+    while (true) {
+        var mid = Math.floor((low + high) / 2);
+        var value = arr[mid];
+
+        if (value < target) {
+            low = mid + 1;
+        }
+        else if (value > target) {
+            high = mid - 1;
+        }
+        else {
+            return mid;
+        }
+    }
+}

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/13.js

@@ -1,0 +1,19 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'd';3;false
+function binary_search(arr, target){
+    var low = 0;
+    var high = arr.length - 1;
+    while (true) {
+        var mid = Math.floor((low + high) / 2);
+        var value = arr[mid];
+
+        if (value < target) {
+            low = mid + 1;
+        }
+        else if (value > target) {
+            high = mid - 1;
+        }
+        else {
+            return mid;
+        }
+    }
+}

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/14.js

@@ -1,0 +1,19 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'e';4;false
+function binary_search(arr, target){
+    var low = 0;
+    var high = arr.length - 1;
+    while (true) {
+        var mid = Math.floor((low + high) / 2);
+        var value = arr[mid];
+
+        if (value < target) {
+            low = mid + 1;
+        }
+        else if (value > target) {
+            high = mid - 1;
+        }
+        else {
+            return mid;
+        }
+    }
+}

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/15.js

@@ -1,0 +1,19 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'f';5;false
+function binary_search(arr, target){
+    var low = 0;
+    var high = arr.length - 1;
+    while (true) {
+        var mid = Math.floor((low + high) / 2);
+        var value = arr[mid];
+
+        if (value < target) {
+            low = mid + 1;
+        }
+        else if (value > target) {
+            high = mid - 1;
+        }
+        else {
+            return mid;
+        }
+    }
+}

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/16.js

@@ -1,0 +1,19 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'g';Interrupted;false
+function binary_search(arr, target){
+    var low = 0;
+    var high = arr.length - 1;
+    while (true) {
+        var mid = Math.floor((low + high) / 2);
+        var value = arr[mid];
+
+        if (value < target) {
+            low = mid + 1;
+        }
+        else if (value > target) {
+            high = mid - 1;
+        }
+        else {
+            return mid;
+        }
+    }
+}

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/17.js

@@ -1,0 +1,19 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'g';undefined;true
+function binary_search(arr, target){
+    var low = 0;
+    var high = arr.length - 1;
+    while (low <= high) {
+        var mid = Math.floor((low + high) / 2);
+        var value = arr[mid];
+
+        if (value < target) {
+            low = mid + 1;
+        }
+        else if (value > target) {
+            high = mid - 1;
+        }
+        else {
+            return mid;
+        }
+    }
+}

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/18.js

@@ -1,0 +1,20 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'g';-1;true
+function binary_search(arr, target){
+    var low = 0;
+    var high = arr.length - 1;
+    while (low <= high) {
+        var mid = Math.floor((low + high) / 2);
+        var value = arr[mid];
+
+        if (value < target) {
+            low = mid + 1;
+        }
+        else if (value > target) {
+            high = mid - 1;
+        }
+        else {
+            return mid;
+        }
+    }
+    return -1;
+}

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/2.js

@@ -1,0 +1,5 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'd';undefined;true
+function binary_search(arr, target){
+    var low = 0;
+    var high = arr.length - 1;
+}

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/3.js

@@ -1,0 +1,7 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'd';undefined;true
+function binary_search(arr, target){
+    var low = 0;
+    var high = arr.length - 1;
+
+    var mid = (low + high) / 2;
+}

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/4.js

@@ -1,0 +1,7 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'd';undefined;true
+function binary_search(arr, target){
+    var low = 0;
+    var high = arr.length - 1;
+
+    var mid = Math.floor((low + high) / 2);
+}

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/5.js

@@ -1,0 +1,8 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'd';undefined;true
+function binary_search(arr, target){
+    var low = 0;
+    var high = arr.length - 1;
+
+    var mid = Math.floor((low + high) / 2);
+    var value = arr[mid];
+}

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/6.js

@@ -1,0 +1,18 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'd';undefined;true
+function binary_search(arr, target){
+    var low = 0;
+    var high = arr.length - 1;
+
+    var mid = Math.floor((low + high) / 2);
+    var value = arr[mid];
+
+    if (value < target) {
+        low = mid + 1;
+    }
+    else if (value > target) {
+        high = mid - 1;
+    }
+    else {
+        return mid;
+    }
+}

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/7.js

@@ -1,0 +1,18 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'b';undefined;false
+function binary_search(arr, target){
+    var low = 0;
+    var high = arr.length - 1;
+
+    var mid = Math.floor((low + high) / 2);
+    var value = arr[mid];
+
+    if (value < target) {
+        low = mid + 1;
+    }
+    else if (value > target) {
+        high = mid - 1;
+    }
+    else {
+        return mid;
+    }
+}

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/8.js

@@ -1,0 +1,18 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'c';2;false
+function binary_search(arr, target){
+    var low = 0;
+    var high = arr.length - 1;
+
+    var mid = Math.floor((low + high) / 2);
+    var value = arr[mid];
+
+    if (value < target) {
+        low = mid + 1;
+    }
+    else if (value > target) {
+        high = mid - 1;
+    }
+    else {
+        return mid;
+    }
+}

### 2023-07-27 LiveFromDAP/performance/scenario/binary_search/javascript/9.js

@@ -1,0 +1,19 @@
+['a', 'b', 'c', 'd', 'e', 'f'], 'd';3;true
+function binary_search(arr, target){
+    var low = 0;
+    var high = arr.length - 1;
+    while (true) {
+        var mid = Math.floor((low + high) / 2);
+        var value = arr[mid];
+
+        if (value < target) {
+            low = mid + 1;
+        }
+        else if (value > target) {
+            high = mid - 1;
+        }
+        else {
+            return mid;
+        }
+    }
+}

### 2023-07-27 LiveFromDAP/performance/scenario/evaluate_javascript.py

@@ -1,0 +1,49 @@
+import os
+import json
+import re
+import time
+from Scenario import Scenario
+from livefromdap import JavascriptLiveAgent
+
+
+if not os.path.exists("tmp"):
+    os.makedirs("tmp")
+
+method = "binary_search"
+source_path = os.path.abspath("tmp/binary_search.js")
+scenario = Scenario(os.path.abspath("binary_search"), "javascript", ".js", os.path.abspath("tmp/binary_search.js"))
+times = {
+    "init": 0,
+    "scenario":{}
+}
+
+# Python
+print(" == Starting == ")
+t1 = time.time()
+agent = JavascriptLiveAgent(debug=False)
+agent.start_server()
+agent.initialize()
+t2 = time.time()
+times["init"] = t2-t1
+
+for i, (args_str, ret, _) in enumerate(scenario):
+    print("Iteration: ", i, end="\r")
+    times['scenario'][i] = {}
+    args = re.split(r',(?![^\[\]\(\)\{\}]*[\]\)\}])', args_str)
+    args = list(map(lambda x: x.strip(), args))
+    t1 = time.time()
+    agent.load_code(source_path)
+    t2 = time.time()
+    times['scenario'][i]['load'] = t2-t1
+    
+    t1 = time.time()
+    res, st = agent.execute(source_path, method, args,max_steps=80)
+    t2 = time.time()
+    times['scenario'][i]['exec'] = t2-t1
+    times['scenario'][i]['step'] = len(st)
+    assert res == ret
+    
+agent.stop_server()
+print("")
+with open("binary_search_javascript_times.json", 'w') as file:
+    json.dump(times, file, indent=4)

### 2023-07-27 paper/paper.tex
Add JS to paper
Rewrite DAP part + add JS

@@
 \usepackage{subcaption} % For subfigures
+\usepackage{array, caption, tabularx,  ragged2e,  booktabs}
 
@@
 \newcommand{\code}[1]{\colorbox{codegray}{\texttt{#1}}}
+
+\definecolor{javared}{rgb}{0.6,0,0} % for strings
+\definecolor{javagreen}{rgb}{0.25,0.5,0.35} % comments
+\definecolor{javapurple}{rgb}{0.5,0,0.35} % keywords
+\definecolor{javadocblue}{rgb}{0.25,0.35,0.75} % javadoc
+
+\lstset{language=Java,
+keywordstyle=\color{javapurple},
+stringstyle=\color{javared},
+commentstyle=\color{javagreen}}
 
@@
 
-By exploiting this protocol, we have created a new language parametric backend, which we have implemented for the C, Python and Java programming languages.
+By exploiting this protocol, we have created a new language parametric backend, which we have implemented for the C, Python, Java and Javascript programming languages.
 These languages are chosen to cover both compiled and interpreted languages.
@@
 These methods depend on both the implementation of the debugging server and the keep-alive agent.
-The debug server initialisation parameters are specific to each language.
+The differences between the implementations are shown in the Table \ref{tab:implementation-differences}.
 
-The way in which the code is loaded also depends on the language. 
-For interpreted languages such as Python, the code can be interpreted at runtime and then loaded into the debugger's memory. 
-For Java, we have extended the ClassLoader to add and modify classpaths and classes at runtime in the keep-alive agent. 
-For C, the code is loaded into shared libraries that can be added and reloaded during execution.
+\begin{table}[h]
+    \centering
+    \noindent\setlength\tabcolsep{4pt}%
+    \begin{tabularx}{\linewidth}{lc*{4}{>{\RaggedRight\arraybackslash}X}}
+      \toprule
+      Implementation & Compile & Load Code & Execution Caller \\ [0.5ex]
+      \midrule
+      JDI Java & Java Compiler API & Dynamic ClassLoader & Debugger \\ 
+      \addlinespace
+      DAP Java & \code{javac} from command line & Dynamic ClassLoader & Debuggee \\ 
+      \addlinespace
+      DAP Python & ~ & Interpreted in debug console & Debuggee \\ 
+      \addlinespace
+      DAP C & \code{gcc} from command line & Loaded as shared library & Debugger \\
+      \addlinespace
+      DAP Javascript & ~ & Imported as module & Debuggee \\ 
+      \bottomrule
+    \end{tabularx}
+    \caption{Implementation differences between languages}
+    \label{tab:implementation-differences}
+\end{table}
 
-Method execution for Java and Python is different to that used in C. 
-For Java and Python, calling methods directly from the debugger does not trigger breakpoints for these languages. 
-To remedy this, execution must be initiated from the agent and not by a debugger command. 
-To do this, the Python and Java agents have fields for referencing a method and its arguments; when this information is entered, the agent starts execution.
-For C, as is the case in the backend using JDI, the method call is made from the debugger console.
+With DAP, the compilation step is performed by calling a command line compilation tool (javac and gcc).
+For the JDI version, developed in Java, the Java compilation API has been used for better performance.
+
+The way the code is loaded also depends on the programming language. For Python, the code is loaded by being interpreted by the debug console during execution.
+For Javascript, a line is added at the end of the imported file to export all the functions in the file as a module, which is then loaded into the KeepAliveAgent with \code{require}.
+For Java, as with the JDI version, a DynamicClassLoader is used. For C, the code is loaded into shared libraries that can be added and reloaded at runtime.
+
+Method execution can be divided into two categories.
+For Java with DAP, Python and Javascript, calling methods directly from the debugger does not trigger breakpoints.
+To remedy this, execution must be initiated by the KeepAliveAgent rather than by a debugger command.
+To do this, the KeepAliveAgent code in these languages has fields for referencing a method and its arguments; when this information is entered, the agent starts execution.
+In C, as in Java with JDI, the method is called from the debugger console.
 
@@
 
-We have developed a dynamic programming environment that incorporates live probes through Java, C, and Python debug servers. 
+We have developed a dynamic programming environment that incorporates live probes through Java, C, Python, Javascript debug servers. 
 The user interface consists of a web page that interacts with a local server. 
@@
 
-In our evaluation, we assessed the time required to compile and load code into the debugger memory for Python, C, and Java. 
+In our evaluation, we assessed the time required to compile and load code into the debugger memory for Python, C, Java and Javascript. 
 To accomplish this, we compiled and loaded programs ranging from 5 to 100 lines of code. 
@@
 
-Moving to Sub-figure \ref{subfig:load}, we depict the loading time in the debugger for C, Java, and Python as a function of the number of lines of code. 
+Moving to Sub-figure \ref{subfig:load}, we depict the loading time in the debugger for C, Java, Python and Javascript as a function of the number of lines of code. 
 The data shows that the loading time remains constant concerning the number of lines of code. 
-On average, Python takes 3 ms, Java takes 2.8 ms, and C takes 1 ms for loading.
+%On average, Python takes 3 ms, Java takes 2.8 ms, and C takes 1 ms for loading.
 
@@
   \centering
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.45\textwidth}
       \centering
@@
   \hfill
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.45\textwidth}
       \centering
@@
   \end{subfigure}
-  \hfill
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.45\textwidth}
     \centering
@@
     \label{subfig:scenario-bs-py}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.45\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{img/scenario_bs_javascript.png}
+    \caption{\centering Javascript}
+    \label{subfig:scenario-bs-js}
   \end{subfigure}
@@
 
-The figure \ref{fig:scenario-bs} shows the time and number of stackframes recorded for each stage of the scenario in Python, C and Java. 
+The figure \ref{fig:scenario-bs} shows the time and number of stackframes recorded for each stage of the scenario in Python, C, Java and Javascript. 
 When the initialisation is executed for Java and C, the time taken at each stage follows the same trend as the number of stack frames registered. 

### 2023-07-27 LiveFromDAP/install.sh
add js to install

@@ -77,4 +77,61 @@ else
     exit 1
 fi
 
+# Node Debug
+echo "[Node Debug] Checking for Node"
+# check if node is installed
+# if node -v exit with 0 then it is installed
+if ! node -v &> /dev/null; then
+    echo "You need to install Node.js"
+    exit 1
+fi
+
+# Js DAP
+echo "[Js DAP] Checking for Js DAP..."
+# check if src/livefromdap/bin/js-dap directory exists
+if [ ! -d "src/livefromdap/bin/js-dap" ]; then
+    echo "Downloading Js DAP..."
+    mkdir -p src/livefromdap/bin/js-dap
+    cd src/livefromdap/bin/js-dap
+    git clone https://github.com/microsoft/vscode-js-debug
+    cd vscode-js-debug
+    npm install
+    # check if gulp is installed
+    # if gulp -v exit with 0 then it is installed
+    if ! gulp -v &> /dev/null; then
+        echo "Installing gulp..."
+        npm install --global gulp-cli
+    fi
+    gulp dapDebugServer
+    cd ../..
+    cp -r js-dap/vscode-js-debug/dist/src/* js-dap/
+    rm -rf js-dap/vscode-js-debug
+    cd ../../..
+else
+    echo "Js DAP already installed."
+fi
+
+
+# Js Tree sitter
+echo "[Js Tree sitter] Checking for Js tree sitter..."
+# check if src/livefromdap/bin/treesitter/javascript.so exists
+if [ ! -f "src/livefromdap/bin/treesitter/javascript.so" ]; then
+    echo "Downloading Js tree sitter..."
+    mkdir -p src/livefromdap/bin/treesitter
+    cd src/livefromdap/bin/treesitter
+    git clone https://github.com/tree-sitter/tree-sitter-javascript
+    python -c "from tree_sitter import Language;Language.build_library('javascript.so', ['tree-sitter-javascript'])"
+    rm -rf tree-sitter-javascript
+    cd ../../..
+else
+    echo "Js tree sitter already installed."
+fi
+
+# Make runner files executable
+echo "[Runner] Making runner files executable..."
+cd src/livefromdap
+make runner
+chmod +x runner/*
+cd ../..
+
 

### 2023-07-28 paper/paper.tex
tweak
Discussion

@@
 
-\section{Related Work}
+\section{Discussion \&\ Related Work}
 \label{sec:related-work}
+
+\subsection{Discussion}
+
+While our approach to implement live probes on top of off-the-shelf debuggers provides a viable strategy to add live probes to many mainstream languages, some limitations and open questions remain.
+Below discuss these in more detail.
+
+``All problems in computer science can be solved by another level of indirection,''\footnote{Attributed to David Wheeler (\url{https://en.wikipedia.org/wiki/David_Wheeler_(computer_scientist)})} yet that often creates other problems.
+This state of affairs can be recognized in some of the performance results of the DAP-based live probes discussed above.
+Layering another level of abstraction (the DAP) between the native debuggers and the live probe server, causes performance to degrade, in this case mostly because of the asynchronous nature of the DAP, and serializing data shipped back and forth over sockets (TO CHECK).
+
+In our current approach, the programmer specifies input arguments to functions or methods using special comments in the code.
+This choice is motivated by the fact that it works in every programming language, does not require special UI affordances, and allows example data to be committed to version control systems.
+Nevertheless, while most primitive data types have a convenient, human readable literal notation, this can become cumbersome with complex, hetergeneous, structured data, such as arrays, objects, records, etc.
+Moreover, these literal notations are language specific: how to specify a constant array in C, Java or Javascript is very different, and needs to be parsed into actual data values from their textual representation in the special comments.
+
+Another challenge is most visible in the case of object-oriented languages, such as Java, where methods typically live in classes, and can only be invoked when an instance of a class is available.
+This means the programmer has to specify not just the input parameters, but also values for fields, or arguments to constructors, so that the live probe server can construct the object.
+This situation becomes worse if, e.g., the constructor depends on other objects as argument, which, in turn, require initialization, and so on.
+
+The above problem of objects and initialization hints at a more basic and general question: how to obtain useful example data in the first place.
+In this paper we have assumed the programmer provides the data, but there are other strategies worth considering.
+For instance:
+\begin{itemize}
+  \item As a base case, one could use default values for all arguments (e.g., \lstinline{0}, \lstinline{""}, \lstinline{null}, etc.).
+This has the drawback that such values might not lead to interesting probe displays.
+  \item Random value generation, as is employed in randomized testing, could be used to assign values to parameters.
+However, in this case it is unclear if it leads to the insight the programmer wants, and would be very confusing if the values would change in every iteration.
+  \item If a test suite is available, an off-line tool could observe frequently occurring values and objects passed into methods, through dynamic analysis of the test suite execution.
+Just like with random generation, however, the problem is to make the assignment stable somehow.
+\end{itemize}
+The above strategies absolve the programmer from having to specify the example values, but potentially at the cost of reduced quality of feedback.
+Further research is needed to find an appropriate middle ground between full manual specification, and fully automatic techniques.
+
+Live probes require continual execution of methods and functions.
+This is fine if the code does not perform any harmful side-effects, such as sending out email, writing files, or launching missiles.
+While it is probably undecidable to statically check whether a piece of code might eventually perform such a side effect, we leave it to the programmer to indicate whether the live probing feature is switched on, e.g., by providing the annotation of the input arguments.
+
+Possible future directions to solve this problem would include sandboxed execution~\cite{???}, or intercepting all calls to IO libraries or the operating system.
+The latter technique would also allow the live probes to not only display variable values, but also, e.g., console output.
+A more disciplined solution would be offered by languages with built-in notions of capability-based security (such as, e.g., Newspeak~\cite{Newspeak}), so that the live probes can be configured to execute in an environment where dangerous features (such as IO) are not available.
+
+While the live probes that we have demonstrated support variables updated in loops by displaying them in tabular form, we have not discussed what happens with recursion.
+The live probe execution repeatedly instructs the debugger to step-over function calls, intermediate stack frames are automatically excluded from the stack-recording, \textit{except} for execution arrives back at the function that is currently probed, since it has a breakpoint on entry.
+Our current implementation simply continues, and adds stack frames to the recording, even though we are on a different level of the stack.
+Further research is needed to explicitly deal with this case, and find an appropriate tree-based or nested visualization.
+
+\subsection{Related Work}
 

### 2023-07-28 paper/paper.tex
Discussion

@@
 This situation becomes worse if, e.g., the constructor depends on other objects as argument, which, in turn, require initialization, and so on.
+At minimum, the live probe server could construct basic objects with default values for all constructor arguments, and let the programmer gradually modify improve the assignment using annotations.
 
@@
   \item If a test suite is available, an off-line tool could observe frequently occurring values and objects passed into methods, through dynamic analysis of the test suite execution.
-Just like with random generation, however, the problem is to make the assignment stable somehow.
+Such common values could also be harvested from production runs.
+Just like with random generation, however, the problem is to make the assignments stable somehow.
 \end{itemize}

### 2023-07-31 paper/paper.tex
Paper correction 1

@@
   \centering
-  \begin{minipage}{0.2\textwidth}
+  \begin{minipage}{0.25\textwidth}
+    \centering
     \begin{lstlisting}[language=C]
@@
   \begin{minipage}{0.7\textwidth}
-    \begin{tikzpicture}[x=0.75pt,y=0.75pt,yscale=-1.1,xscale=0.95]
-
-%Shape: Rectangle [id:dp48530454252997635] 
-\draw   (11,13.3) -- (68.6,13.3) -- (68.6,34.5) -- (11,34.5) -- cycle ;
-%Shape: Rectangle [id:dp4741172814816692] 
-\draw   (11,34.5) -- (68.6,34.5) -- (68.6,89.7) -- (11,89.7) -- cycle ;
-%Shape: Rectangle [id:dp39333640528563685] 
-\draw   (83.8,13.3) -- (141.4,13.3) -- (141.4,34.5) -- (83.8,34.5) -- cycle ;
-%Shape: Rectangle [id:dp852338699133017] 
-\draw   (83.8,34.5) -- (141.4,34.5) -- (141.4,89.7) -- (83.8,89.7) -- cycle ;
-%Shape: Rectangle [id:dp7978026394194542] 
-\draw   (155.6,13.3) -- (213.2,13.3) -- (213.2,34.5) -- (155.6,34.5) -- cycle ;
-%Shape: Rectangle [id:dp7423976709217543] 
-\draw   (155.6,34.5) -- (213.2,34.5) -- (213.2,89.7) -- (155.6,89.7) -- cycle ;
-%Shape: Rectangle [id:dp7923849318907303] 
-\draw   (225.8,13.7) -- (283.4,13.7) -- (283.4,34.9) -- (225.8,34.9) -- cycle ;
-%Shape: Rectangle [id:dp8325747263531937] 
-\draw   (225.8,34.9) -- (283.4,34.9) -- (283.4,90.1) -- (225.8,90.1) -- cycle ;
-%Shape: Rectangle [id:dp3066063334990279] 
-\draw   (343,14.1) -- (400.6,14.1) -- (400.6,35.3) -- (343,35.3) -- cycle ;
-%Shape: Rectangle [id:dp9425547748588214] 
-\draw   (343,35.3) -- (400.6,35.3) -- (400.6,90.5) -- (343,90.5) -- cycle ;
-%Straight Lines [id:da5537198175803346] 
-\draw    (69.4,56.1) -- (80,56.1) ;
-\draw [shift={(83,56.1)}, rotate = 180] [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.08]  [draw opacity=0] (7.14,-3.43) -- (0,0) -- (7.14,3.43) -- cycle    ;
-%Straight Lines [id:da9992180079914922] 
-\draw    (141.4,56.1) -- (152,56.1) ;
-\draw [shift={(155,56.1)}, rotate = 180] [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.08]  [draw opacity=0] (7.14,-3.43) -- (0,0) -- (7.14,3.43) -- cycle    ;
-%Straight Lines [id:da24526356738515775] 
-\draw    (212.4,56.1) -- (223,56.1) ;
-\draw [shift={(226,56.1)}, rotate = 180] [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.08]  [draw opacity=0] (7.14,-3.43) -- (0,0) -- (7.14,3.43) -- cycle    ;
-%Straight Lines [id:da7346663349503276] 
-\draw [line width=1.25, dashed]    (284.4,56.1) -- (337.6,56.47) ;
-\draw [shift={(341.6,56.5)}, rotate = 180.4] [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.08]  [draw opacity=0] (9.29,-4.46) -- (0,0) -- (9.29,4.46) -- cycle    ;
-
-% Text Node
-\draw (15.2,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 3};
-% Text Node
-\draw (22.4,39) node [anchor=north west][inner sep=0.75pt]   [align=left] {n : 3};
-% Text Node
-\draw (88,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 4};
-% Text Node
-\draw (95.2,39) node [anchor=north west][inner sep=0.75pt]   [align=left] {n : 3};
-% Text Node
-\draw (94.8,53.2) node [anchor=north west][inner sep=0.75pt]   [align=left] {i \ : 0};
-% Text Node
-\draw (159.8,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 5};
-% Text Node
-\draw (167,39) node [anchor=north west][inner sep=0.75pt]   [align=left] {n : 3};
-% Text Node
-\draw (166.6,53.2) node [anchor=north west][inner sep=0.75pt]   [align=left] {i \ : 0};
-% Text Node
-\draw (230,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 4};
-% Text Node
-\draw (237.2,39) node [anchor=north west][inner sep=0.75pt]   [align=left] {n : 3};
-% Text Node
-\draw (236.8,53.6) node [anchor=north west][inner sep=0.75pt]   [align=left] {i \ : 1};
-% Text Node
-\draw (347.2,17) node [anchor=north west][inner sep=0.75pt]   [align=left] {line : 7};
-% Text Node
-\draw (354.4,39) node [anchor=north west][inner sep=0.75pt]   [align=left] {n : 3};
-% Text Node
-\draw (354,54) node [anchor=north west][inner sep=0.75pt]   [align=left] {i \ : 2};
-    \end{tikzpicture}
+    \centering
+    \includegraphics[width=0.8\linewidth]{img/stackrecording_chain.png}
   \end{minipage}
@@
 \end{figure}
+
+\begin{table}[ht]
+  \centering
+  \begin{tabular}{|c|c|c|c|c|}
+  \hline
+      \textbf{Line 3} & \textbf{Line 4} & \textbf{Line 5} & \textbf{Line 7} & \textbf{Line 8} \\ \hline
+      n=3 & n=3;i=0 & n=3;i=0 & ~ & ~ \\ \hline
+      ~ & n=3;i=1 & n=3;i=1 & ~ & ~ \\ \hline
+      ~ & n=3;i=2 & n=3;i=2 & ~ & ~ \\ \hline
+      ~ & n=3;i=3 & ~ & n=3;i=3 & return 3 \\ \hline
+  \end{tabular}
+  \label{tab:stack-recording}
+  \caption{Stack Recording Example 2}
+\end{table}
 
@@
   \begin{lstlisting}[language=Java]
-public class Debugger{
-  Debugger(){
+public class LiveProbesServer{
+  LiveProbesServer()
+    this.kaa = new KeepAliveAgent();
+  }
+
+  public void run(){
     setBreakpointOnWhileTrue();
-    launchVM(KeepeAliveAgent);
+    launchVM(kaa);
   }
-  public void loadClass(classPath, className){
-    addClassPathToDynamicClassLoader(classPath);
-    loadClass(className);
+  public void loadClass(className){
+    kaa.loadClass(className);
   }
@@
   \end{lstlisting}
-  \caption{JDI Debugger classes}
+  \caption{Live Probes Server ?}
   \label{fig:debugger-class}
@@
     \label{tab:implementation-differences}
+\end{table}
+
+\begin{table}[!ht]
+  \centering
+  \begin{tabular}{|l|l|l|}
+  \hline
+      \textbf{} & SLOC Live Probe Server & SLOC KAA \\ \hline
+      \textbf{Java JDI} & 592 (Java) & 15 (Java) \\ \hline
+      \textbf{DAP Base} & 267 (Python) & ~ \\ \hline
+      \textbf{DAP Java} & 352 (Python) & 91 (Java) \\ \hline
+      \textbf{DAP Python} & 142 (Python) & 26(Python) \\ \hline
+      \textbf{DAP Javascript} & 257 (Python) & 24(Javascript) \\ \hline
+      \textbf{DAP C} & 188 (Python) & 14(C) \\ \hline
+  \end{tabular}
+  \caption{SLOC of the Live Probe Server and Keep Alive Agent (To update after new Java changes)}
+  \label{tab:sloc}
 \end{table}
@@
   \centering
-  \begin{subfigure}[b]{0.45\textwidth}
-      \centering
   \includegraphics[width=\textwidth]{img/execute.png}
-      \caption{\centering Total execution time}
-      \label{subfig:execute-time-total}
-  \end{subfigure}
-  \hfill
-  \begin{subfigure}[b]{0.45\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{img/execute_ps.png}
-      \caption{\centering Execution time per step}
-      \label{subfig:execute-time-ps}
-  \end{subfigure}
   \caption{Execution time in seconds depending on the number of steps.}
@@
 
-Subsequently, we proceeded to evaluate the execution time concerning the number of steps in the debugger. 
-Figure \ref{subfig:execute-time-total} presents the total time taken based on the number of steps during stack recording, while Figure \ref{subfig:execute-time-ps} demonstrates the time per step as a function of the number of steps.
-Notably, the execution time exhibits a linear relationship with the number of steps in the method.
+Next, we evaluated the execution time as a function of the number of steps in the debugger. 
+Figure \ref{fig:execute-time} shows the total time taken as a function of the number of steps during stack registration.
+Note that the execution time shows a linear relationship with the number of steps in the method. 
+In fact, for each implementation we obtain a constant time for each stack frame recorded, corresponding to the time taken to obtain the state of the stack frame and the time taken for the debugger to perform a step.
+The difference in time between the different DAP languages is explained by the implementation of the debug server. 
+With C, the debug server is a simple wrapper around gdb, which has very good performance.
 

### 2023-08-01 paper/paper.tex
Clean and add sloc

@@
 
-\begin{table}[!ht]
-  \centering
-  \begin{tabular}{|l|l|l|}
-  \hline
-      \textbf{} & SLOC Live Probe Server & SLOC KAA \\ \hline
-      \textbf{Java JDI} & 592 (Java) & 15 (Java) \\ \hline
-      \textbf{DAP Base} & 267 (Python) & ~ \\ \hline
-      \textbf{DAP Java} & 352 (Python) & 91 (Java) \\ \hline
-      \textbf{DAP Python} & 142 (Python) & 26(Python) \\ \hline
-      \textbf{DAP Javascript} & 257 (Python) & 24(Javascript) \\ \hline
-      \textbf{DAP C} & 188 (Python) & 14(C) \\ \hline
-  \end{tabular}
-  \caption{SLOC of the Live Probe Server and Keep Alive Agent (To update after new Java changes)}
-  \label{tab:sloc}
-\end{table}
-
 With DAP, the compilation step is performed by calling a command line compilation tool (javac and gcc).
@@
 In C, as in Java with JDI, the method is called from the debugger console.
+
+\subsection{Implementation specifications}
+
+\begin{table}[!ht]
+  \centering
+  \noindent\setlength\tabcolsep{4pt}%
+  \begin{tabularx}{0.7\linewidth}{lc*{4}{>{\RaggedRight\arraybackslash}X}}
+      \toprule
+      Implementation & SLOC Live Probe Server & SLOC KAA \\
+      \midrule
+      Java JDI & 527 (Java) & 77 (Java) \\ 
+      DAP Base & 267 (Python) & ~ \\ 
+      DAP Python & 142 (Python) & 26(Python) \\ 
+      DAP Java & 352 (Python) & 115 (Java) \\ 
+      DAP Javascript & 257 (Python) & 24(Javascript) \\ 
+      DAP C & 188 (Python) & 14(C) \\
+      \bottomrule
+  \end{tabularx}
+  \caption{SLOC of the Live Probe Server and Keep Alive Agent}
+  \label{tab:sloc}
+\end{table}
+
+The table \ref{tab:sloc} shows the number of lines of code for the server part and the KAA for each implementation, with the language in brackets.
+The DAP Base line corresponds to the common code used by the other DAP-based implementations.
+The tests were carried out with OpenJDK 20.0.2 for Java, Python 3.11.2 for Python and Node 18.16.1 for Javascript.
+The C code was compiled using GCC 13.1.1.
+The machine used to carry out the evaluations has an AMD Ryzen 5 2500U CPU and 8Gb of RAM.
 

### 2023-08-01 JavaProbes/src/main/java/debugger/VariableMap.java

@@ -1,116 +1,0 @@
-package debugger;
-
-import org.json.JSONArray;
-import org.json.JSONObject;
-
-import org.javatuples.Pair;
-
-import java.util.ArrayList;
-import java.util.HashMap;
-import java.util.List;
-import java.util.TreeMap;
-
-public class VariableMap {
-
-    private final TreeMap<Integer, List<HashMap<String, String>>> diffMap = new TreeMap<>();
-    public final TreeMap<Integer, List<HashMap<String, String>>> valueMap = new TreeMap<>();
-
-    private final List<Pair<Integer,HashMap<String, String>>> stackFrameHistory = new ArrayList<>();
-    private Integer lastLine = null;
-    private HashMap<String, String> lastValue = null;
-
-    public VariableMap() {
-        super();
-    }
-
-    public void addStackFrame(Integer lineNumber, HashMap<String, String> map) {
-        // History of stack frames in order of execution
-        stackFrameHistory.add(new Pair<>(lineNumber, map));
-
-        // Values over time
-        if (!valueMap.containsKey(lineNumber)) {
-            valueMap.put(lineNumber, new ArrayList<>());
-        }
-        valueMap.get(lineNumber).add(map);
-
-        // Diff between values
-        if (lineNumber == -1) {
-            diffMap.put(lineNumber, new ArrayList<>());
-            diffMap.get(lineNumber).add(map);
-        }
-        else{
-            if (lastValue != null || lastLine != null) {
-                // We need to see difference between the last value and the current value
-                HashMap<String, String> diff = getDiff(lastValue, map);
-                if (diff.size() > 0) {
-                    // There is a difference, so we need to add it to the list
-                    diffMap.computeIfAbsent(lastLine, k -> new ArrayList<>());
-                    diffMap.get(lastLine).add(diff);
-                }
-            }
-            lastValue = map;
-            lastLine = lineNumber;
-        }
-    }
-
-
-    private HashMap<String, String> getDiff(HashMap<String, String> previous, HashMap<String, String> current) {
-        HashMap<String, String> diff = new HashMap<>();
-        for (String key : current.keySet()) {
-            if (!previous.containsKey(key) || !previous.get(key).equals(current.get(key))) {
-                diff.put(key, current.get(key));
-            }
-        }
-        return diff;
-    }
-
-
-    public JSONObject valueToJSON() {
-        // Convert the treemap to a dictionary, the list to a list, and the hashmap to a dictionary
-        // Then convert the dictionary to a JSON string
[…]

### 2023-08-01 JavaProbes/src/main/java/prettyprinter/PrettyPrinter.java

@@ -1,7 +1,6 @@
 package prettyprinter;
 
 import debugger.StackRecording;
-import debugger.VariableMap;
 
 import java.io.BufferedReader;
 import java.io.File;

### 2023-08-01 JavaProbes/src/main/java/prettyprinter/SourceCodeAnnotator.java

@@ -4,10 +4,6 @@ import com.github.javaparser.StaticJavaParser;
 import com.github.javaparser.ast.CompilationUnit;
 import debugger.StackFrame;
 import debugger.StackRecording;
-import debugger.VariableMap;
-
-
-import java.util.List;
 
 
 public class SourceCodeAnnotator {

### 2023-08-01 LiveFromDAP/performance/execute_performance.py

@@ -176,7 +176,7 @@ for _ in range(3):
 for i in range(33):
     print("Iteration: ", i, end="\r")
     t1 = time.time()
-    _, st = agent.execute(js_source_path, js_method, [i+1], max_steps=1000000)
+    _, st = agent.execute(js_source_path, js_method, [str(i+1)], max_steps=1000000)
     t2 = time.time()
     times["js"][len(st.stackframes)] = t2 - t1
 

### 2023-08-01 cloc.py

@@ -1,0 +1,53 @@
+import os
+import json
+
+BASE_COMMAND = "cloc --vcs git --json"
+
+report = {
+    "Java JDI":{},
+    "DAP Base":{},
+    "DAP Python":{},
+    "DAP Java":{},
+    "DAP JavaScript":{},
+    "DAP C":{},
+}
+
+def extract(root_path, files):
+    command = BASE_COMMAND + " " + " ".join([os.path.join(root_path, f) for f in files])
+    result = os.popen(command).read()
+    return json.loads(result)
+
+# Java JDI server
+root_path = "JavaProbes/src/main/java/debugger/"
+files = ["Debugger.java", "MirrorCreator.java", "ObjectInvocationRequest.java", "StackFrame.java", "StackRecording.java"]
+report["Java JDI"]["Server"] = extract(root_path, files)["Java"]["code"]
+# Java JDI KAA
+files = ["DebugAgent.java", "DynamicClassLoader.java", "DynamicClassLoaderFactory.java"]
+report["Java JDI"]["KAA"] = extract(root_path, files)["Java"]["code"]
+
+# DAP Base
+root_path = "LiveFromDAP/src/livefromdap/"
+files = ["agent/BaseLiveAgent.py", "utils/StackRecording.py"]
+report["DAP Base"]["Server"] = extract(root_path, files)["Python"]["code"]
+report["DAP Base"]["KAA"] = 0
+
+# DAP Python
+report["DAP Python"]["Server"] = extract(root_path, ["agent/PythonLiveAgent.py"])["Python"]["code"]
+report["DAP Python"]["KAA"] = extract(root_path, ["runner/py_runner.py"])["Python"]["code"]
+
+# DAP Java
+report["DAP Java"]["Server"] = extract(root_path, ["agent/JavaLiveAgent.py"])["Python"]["code"]
+files = ["runner/Runner.java", "runner/DynamicClassLoader.java", "runner/DynamicClassLoaderFactory.java"]
+report["DAP Java"]["KAA"] = extract(root_path, files)["Java"]["code"]
+
+# DAP JavaScript
+report["DAP JavaScript"]["Server"] = extract(root_path, ["agent/JavascriptLiveAgent.py"])["Python"]["code"]
+report["DAP JavaScript"]["KAA"] = extract(root_path, ["runner/js_runner.js"])["JavaScript"]["code"]
+
+# DAP C
[…]

### 2023-08-01 LiveFromDAP/src/prettyprinter/CPrettyPrinter.py
All pretty printer use tree_sitter

@@ -1,107 +1,49 @@
-from pycparser import c_ast, parse_file
+import os
+from .TreeSitterPrettyPrinter import TreeSitterPrettyPrinter
 
-from livefromdap.utils.StackRecording import StackRecording
-
-
-class CPrettyPrinter(c_ast.NodeVisitor):
-
-    def __init__(self, target_path, method):
-        self.target_path = target_path
-        self.filename = target_path.split("/")[-1]
-        self.method = method
-        self.changed_source = False
-        # Get the number of lines in the file
+class CPrettyPrinter(TreeSitterPrettyPrinter):
+    setup_function_query = """
+        (function_definition
+            (function_declarator 
+                declarator: (identifier) @fname
+                (#match? @fname "{method}")
+                parameters: (parameter_list [
+                    (parameter_declaration
+                        declarator: (identifier) @fparam
+                    )
+                    (parameter_declaration
+                        declarator: (array_declarator
+                            declarator: (identifier) @fparam
+                        ) 
+                    )]
+                )
+            ) 
+        )@fdecl
+    """
     
-    def pretty_print(self, stacktrace : StackRecording, return_value=None):
-        self.return_value = return_value
-        self.stacktrace = stacktrace
-        self.ast = parse_file(self.target_path, use_cpp=True)
-        self._visit_ast(self.ast)
-        self.ast = None
-        self.return_value = None
-        self.stacktrace = None
-        return "\n".join(self.output)
+    assignment_query = """
+        (init_declarator
+            declarator: (identifier) @varname
+        )
+        (assignment_expression
+            left: (identifier) @assignleft
+        )
+    """
     
-    def _visit_ast(self, ast):
-        with open(self.target_path, "r") as f:
-            self.lines = f.readlines()
-        self.output = ["" for _ in range(len(self.lines))]
-        self.visit(ast)
-
-    def get_value_string(self,stackframe, variable_name):
-        vtype = stackframe.get_type(variable_name)
-        value = stackframe.get_variable(variable_name)
-        if value is not None and vtype is not None and "*" in vtype:
-            return f"{vtype}[...]"
-        return value
-
+    while_query = """
+        (while_statement
+            condition : (
+                parenthesized_expression (
+                    binary_expression
+                        left : (identifier)* @whileleft
+                        right : (identifier)* @whileright
+                )
+            )
+        )
+    """
     
-    def get_variable(self, line, variable_name):
[…]

### 2023-08-01 paper/paper.tex
Add second demo to paper

@@
       \midrule
-      Java JDI & 527 (Java) & 77 (Java) \\ 
+      Java JDI & 521 (Java) & 77 (Java) \\ 
       DAP Base & 267 (Python) & ~ \\ 
@@
 \label{sec:demo-small-c}
+
+We have developed a dynamic programming environment that integrates live probes via Java, C, Python and Javascript debugging servers.
+
+Each time changes are made in the code editor, the code is sent to the server, which then checks for parseability and syntax errors. 
+In addition, if the code contains a comment beginning with "@", followed by a function call, the server attempts to create a stack recording of that function, with the parameters provided.
+
+The user interface consists of a web page that interacts with a local server. The demo features two display methods.
+
+\subsubsection{Bret Victor's style}
+\label{sec:bret-victor-style}
 \begin{figure}[htbp]
@@
 
-We have developed a dynamic programming environment that incorporates live probes through Java, C, Python, Javascript debug servers. 
-The user interface consists of a web page that interacts with a local server. 
-In Figure \ref{fig:demo}, we present a screenshot depicting a session with a binary\_search function in C. 
-On the left-hand side, there is a code editor (based on CodeMirror 5), while on the right, a panel showcases live probes.
-
-Whenever changes are made in the code editor, the code is sent to the server, which then checks for parseability and syntax errors. 
-Additionally, if the code contains a comment starting with "@", followed by a function call, the server attempts to create a stack recording of that function, along with the provided parameters. 
-This stack recording is subsequently processed through a parser to generate live probes, offering the following functionalities:
+The first mode is inspired by Bret Victor's presentation. The screenshot in Figure \ref{fig:demo} shows a session with a C binary search function. 
+On the left side there is a code editor (based on CodeMirror 5), while on the right a panel displays live probes.
+The display on the right is generated from source code and stackrecording with a tree sitter to display probes in different locations:
 
 \begin{itemize}
-  \item Displaying input arguments and return values for each function call.
-  \item Showing the value of each variable definition or assignment. In cases where the assignment is within a loop or called multiple times, all the different values are displayed.
-  \item Presenting the comparison values for each loop (for and while).
+  \item Display of input arguments and return values for each function call.
+  \item Displays the value of each variable definition or assignment. In cases where the assignment is in a loop or is called several times, all the different values are displayed.
+  \item Display comparison values for each loop (for and while).
 \end{itemize}
+
+The use of the tree-sitter is not required to display probes.
+In fact, it could simply display the variables when they are modified.
+However, using the tree sitter also allows us to be able to display the state of variables where they are used, such as in comparisons in loops.
+
+\subsubsection{Stack explorer}
+
+\begin{figure}[htbp]
+  \centering
+  \begin{subfigure}[b]{0.4\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{img/demo/stack/0.png}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.4\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{img/demo/stack/1.png}
+  \end{subfigure}
+  \vfill
+  \begin{subfigure}[b]{0.4\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{img/demo/stack/2.png}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.4\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{img/demo/stack/3.png}
+  \end{subfigure}
+  \caption{Demo of stack explorer}
+  \label{fig:demo-stack}
+\end{figure}
+
+The second mode, featured in Figure \ref{fig:demo-stack}, displays the stack recording in the form of a history.
+The screenshots show the different stack states recorded.
+The right-hand panel shows the stackframe state and a slider for exploring the different states.
+The location in the source code corresponding to the state where the stackframe has been registered is highlighted in yellow in the Codemirror editor on the left.
 

### 2023-08-01 LiveFromDAP/src/webdemo/AutoAgent.py

@@ -9,6 +9,7 @@ from livefromdap.agent.CLiveAgent import CLiveAgent
 from livefromdap.agent.JavaLiveAgent import JavaLiveAgent
 from livefromdap.agent.JavascriptLiveAgent import JavascriptLiveAgent
 from livefromdap.agent.PythonLiveAgent import PythonLiveAgent
+from livefromdap.utils.StackRecording import Stackframe
 from prettyprinter.CPrettyPrinter import CPrettyPrinter
 from pycparser import c_parser, parse_file, c_generator
 import ast as python_ast
@@ -259,6 +260,7 @@ class AutoPythonLiveAgent(AutoLiveAgent):
     def construct_result_json(self, method, output):
         return_value, stacktrace = output
         if self.raw:
+            stacktrace.last_stackframe.variables.append({"name":"return", "value":return_value})
             return json.dumps({
                 "return_value": return_value,
                 "stacktrace": stacktrace.to_json()

### 2023-08-02 JavaProbes/src/main/java/debugger/DebugAgent.java
Clean file and add type checking

@@ -13,7 +13,6 @@ public class DebugAgent {
     }
 
     public void addPath(String path) {
-        System.out.println("Adding path: " + path);
         try {
             File file = new File(path);
             if(file.exists()) {
@@ -26,7 +25,6 @@ public class DebugAgent {
     }
 
     public Class loadClass(String className) {
-        System.out.println("Loading class: " + className);
         try {
             Class clazz = classLoaderFactory.loadClass(className);
             forName(className, true, classLoaderFactory.getClassLoader()); // additional to force class preparation

### 2023-08-02 LiveFromDAP/performance/execute_performance.py

@@ -171,7 +171,7 @@ agent.load_code(js_source_path)
 
 # warm up
 for _ in range(3):
-    agent.execute(js_source_path, js_method, [1], max_steps=1000000)
+    agent.execute(js_source_path, js_method, ["1"], max_steps=1000000)
     
 for i in range(33):
     print("Iteration: ", i, end="\r")

### 2023-08-02 LiveFromDAP/performance/general_performance.py

@@ -104,22 +104,22 @@ java_args = ["new int[]{1,2,3,4,5,6}", "9"]
 
 times = {
     "c": {
-        "initialize": 0,
-        "compile": 0,
-        "load_code": 0,
-        "execute": 0
+        "initialize": 0.,
+        "compile": 0.,
+        "load_code": 0.,
+        "execute": 0.
     },
     "python": {
-        "initialize": 0,
-        "compile": 0,
-        "load_code": 0,
-        "execute": 0
+        "initialize": 0.,
+        "compile": 0.,
+        "load_code": 0.,
+        "execute": 0.
     },
     "java": {
-        "initialize": 0,
-        "compile": 0,
-        "load_code": 0,
-        "execute": 0
+        "initialize": 0.,
+        "compile": 0.,
+        "load_code": 0.,
+        "execute": 0.
     }
 }
 

### 2023-08-02 LiveFromDAP/src/livefromdap/agent/BaseLiveAgent.py

@@ -1,6 +1,7 @@
 import subprocess
 import os
 from abc import ABC, abstractmethod
+from debugpy.common.messaging import JsonIOStream
 
 from livefromdap.utils.StackRecording import StackRecording
 
@@ -50,8 +51,10 @@ class BaseLiveAgent(BaseLiveAgentInterface):
     This class should not be used directly, but should be inherited by a specific LiveAgent"""
 
     seq : int = 0
+    
+    io : JsonIOStream
 
-    def __init__(self, *args, **kwargs):
+    def __init__(self, **kwargs):
         self.debug = kwargs.get("debug", False)
         self.seq = 0
         
@@ -290,7 +293,7 @@ class BaseLiveAgent(BaseLiveAgentInterface):
         self.io.write_json(evaluate_request)
         return self.wait("response", command="evaluate")
 
-    def wait(self, type: str, event : str= None, command : str = None) -> dict:
+    def wait(self, type: str, event : str = "", command : str = "") -> dict:
         """Wait for a specific message from the debuggee
 
         Args:
@@ -305,7 +308,7 @@ class BaseLiveAgent(BaseLiveAgentInterface):
             dict: the message received
         """
         while True:
-            output = self.io.read_json()
+            output : dict = self.io.read_json() # type: ignore
             if self.debug: print(output)
             if output["type"] == "request" and output["command"] == "runInTerminal":
                 if self._handleRunInTerminal(output):

### 2023-08-02 LiveFromDAP/src/livefromdap/agent/CLiveAgent.py

@@ -25,7 +25,7 @@ class CLiveAgent(BaseLiveAgent):
             ) @funcdef
         """
         self.current_loaded_shared_libraries = None
-        self.main_thread_id = None
+        self.main_thread_id = 0
 
     def start_server(self):
         """Create a subprocess with the agent"""
@@ -170,11 +170,11 @@ class CLiveAgent(BaseLiveAgent):
         self.wait("event", event="stopped")
         end_lines = [self.get_end_line(source_file, method)]
         stackrecording = StackRecording()
-        self.initial_height = None
+        self.initial_height = -1
         i = 0
         while True:
             stackframes = self.get_stackframes(thread_id=self.main_thread_id)
-            if self.initial_height is None:
+            if self.initial_height == -1:
                 self.initial_height = len(stackframes)
             frame_id = stackframes[0]["id"]
             if stackframes[0]["name"] == "main()":

### 2023-08-02 LiveFromDAP/src/livefromdap/agent/JavaLiveAgent.py

@@ -10,7 +10,6 @@ from .JavaParams import *
 class JavaLiveAgent(BaseLiveAgent):
     def __init__(self, *args, **kwargs):
         super().__init__(*args, **kwargs)
-        self.ls_server = None
         self.ls_server_path = kwargs.get("ls_server_path", os.path.join(os.path.dirname(__file__), "..", "bin", "jdt-language-server", "bin", "jdtls"))
         self.debug_jar_path = kwargs.get("debug_jar_path", os.path.join(os.path.dirname(__file__), "..", "bin", "com.microsoft.java.debug.plugin.jar"))
         self.runner_path = kwargs.get("runner_path", os.path.abspath(os.path.join(os.path.dirname(__file__), "..", "runner")))
@@ -57,7 +56,7 @@ class JavaLiveAgent(BaseLiveAgent):
         }
         })
         while True:
-            response = self.ls_io.read_json()
+            response : dict = self.ls_io.read_json() # type: ignore
             if self.debug: print("[LanguageServer]", response)
             if "method" in response and response["method"] == "client/registerCapability" and response["params"]["registrations"][0]["method"] == "workspace/executeCommand":
                 self.ls_capabilities += response["params"]["registrations"][0]["registerOptions"]["commands"]
@@ -72,7 +71,7 @@ class JavaLiveAgent(BaseLiveAgent):
         runner_file_path = os.path.join(self.runner_path, self.runner_file)
         self.lsp_add_document(runner_file_path)
         while True:
-            response = self.ls_io.read_json()
+            response = self.ls_io.read_json() # type: ignore
             if self.debug: print("[LanguageServer]", response)
             if "method" in response and response["method"] == "client/registerCapability" and response["params"]["registrations"][0]["method"] == "workspace/executeCommand":
                 self.ls_capabilities += response["params"]["registrations"][0]["registerOptions"]["commands"]
@@ -92,7 +91,7 @@ class JavaLiveAgent(BaseLiveAgent):
                 }
             })
             while True:
-                response = self.ls_io.read_json()
+                response = self.ls_io.read_json() # type: ignore
                 if self.debug: print("[LanguageServer]",response)
                 if "id" in response and response["id"] == 30:
                     break
@@ -111,7 +110,7 @@ class JavaLiveAgent(BaseLiveAgent):
             })
         
             while True:
-                response = self.ls_io.read_json()
+                response = self.ls_io.read_json() # type: ignore
                 if self.debug: print("[LanguageServer]",response)
[…]

### 2023-08-02 LiveFromDAP/src/livefromdap/agent/JavascriptLiveAgent.py

@@ -13,7 +13,7 @@ class JavaScriptPreprocessor:
     This is done by adding module.exports = {function_name: function_name} to the end of the file
     """
     
-    def __init__(self, input_path, output_path):
+    def __init__(self, input_path : str, output_path : str):
         self.input_path = input_path
         self.output_path = output_path
         self.lang = Language(os.path.abspath(os.path.join(os.path.dirname(__file__), "..", "bin","treesitter","javascript.so")), 'javascript')
@@ -90,18 +90,18 @@ class JavascriptLiveAgent(BaseLiveAgent):
         while True:
             # check if the process returned an error
             if self.server_process.poll() is not None:
-                error = self.server_process.stderr.readline()
+                error = self.server_process.stderr.readline() # type: ignore
                 # if address already in use
                 if b"Error: listen EADDRINUSE: address already in use" in error:
                     # fidn the process that is using the port
                     p = subprocess.Popen(["lsof", "-i", ":8123"], stdout=subprocess.PIPE)
                     p.wait()
-                    lsof_output = p.stdout.readlines()
+                    lsof_output = p.stdout.readlines() # type: ignore
                     pid = lsof_output[-1].split()[1]
                     subprocess.Popen(["kill", pid])
                     self.start_server()
                     return
-            output = self.server_process.stdout.readline()
+            output = self.server_process.stdout.readline() # type: ignore
             # wait for : Debug server listening at 127.0.0.1:8123
             if b"Debug server listening" in output:
                 break
@@ -254,7 +254,7 @@ class JavascriptLiveAgent(BaseLiveAgent):
         self.configuration_done()
 
 
-    def load_code(self, path):
+    def load_code(self, path : str):
         """For this agent, loading code is loading a class in the runner"""
         jspp = JavaScriptPreprocessor(path, path)
         self._entry_line[path] = jspp.functions
@@ -262,13 +262,13 @@ class JavascriptLiveAgent(BaseLiveAgent):
         self.evaluate(f'loadFile("{path}")', frame_id)
 
 
-    def execute(self, path, method, args, max_steps=100):
+    def execute(self, path : str, method : str, args : list[str], max_steps : int =100) -> tuple[str, StackRecording]:
         """Execute a method with the given arguments"""
         if not path in self._entry_line:
             self.load_code(path)
             
         if not method in self._entry_line[path]:
-            return
+            return "Error", StackRecording()
         
[…]

### 2023-08-02 LiveFromDAP/src/prettyprinter/TreeSitterPrettyPrinter.py

@@ -2,6 +2,9 @@ from tree_sitter import Language,Parser
 from livefromdap.utils.StackRecording import StackRecording
 
 class TreeSitterPrettyPrinter():
+    tree_sitter_path : str
+    tree_sitter_name : str
+    
     def __init__(self, file_path, method_name):
         self.file_path = file_path
         self.method_name = method_name
@@ -37,7 +40,7 @@ class TreeSitterPrettyPrinter():
     
     
     def setup_function(self, formatter : dict):
-        function_params_query = self.lang.query(self.setup_function_query.format(**formatter))
+        function_params_query = self.lang.query(self.setup_function_query.format(**formatter)) # type: ignore
         captures = function_params_query.captures(self.ast.root_node)
         params = []
         for capture, capture_type in captures:
@@ -53,7 +56,7 @@ class TreeSitterPrettyPrinter():
         self.output[self.function_start[0]] = fdef_string
         
     def visit_assignment(self):
-        vardecl_query = self.lang.query(self.assignment_query)
+        vardecl_query = self.lang.query(self.assignment_query) # type: ignore
         
         captures = vardecl_query.captures(self.ast.root_node)
         for capture, capture_type in captures:
@@ -68,7 +71,7 @@ class TreeSitterPrettyPrinter():
                 self.output[capture.start_point[0]] += f"{varname} = {value}"
     
     def visit_while(self):
-        while_decl_query = self.lang.query(self.while_query)
+        while_decl_query = self.lang.query(self.while_query) # type: ignore
         captures = while_decl_query.captures(self.ast.root_node)
         variables = {}
         for capture, capture_type in captures:

### 2023-08-02 LiveFromDAP/src/webdemo/AutoAgent.py

@@ -25,9 +25,8 @@ class ThreadWithReturnValue(Thread):
         self._return = None
 
     def run(self):
-        if self._target is not None:
-            self._return = self._target(*self._args,
-                                                **self._kwargs)
+        if self._target is not None: # type: ignore
+            self._return = self._target(*self._args, **self._kwargs) # type: ignore
     def join(self, *args):
         Thread.join(self, *args)
         return self._return
@@ -61,7 +60,6 @@ class AutoCLiveAgent(AutoLiveAgent):
         self.agent = CLiveAgent(debug=False)
         self.agent.start_server()
         self.agent.initialize()
-        print("Agent initialized")
         self.buzy = False
         self.c_parser = c_parser.CParser()
         self.c_generator = c_generator.CGenerator()
@@ -117,7 +115,6 @@ class AutoCLiveAgent(AutoLiveAgent):
     def construct_result_json(self, method, output):
         return_value, stacktrace = output
         if self.raw:
-            print("Sending raw output")
             return json.dumps({
                 "return_value": return_value,
                 "stacktrace": stacktrace.to_json()
@@ -304,7 +301,6 @@ class AutoJavascriptLiveAgent(AutoLiveAgent):
         changed = False
         try:
             ast = self.parser.parse(bytes(code, "utf8"))
-            print("parsing ", ast.root_node.sexp())
             # Query to find ERROR or MISSING nodes
             query = self.lang.query("""
             (ERROR) @error
@@ -346,7 +342,6 @@ class AutoJavascriptLiveAgent(AutoLiveAgent):
         return output
 
     def execute(self, method, args):
-        print("Executing", args)
         output = self.agent.execute(self.source_path, method, args)
         if output is None:
             return ""

### 2023-08-02 LiveFromDAP/src/webdemo/main.py

@@ -164,15 +164,14 @@ def stackt(language):
 
 @socketio.on('disconnect')
 def on_disconnect():
-    if request.sid in sessions_to_sid and (session_id:=sessions_to_sid[request.sid]) in sessions:
+    if request.sid in sessions_to_sid and (session_id:=sessions_to_sid[request.sid]) in sessions: # type: ignore
         sessions[session_id].agent.agent.stop_server()
         del sessions[session_id]
 
 @socketio.on('join')
 def on_join(data):
-    print("join")
     session_id = data.get("session_id")
-    sessions_to_sid[request.sid] = session_id
+    sessions_to_sid[request.sid] = session_id # type: ignore
     language = data.get("language")
     join_room(session_id)
     if not session_id in sessions:

### 2023-08-02 paper/paper.tex
add performance JDI and clean image for demo 2

@@
   \centering
-  \begin{subfigure}[b]{0.4\textwidth}
+  \begin{subfigure}[b]{0.45\textwidth}
       \centering
@@
   \hfill
-  \begin{subfigure}[b]{0.4\textwidth}
+  \begin{subfigure}[b]{0.45\textwidth}
     \centering
@@
   \vfill
-  \begin{subfigure}[b]{0.4\textwidth}
+  \par\bigskip
+  \begin{subfigure}[b]{0.45\textwidth}
     \centering
@@
   \centering
-  \includegraphics[width=\textwidth]{img/execute.png}
-  \caption{Execution time in seconds depending on the number of steps.}
+  \includegraphics[width=0.5\textwidth]{img/execute.png}
+  \caption{Execution time with DAP in seconds depending on the number of steps.}
   \label{fig:execute-time}
+\end{figure}
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.5\textwidth]{img/execute_java_jdi.png}
+  \caption{Comparison of execution time with DAP and JDI in seconds depending on the number of steps.}
+  \label{fig:execute-java-jdi}
 \end{figure}
@@
 With C, the debug server is a simple wrapper around gdb, which has very good performance.
+Figure \ref{fig:execute-java-jdi} shows Java execution times with DAP and JDI. 
+The y-axis is in logarithmic scale. 
+There is a significant difference in performance between the two implementations, despite the fact that the DAP server implementation in Java also uses JDI.
+We have not found the cause of this difference in performance. % TODO : find it
 
@@
 
+\begin{table}
 \begin{enumerate}
@@
 \end{enumerate}
+\caption{Steps of the binary search scenario.}
+\label{table:scenario-bs}
+\end{table}
 

### 2023-08-02 LiveFromDAP/src/livefromdap/agent/BaseLiveAgent.py

@@ -314,8 +314,8 @@ class BaseLiveAgent(BaseLiveAgentInterface):
                 if self._handleRunInTerminal(output):
                     continue
             if output["type"] == type:
-                if event is None or output["event"] == event:
-                    if command is None or output["command"] == command:
+                if event == "" or output["event"] == event:
+                    if command == "" or output["command"] == command:
                         return output
             if output["type"] == "event" and output["event"] == "terminated":
                 raise DebuggeeTerminatedError()

### 2023-08-04 paper/paper.tex
tweaks

@@
 
-\begin{table}[ht]
+\begin{table}[t]
   \centering
-  \begin{tabular}{|c|c|c|c|c|}
-  \hline
-      \textbf{Line 3} & \textbf{Line 4} & \textbf{Line 5} & \textbf{Line 7} & \textbf{Line 8} \\ \hline
-      n=3 & n=3;i=0 & n=3;i=0 & ~ & ~ \\ \hline
-      ~ & n=3;i=1 & n=3;i=1 & ~ & ~ \\ \hline
-      ~ & n=3;i=2 & n=3;i=2 & ~ & ~ \\ \hline
-      ~ & n=3;i=3 & ~ & n=3;i=3 & return 3 \\ \hline
-  \end{tabular}
-  \label{tab:stack-recording}
-  \caption{Stack Recording Example 2}
+  \includegraphics[width=\textwidth]{img/matrix.pdf}
+  \caption{Tracking variable values across time and space}
 \end{table}
+
+% \begin{table}[ht]
+%   \centering
+%   \begin{tabular}{|c|c|c|c|c|}
+%   \hline
+%       \textbf{Line 3} & \textbf{Line 4} & \textbf{Line 5} & \textbf{Line 7} & \textbf{Line 8} \\ \hline
+%       n=3 & n=3;i=0 & n=3;i=0 & ~ & ~ \\ \hline
+%       ~ & n=3;i=1 & n=3;i=1 & ~ & ~ \\ \hline
+%       ~ & n=3;i=2 & n=3;i=2 & ~ & ~ \\ \hline
+%       ~ & n=3;i=3 & ~ & n=3;i=3 & return 3 \\ \hline
+%   \end{tabular}
+%   \label{tab:stack-recording}
+%   \caption{Stack Recording Example 2}
+% \end{table}
 
@@
 \begin{table}[!ht]
-  \centering
+  %\centering
   \noindent\setlength\tabcolsep{4pt}%
-  \begin{tabularx}{0.7\linewidth}{lc*{4}{>{\RaggedRight\arraybackslash}X}}
+  \begin{tabular}{lrlrl}
       \toprule
-      Implementation & SLOC Live Probe Server & SLOC KAA \\
+      Implementation & \multicolumn{2}{l}{\#SLOC Probe Server} & \multicolumn{2}{l}{\#SLOC Keep-Alive Agent} \\
       \midrule
-      Java JDI & 521 (Java) & 77 (Java) \\ 
-      DAP Base & 267 (Python) & ~ \\ 
-      DAP Python & 142 (Python) & 26(Python) \\ 
-      DAP Java & 352 (Python) & 115 (Java) \\ 
-      DAP Javascript & 257 (Python) & 24(Javascript) \\ 
-      DAP C & 188 (Python) & 14(C) \\
+      Java JDI & 521 & Java & 77 & Java \\ 
+      DAP Base & 267  & Python & --- & \\ 
+      DAP Python & 142 & Python & 26 & Python \\ 
+      DAP Java & 352 & Python & 115 & Java \\ 
+      DAP Javascript & 257 & Python & 24 & Javascript \\ 
+      DAP C & 188 & Python & 14 & C \\
       \bottomrule
-  \end{tabularx}
-  \caption{SLOC of the Live Probe Server and Keep Alive Agent}
+  \end{tabular}
+  \caption{Size of the language specific components in SLOC and their implementation languages.}
   \label{tab:sloc}

### 2023-08-08 paper/paper.tex
title
Abstract

@@
 \usepackage{subcaption} % For subfigures
-\usepackage{array, caption, tabularx,  ragged2e,  booktabs}
+\usepackage{array, caption, tabularx,  ragged2e,  booktabs, xspace}
 
@@
 %%
+
+\def\coolname{\textsc{LiveRec}\xspace}
 
@@
 
-\title{Language-Parametric Techniques for Live Probes}
+\title{\textsc{LiveRec}: Framing Debug Protocols to Hotwire Probes}
 %\subtitle{Preparing Articles for Programming}% optional
@@
   In the first part of his 2012 presentation ``Inventing on Principle''~\cite{InventingOnPrinciple}, Bret Victor gives a demo of live code editor for Javascript which shows the dynamic history of values of variables in real time.
-This form of live programming has become known as ``live probes''~\cite{MakingMethodsLive,UsableLiveProgramming,ExampleBasedGraalVM}.
-Using live probes the programmer has a permanent and continuous insight into the dynamic evolution of function or method variables, thus improving feedback and developer experience.
+This form of live programming has become known as ``probes''~\cite{MakingMethodsLive,UsableLiveProgramming,ExampleBasedGraalVM}.
+Probes provide the programmer with permanent and continuous insight into the dynamic evolution of function or method variables, thus improving feedback and developer experience.
 
@@
 Later work~\cite{ExampleBasedGraalVM} provides an implementation approach, but this requires a programming language to be implemented on top of the GraalVM runtime~\cite{GraalVM}.
-In this paper we investigate generic engineering strategies that can be applied in the context of many programming languages, without requiring dedicated compiler or runtime support.
+In this paper we present \coolname, a generic approach for implementing probes which can be applied in the context of many programming languages, without requiring dedicated compiler or runtime support.
 
@@
   \textit{Approach.}
-  In this paper we explore how debugger protocols can be exploited to implement live probes.
+  \coolname is based on reusing existing debug protocols to implement probes.
 Methods or functions are compiled after every code change and executed inside the debugger.
@@
   \textit{Knowledge.}
-  It turns out that debugger protocols are rich enough for implementing live probes.
-Step-wise execution, hot code swapping and stack frame inspection provide the right granularity and sufficient information to realize live probes, without modifying compilers or language runtimes.
+  It turns out that debug protocols are rich enough for implementing live probes.
+Step-wise execution, hot code swapping, and stack frame inspection provide the right granularity and sufficient information to realize live probes, without modifying compilers or language runtimes.
 Furthermore, it turns out that the recently proposed Debugger Adapter Protocol (DAP)~\cite{DAP} provides an even more generic approach of implementing live probes, but, in some cases, at the cost of a significant performance penalty.
@@
   \textit{Grounding.}
-  We have implemented live probes using stack recording natively for Java through the Java Debug Interface (JDI)~\cite{JDI}, and through the DAP for Java, Python, C, and Javascript, all with minimum glue code.
-We evaluate the run-time performance of all four live probes prototypes, decomposed into: compile-after-change, hotswap, single step overhead, and stack recording overhead.
-Our results show that live probes on top of native debugger APIs are performant enough for interactive use.
+  We have applied \coolname to implement probes using stack recording natively for Java through the Java Debug Interface (JDI)~\cite{JDI}, and through the DAP for Java, Python, C, and Javascript, all requiring just reasonable amounts glue code.
+We evaluate the run-time performance of all four probes prototypes, decomposed into: compile-after-change, hotswap, single step overhead, and stack recording overhead.
+Our results show that live probes on top of native debug APIs are performant enough for interactive use.
 In the case of DAP, it depends on characteristics of the programming language.
@@
   \textit{Importance.} Live programming improves the programmer experience by providing immediate feedback about a program's execution and eliminating disruptive edit-compile-restart sequences.
-Live probes are one way to shorten the programmer feedback loop at the level of functions and methods.
-Although live probes are not new, and have been implemented in (prototype) systems, our approach of building live probes on top of existing and generic debugger protocols opens up live probes for a host of mainstream programming languages, with minimum effort.
+Probes are one way to shorten the programmer feedback loop at the level of functions and methods.
+Although probes are not new, and have been implemented in (prototype) systems, \coolname's approach of building live probes on top of existing and generic debug protocols opens up probes for a host of mainstream programming languages, with minimum effort.
 \end{abstract}

### 2023-08-08 paper/paper.tex
Intro

@@
 While all the prototypes demonstrated by Victor are impressive, for the purpose of this paper we zoom in on one of them: a live editor for Javascript with a side pane showing the history of values of local function variables during an execution (see Figure~\ref{FIG:bret} below).
-Although not mentioned by Victor himself, this live programming features has become known as ``live probes''~\cite{UsableLiveProgramming,ExampleBasedGraalVM}.
+Although not mentioned by Victor himself, this live programming features has become known as ``probes''~\cite{UsableLiveProgramming,ExampleBasedGraalVM}.
 
-The question left unanswered in Victor's presentation is: how to build such live probes?
+The question left unanswered in Victor's presentation is: how to build such probes?
 While there has been research showcasing implementation approaches~\cite{ExampleBasedGraalVM}, they depend on a specific runtime system (e.g., GraalVM), or employ language specific ``hacks''~\cite{LiveLiterals}, and hence cannot be easily transferred to other languages.
-So a more specific question is: how to  engineer live probes \textit{generically}, without dedicated compiler or runtime support?
+So a more specific question is: how to  engineer probes \textit{generically}, without dedicated compiler or runtime support?
 
-In this paper we present an approach to implementing live probes reusing standard debugger protocols and APIs.
-Most language support an API to facilitate the implementation of debuggers and other run-time tools (e.g., profilers, tracers, etc.).
+In this paper we present \coolname, an approach to implement probes by reusing standard debugger protocols and APIs.
+Most languages support an API to facilitate the implementation of debuggers and other run-time tools (e.g., profilers, tracers, etc.).
 It turns out that such facilities are sufficiently powerful for implementing live probes.
-The key to the approach is to (re)execute a function or method inside the debugger, step by step, and at each step record a snapshot of the current stack-frame.
+The key to the approach of \coolname is to (re)execute a function or method inside the debugger, step by step, and at each step record a snapshot of the current stack-frame.
 After the method has finished executing, the resulting stack recording is communicated back to the IDE and visualized alongside the code, matching stack recording entries to the source code using source location information.
 
-We first detail how to implement live probes using a native debugger interface, the Java Debugger Interface (JDI).
+We first detail how to implement probes using a native debugger interface, the Java Debug Interface (JDI).
 Then we generalize the approach to employ the recently proposed Debugger Adapter Protocol (DAP).
-This implementation is showcased with live probe implementations in Java (again), C, Python, and Javascript. 
-As a result, live probes can be supported in any programming language that provides an implementation of the DAP, with only a limited amount of configuration code.
+This implementation is showcased with probe implementations in Java (again), C, Python, and Javascript. 
+As a result, probes can be supported in any programming language that provides an implementation of the DAP, with only a limited amount of configuration code.
 
-Since live programming (and hence live probes) is all about immediate feedback, it is important that the performance overhead of our implementation strategy does not impede practical use.
-We therefore provide a detail analysis and measurement of the performance overhead of (re)compilation, hot-swapping, step-wise execution, and stack recording.
-To assess the overall usability of the approach, we furthermore decompose a full editing scenario (derived from Victor's presentation) in $X$ logical steps, replay it on every live probe implementation, and measure the performance of each step.
+Since live programming (and hence probes) is all about immediate feedback, it is important that the performance overhead of \coolname's strategy does not impede practical use.
+We therefore provide a detailed analysis and measurement of the performance overhead of (re)compilation, hot-swapping, step-wise execution, and stack recording.
+To assess the overall usability of the approach, we furthermore decompose a full editing scenario (derived from Victor's presentation) in $19$ logical steps, replay it on every  probe implementation, and measure the performance of each edit step.
 
@@
 \begin{itemize}
-  \item We analyze the requirements for implementing live probes (Section~\ref{sec:problem-overview})
-  \item We introduce \textit{stack recording} as a technique to obtain the required information through standard debugger protocols (Section~\ref{sec:stack-recording})
-  \item We implement live probes through stack recording on top of the Java Debugger Interface (JDI) (Section~\ref{sec:live-probes-java})
-  \item We then generalize the approach via the Debugger Adapter Protocol (DAP) (Section~\ref{sec:generalizing-live-probes})
-   \item We evaluate the feasibility of the approach by implementing live probes for Java, Python, C and Javascript on top the DAP, investigate the performance overhead of the individual steps, and reproduce an editing scenario to assess practicality (Section~\ref{sec:evaluation})
+  \item We analyze the requirements for implementing probes (Section~\ref{sec:problem-overview})
+  \item We provide an overview of \coolname and introduce \textit{stack recording} as a technique to obtain the required information through standard debug protocols (Section~\ref{sec:stack-recording})
+  \item We apply \coolname on top of the Java Debugger Interface (JDI) (Section~\ref{sec:live-probes-java})
+  \item We then show how to generalize \coolname in the context of the Debugger Adapter Protocol (DAP) (Section~\ref{sec:generalizing-live-probes})
+   \item We evaluate the feasibility of \coolname by implementing probes for Java, Python, C and Javascript on top the DAP, investigate the performance overhead of the individual steps, and reproduce an editing scenario to assess practicality (Section~\ref{sec:evaluation}).
 \end{itemize}
+We conclude the paper with a discussion of limitations and related work, and present concluding remarks with directions for further research.
 

### 2023-08-10 paper/paper.tex
add profiling

--- 2023-08-04 paper/paper.tex
@@
 The y-axis is in logarithmic scale. 
-There is a significant difference in performance between the two implementations, despite the fact that the DAP server implementation in Java also uses JDI.
-We have not found the cause of this difference in performance. % TODO : find it
+There is a significant difference in performance between the two implementations, despite the fact that the DAP server implementation in Java also uses JDWP.
+By profiling the execution of the java debug server, we found two causes for this performance gap:
+\begin{itemize}
+  \item The first is due to the "evaluate" request in the DAP protocol.
+To run the function and define the arguments in the DAP implementation, we use "evaluate" requests.
+These requests require an additional compilation step in the debug server compared with the version using JDI.
+In the JDI version, the arguments are actually copied into the debugger directly from the server coded in Java.
+To improve the performance of our DAP implementation, if the arguments are not changed between two executions, they are kept in the KAA.
+  \item The second cause is the native performance of the debugger server.
+Our method automates debugger procedures, and the observed performance of these procedures is sufficient for human use.
+For example, obtaining the stacktrace takes an average of 40 ms.
+When used by a person, this delay is less than 0.1 seconds and is considered instantaneous.
+However, when we perform stack recording, this request will be executed a significant number of times (depending on the size of the code).
+\end{itemize}
 

### 2023-08-10 paper/paper.tex
Rewrite JDI paragraph

@@
 
-We have developed a debugger capable of stack recording using JDI, a debugging interface for Java. 
-A simplified pseudo code of this class is presented in Figure \ref{fig:debugger-class}.
-The class includes a method to initialise the JVM (line 2), a method to search or reload a class in the VM (line 5) and a method to generate a stack recording of a method (line 10).
+To enable stack recording, we developed a debugger leveraging the Java Debug Interface (JDI) library. 
+A simplified pseudo-code representation of this debugger is displayed in Figure \ref{fig:debugger-class}. 
+The class contains a method for starting the JVM (at line 2), a process to load or reload a class within the VM (at line 5), and a procedure for generating a method stack record (at line 10).
+Although in practice, the implementation involves multiple interacting threads, these details have been omitted from this overview in favor of clarity and simplicity.
 
-For initialisation, a JVM containing the instrumentation provided by JDI is launched on the KeepAliveAgent class \ref{fig:java-keep-alive-agent} and a breakpoint is set on the infinite loop in the \code{main} function (line 9).
-This breakpoint is used to keep the debugger alive in a wait state, which is necessary to use the JDI primitives \code{mirrorOf}, \code{newInstance} and \code{invokeMethod}.
+Initialization involves the launch of a JVM that features JDI instrumentation, utilizing the KeepAliveAgent class (in Figure \ref{fig:java-keep-alive-agent}). 
+Concurrently, a breakpoint is established at the infinite loop within the \code{main} function, visible at line 9. 
+This breakpoint's function is to maintain the debugger in a state of suspension.
+This waiting state is requisite for the use of JDI primitives such as \code{mirrorOf}, \code{newInstance}, and \code{invokeMethod}.
 
-To load a class, we use a dynamic ClassLoader which allows us to add class paths and load a class during execution. 
-When the code changes, the class is reloaded into the KeepAliveAgent. 
-This method has one limitation: it only works if the class signature has not changed, and only the contents of the methods have changed. 
-If the structure has changed, the KeepAliveAgent must be restarted. 
-This limitation can be overcome with certain JVM implementations\cite{KABANOV201151}.
+Class loading is achieved using a dynamic ClassLoader, which gives us the ability to append class paths and load a class during execution. 
+Modifications to the code cause the class to be reloaded into the KeepAliveAgent. 
+However, there is a caveat to this procedure - it only works if the class signature remains consistent and only the method content is changed. 
+Any changes to the structure will require a restart of the KeepAliveAgent. 
+It is to be noted that specific JVM implementations can bypass this limitation \cite{}.
 
-To execute the target method (line 10 in Debugger\ref{fig:debugger-class}), the arguments are created in the client JVM, then passed to the debugger JVM using the \code{mirrorOf} and \code{newInstance} methods of the reflection API (line 11).
-A breakpoint is then placed at the input of the target method (line 12).
-The target method is invoked using \code{invokeMethod}. 
-Stack recording is then performed using the algorithm described in the previous paragraph. 
-As long as the method is not finished, the debugger registers the stackframe and takes a step in the debuggee.
+To execute a function (line 10 in Debugger class, Figure \ref{fig:debugger-class}), the debugger creates the values associated with the arguments locally. 
+It then transmits these values to the JVM containing the KAA using the \code{mirrorOf} and \code{newInstance} methods of JDI's Reflection API (line 11). 
+The debugger then sets a breakpoint at the target method's input (line 12) and launches the target method using \code{invokeMethod}. 
+The debugger then registers the stack using the algorithm described in the previous paragraph.
+Until the method is completed, the debugger registers the stack frame and takes a step forward in debugging.
 

### 2023-08-10 paper/paper.tex
clear installation

@@
       \midrule
-      Java JDI & 521 & Java & 77 & Java \\ 
-      DAP Base & 267  & Python & --- & \\ 
+      Java JDI & 521 & Java & 75 & Java \\ 
+      DAP Base & 270  & Python & --- & \\ 
       DAP Python & 142 & Python & 26 & Python \\ 
-      DAP Java & 352 & Python & 115 & Java \\ 
-      DAP Javascript & 257 & Python & 24 & Javascript \\ 
-      DAP C & 188 & Python & 14 & C \\
+      DAP Java & 360 & Python & 114 & Java \\ 
+      DAP Javascript & 256 & Python & 24 & Javascript \\ 
+      DAP C & 181 & Python & 14 & C \\
       \bottomrule
@@
 The tests were carried out with OpenJDK 20.0.2 for Java, Python 3.11.2 for Python and Node 18.16.1 for Javascript.
-The C code was compiled using GCC 13.1.1.
+The C code was compiled using GCC 13.1.1 and GDB 13.2 was used for stack recording.
 The machine used to carry out the evaluations has an AMD Ryzen 5 2500U CPU and 8Gb of RAM.

### 2023-08-10 LiveFromDAP/install.sh

@@ -1,6 +1,19 @@
 #!/bin/bash
 # This script will download and install the debug servers for LiveFromDAP
 
+
+echo "[Java Debug] Checking for Java"
+if ! java --version &> /dev/null; then
+    echo "You need to install java"
+    exit 1
+fi
+
+echo "[Java Debug] Checking for Javac"
+if ! javac --version &> /dev/null; then
+    echo "You need to install javac"
+    exit 1
+fi
+
 # JDT LS
 echo "[JDT LS] Checking for JDT LS..."
 # check if src/livefromdap/bin/jdt-language-server directory exists
@@ -39,13 +52,24 @@ else
 fi
 
 # GDB Debug
-echo "[C Debug] Checking for gdb"
+echo "[C Debug] Checking for GCC"
+# check if node is installed
+# if node -v exit with 0 then it is installed
+if ! gcc -v &> /dev/null; then
+    echo "You need to install GCC"
+    exit 1
+fi
+
+# GDB Debug
+echo "[C Debug] Checking for GDB"
 # check if node is installed
 # if node -v exit with 0 then it is installed
 if ! gdb -v &> /dev/null; then
     echo "You need to install GDB"
     exit 1
 fi
+
+
 
 # C Debug
 echo "[C Debug] Checking for C debug server..."
@@ -92,6 +116,14 @@ echo "[Node Debug] Checking for Node"
 # if node -v exit with 0 then it is installed
 if ! node -v &> /dev/null; then
     echo "You need to install Node.js"
+    exit 1
+fi
+
+echo "[Node Debug] Checking for NPM"
+# check if node is installed
+# if node -v exit with 0 then it is installed
+if ! npm --version &> /dev/null; then
+    echo "You need to install NPM"
     exit 1
 fi
 
@@ -148,7 +180,7 @@ fi
 # Make runner files executable
 echo "[Runner] Making runner files executable..."
 cd src/livefromdap
-make runner
+make all
 chmod +x runner/*
 cd ../..
 

### 2023-08-10 LiveFromDAP/src/livefromdap/agent/BaseLiveAgent.py

@@ -1,6 +1,7 @@
 import subprocess
 import os
 from abc import ABC, abstractmethod
+from typing import Any
 from debugpy.common.messaging import JsonIOStream
 
 from livefromdap.utils.StackRecording import StackRecording
@@ -51,10 +52,11 @@ class BaseLiveAgent(BaseLiveAgentInterface):
     This class should not be used directly, but should be inherited by a specific LiveAgent"""
 
     seq : int = 0
+    debug : bool = False
     
     io : JsonIOStream
 
-    def __init__(self, **kwargs):
+    def __init__(self, **kwargs : Any):
         self.debug = kwargs.get("debug", False)
         self.seq = 0
         

### 2023-08-10 LiveFromDAP/src/livefromdap/agent/CLiveAgent.py

@@ -23,6 +23,7 @@ class CLiveAgent(BaseLiveAgent):
                     (#match? @fname "{function_name}")
                 )
             ) @funcdef
+            (return_statement) @return
         """
         self.current_loaded_shared_libraries = None
         self.main_thread_id = 0
@@ -120,16 +121,22 @@ class CLiveAgent(BaseLiveAgent):
         compilation = subprocess.run(self.compile_command.format(target_input=input_file, target_output=output_file), shell=True, check=True)
         return compilation.returncode
     
-    def get_end_line(self, path, function_name):
+    def get_end_line(self, path : str, function_name : str) -> list[int]:
         with open(path, 'r') as file:
             code = file.read()
         query = self.lang.query(self.end_function_query_string.format(function_name=function_name))
         captures = query.captures(self.parser.parse(bytes(code, "utf8")).root_node)
-        captures = [c for c in captures if c[1] == "funcdef"]
+        captures = [c for c in captures if c[1] == "funcdef" or c[1] == "return"]
         if len(captures) == 0:
             raise Exception(f"Function {function_name} not found")
         else:
-            return captures[0][0].end_point[0]+1
+            res = []
+            for capture in captures:
+                if capture[1] == "funcdef":
+                    res.append(capture[0].end_point[0] + 1)
+                if capture[1] == "return":
+                    res.append(capture[0].start_point[0] + 1)
+            return res
     
     def setup_runner_breakpoint(self):
         self.set_breakpoint(self.runner_path, [14])
@@ -162,14 +169,15 @@ class CLiveAgent(BaseLiveAgent):
         recorded_stackframe = Stackframe(line, column, height, variables)
         stackrecording.add_stackframe(recorded_stackframe)
     
-    def execute(self, source_file, method, args, max_steps=300):
+    def execute(self, source_file : str, method : str, args : list[str], max_steps : int=300) -> tuple[str, StackRecording]:
         self.set_function_breakpoint([method])
         command = f"-exec call {method}({','.join(args)})"
         frame_id = self.get_stackframes(thread_id=self.main_thread_id)[0]["id"]
         self.evaluate(command, frame_id)
         self.wait("event", event="stopped")
-        end_lines = [self.get_end_line(source_file, method)]
+        end_lines : list[int] = self.get_end_line(source_file, method)
         stackrecording = StackRecording()
+        return_value = ""
         self.initial_height = -1
         i = 0
         while True:
@@ -182,7 +190,7 @@ class CLiveAgent(BaseLiveAgent):
[…]

### 2023-08-10 LiveFromDAP/src/livefromdap/agent/JavaLiveAgent.py

@@ -1,23 +1,28 @@
 import os
 import subprocess
-from debugpy.common.messaging import JsonIOStream
-from debugpy.common import sockets
+from typing import Any
+from debugpy.common.messaging import JsonIOStream # type: ignore
+from debugpy.common import sockets # type: ignore
 
 from livefromdap.utils.StackRecording import Stackframe, StackRecording
 from .BaseLiveAgent import BaseLiveAgent
 from .JavaParams import *
 
 class JavaLiveAgent(BaseLiveAgent):
-    def __init__(self, *args, **kwargs):
-        super().__init__(*args, **kwargs)
-        self.ls_server_path = kwargs.get("ls_server_path", os.path.join(os.path.dirname(__file__), "..", "bin", "jdt-language-server", "bin", "jdtls"))
+    def __init__(self, *args : str, **kwargs : str): 
+        super().__init__(*args, **kwargs) # type: ignore
+        self.ls_server_path : str = kwargs.get("ls_server_path", os.path.join(os.path.dirname(__file__), "..", "bin", "jdt-language-server", "bin", "jdtls"))
+        assert os.path.exists(self.ls_server_path), f"Could not find language server at {self.ls_server_path}"
         self.debug_jar_path = kwargs.get("debug_jar_path", os.path.join(os.path.dirname(__file__), "..", "bin", "com.microsoft.java.debug.plugin.jar"))
+        assert os.path.exists(self.debug_jar_path), f"Could not find debug jar at {self.debug_jar_path}"
         self.runner_path = kwargs.get("runner_path", os.path.abspath(os.path.join(os.path.dirname(__file__), "..", "runner")))
         self.runner_file = kwargs.get("runner_file", "Runner.java")
-        self.project_name = None
+        assert os.path.exists(os.path.join(self.runner_path, self.runner_file)), f"Could not find runner file at {os.path.join(self.runner_path, self.runner_file)}"
+        self.project_name = ""
         self.method_loaded= None
         self.loaded_classes = {}
         self.loaded_class_paths = []
+        self.last_args = None
 
     def start_ls_server(self):
         self.ls_server = subprocess.Popen(
@@ -26,7 +31,7 @@ class JavaLiveAgent(BaseLiveAgent):
             stdout=subprocess.PIPE,
             stderr=subprocess.PIPE
         )
-        self.ls_capabilities = []
+        self.ls_capabilities : list[str] = []
         self.ls_io = JsonIOStream.from_process(self.ls_server)
         self.ls_io.write_json({
[…]

### 2023-08-10 LiveFromDAP/src/livefromdap/runner/Runner.java

@@ -53,7 +53,7 @@ public class Runner {
         Runner runner = new Runner();
         Object result;
         while (true) {
-            if (runner.method != null && runner.args != null) {
+            if (runner.method != null) {
                 try {
                     // check if static
                     if (Modifier.isStatic(runner.method.getModifiers())) {
@@ -81,7 +81,6 @@ public class Runner {
                         }
                         result = runner.method.invoke(classInstance, runner.args);
                     }
-                    runner.args = null;
                     runner.method = null;
                 } catch (Exception e) {
                     e.printStackTrace();

### 2023-08-10 LiveFromDAP/src/webdemo/AutoAgent.py

@@ -13,7 +13,7 @@ from livefromdap.utils.StackRecording import Stackframe
 from prettyprinter.CPrettyPrinter import CPrettyPrinter
 from pycparser import c_parser, parse_file, c_generator
 import ast as python_ast
-import javalang
+import javalang # type: ignore
 from prettyprinter.JavaPrettyPrinter import JavaPrettyPrinter
 from prettyprinter.JavascriptPrettyPrinter import JavascriptPrettyPrinter
 from prettyprinter.PythonPrettyPrinter import PythonPrettyPrinter
@@ -115,6 +115,7 @@ class AutoCLiveAgent(AutoLiveAgent):
     def construct_result_json(self, method, output):
         return_value, stacktrace = output
         if self.raw:
+            stacktrace.last_stackframe.variables.append({"name":"return", "value":return_value})
             return json.dumps({
                 "return_value": return_value,
                 "stacktrace": stacktrace.to_json()
@@ -190,6 +191,7 @@ class AutoJavaLiveAgent(AutoLiveAgent):
     def construct_result_json(self, method, output):
         return_value, stacktrace = output
         if self.raw:
+            stacktrace.last_stackframe.variables.append({"name":"return", "value":return_value})
             return json.dumps({
                 "return_value": return_value,
                 "stacktrace": stacktrace.to_json()
@@ -333,6 +335,7 @@ class AutoJavascriptLiveAgent(AutoLiveAgent):
     def construct_result_json(self, method, output):
         return_value, stacktrace = output
         if self.raw:
+            stacktrace.last_stackframe.variables.append({"name":"return", "value":return_value})
             return json.dumps({
                 "return_value": return_value,
                 "stacktrace": stacktrace.to_json()

### 2023-08-10 LiveFromDAP/src/webdemo/main.py
add script to launch demo fast

@@ -194,9 +194,9 @@ def handle_json(json_msg):
         return
     sessions[session_id].queue.put(json_msg)
     
-        
+def run():
+    socketio.run(app)
 
 
 if __name__ == '__main__':
-    #no reload because of the global variables
-    socketio.run(app)
+    run()

### 2023-08-10 paper/paper.tex
add dap request

@@
 This backend offers an interface common to all three languages, which includes methods for starting and initialising the debugging server, loading and reloading code in the debugger, and executing a method while performing stack recording.
+Table \ref{tab:dap-req} contains a list of the requests from the DAP that we have used to implement the interface.
+
+\begin{table}[h]
+  \centering
+  \noindent\setlength\tabcolsep{4pt}%
+  \begin{tabularx}{\linewidth}{ll*{2}{>{\RaggedRight\arraybackslash}X}}
+    \toprule
+    Request & Usage \\ [0.5ex]
+    \midrule
+    Initialize and Launch & Initialize the DAP server and launch the debuggee \\ 
+    \addlinespace
+    SetBreakpoints & Set breakpoints at location in source code \\ 
+    \addlinespace
+    StackTrace & Get the state of the current stacktrace \\ 
+    \addlinespace
+    Scops and Variables & Get the current scopes and variables in it \\ 
+    \addlinespace
+    Evaluate & Evaluate expression in the debugee \\ 
+    \addlinespace
+    Next and Continue & Request to step or continue in the debuggee \\ 
+    \addlinespace
+    \bottomrule
+  \end{tabularx}
+  \caption{DAP Requests}
+  \label{tab:dap-req}
+\end{table}
 
@@
 For Javascript, a line is added at the end of the imported file to export all the functions in the file as a module, which is then loaded into the KeepAliveAgent with \code{require}.
-For Java, as with the JDI version, a DynamicClassLoader is used. For C, the code is loaded into shared libraries that can be added and reloaded at runtime.
+For Java, as with the JDI version, a Dynamic ClassLoader is used. For C, the code is loaded into shared libraries that can be added and reloaded at runtime.
 

### 2023-09-01 paper/paper.tex
Reviewing sections 1 and 2. Added comments and suggestions in the text, below the sentences.
Inventing a new figure -- tracking variable values across time and space
Updated figure -- tracking variable values across time and space

@@
 }
+
+\def\iC#1{\lstinline[language=C]{#1}\xspace}
 
@@
 Furthermore, it turns out that the recently proposed Debugger Adapter Protocol (DAP)~\cite{DAP} provides an even more generic approach of implementing live probes, but, in some cases, at the cost of a significant performance penalty.
+%vRozen: "It turns out" -- heavy 'it' at start of sentence -- repetitive 2x. Suggest rephrasing: This paper establishes that ...  or simply: We show that ...
 
@@
 In the case of DAP, it depends on characteristics of the programming language.
+  %vRozen: probes prototypes -- heavy alliteration -- try saying it three times fast. Suggested phrasing just 'prototypes'.
+  %vRozen: 'hotswap', 'hot swap' or 'hot-swap' ?
 
@@
 By erasing the distinction between a program on the one hand, and its execution on the other, programmers may enjoy a more fluid experience, with immediate dynamic feedback after every code change.
+%vRozen: 'may enjoy' ? -- Perhaps remove 'may' here.
 
@@
 Although not mentioned by Victor himself, this live programming features has become known as ``probes''~\cite{UsableLiveProgramming,ExampleBasedGraalVM}.
+%vRozen: Singular / plural mismatch. 'these features have' or 'this feature has'. I get that 'probes' invites using the plural.
+%vRozen: I'm missing a definition of probes.
+Probes are a specific live programming mechanism that adds immediate feedback to imperative languages, reminiscent of voltage probes, for relating the inputs of functions to their outputs.
 
@@
 So a more specific question is: how to  engineer probes \textit{generically}, without dedicated compiler or runtime support?
+%vRozen: ``hacks'' sounds a bit derogative. Let's stay friends with the reviewers :D
 
@@
 After the method has finished executing, the resulting stack recording is communicated back to the IDE and visualized alongside the code, matching stack recording entries to the source code using source location information.
+%vRozen: 'It turns out' -- heavy it.
+%vRozen: 'stack-frame' or 'stack frame' ?
+%vRozen: Long final sentence -- information-dense -- high mental load -- consider breaking into parts.
 
@@
 To assess the overall usability of the approach, we furthermore decompose a full editing scenario (derived from Victor's presentation) in $19$ logical steps, replay it on every  probe implementation, and measure the performance of each edit step.
+%Rozen: Repeats 'hence' -- rephrase: therefore
 
@@
 We conclude the paper with a discussion of limitations and related work, and present concluding remarks with directions for further research. 
+%vRozen: To me the paper's scientific contributions and its structure are not the same thing.
+The paper contributes 'stack recording', a generic approach for probes.
+The paper is structured as follows: ...
 
@@
 The basic concept of live probes is illustrated in Figure~\ref{FIG:bret}, showing two panes. 
+%vRozen: 'live probes' -- \coolname?
 The left pane shows an editor containing the source code of a possible implementation of binary search in Javascript. 
+%vRozen: The left pane shows an editor containing JavaScript sources of an example implementation of binary search.
 The right pane shows the values of variables matched up to the source line they were last updated.
+%vRozen: The right pane shows the values the variables have at the source lines they were last updated.
 The top two rows specify example input data that the function is executed on. 
-The consecutive values of variables that are updated in a loop (e.g.
-\lstinline{low}, \lstinline{high}, \lstinline{mid} and \lstinline{value}) are displayed adjacently.
+%vRozen: Bit passive -- sentence doesn't sing -- can't put my finger on it.
+The consecutive values of variables that are updated in a loop (e.g. \lstinline{low}, \lstinline{high}, \lstinline{mid} and \lstinline{value}) are displayed adjacently. 
+%vRozen: Bit passive -- sentence doesn't sing -- can't put my finger on it.
 So for \lstinline{low}, the history of values is 0, 3, and 5.
 Note that the update of \lstinline{high} in the else-if branch (apparently) never executed, so there is no (redundant) value display in the right pane.
+%vRozen: Long sentence, rephrase, remove 'so'.
 Finally, the last line shows the final return value of the function: apparently the \lstinline{key} was not found in the \lstinline[language=c]{array}.
@@
 What makes this \textit{live} is that while the programmer is editing the program on the left (or the input arguments to the function), the value display on the right continuously updates so that it is apparent what is going on at run time.
+%vRozen: Rephrase.
+What is 'this' ?
+Rephrase.
+This scenario exemplifies live programming.
+Specifically, the values displayed on the right continuously update while the programmer is editing the program on the left.
 
@@
 \end{itemize}
+%vRozen: consider starting the above sentences with a verb in simple present. Remember requirements have actors. Is that the case here? Maybe they are technical challenges.
 
@@
 This program consists of a while-true-loop, halted on a breakpoint, and will be the verhicle for executing methods, without having to restart the debugger upon each request.
+%vRozen: Long sentences.
+Consider breaking into parts.
+For instance: The basic intuition behind our approach is to meet these requirements as follows.
+Each editor starts a custom debugger, implemented against some debugger API.
+Instead of debugging the program directly, a this debugger runs a driver program called the Keep Alive Agent (KAA).
+...
 
@@
 To keep stack recording local to the current method, the debugger never steps \textit{into} a call to another method.
+%vRozen: 'hotswap', 'hot swap' or 'hot-swap' ?
+%vRozen: 'some kind' -- vague -- a particular ?
 
@@
 These snapshots are stored in a data structure, called \textit{stack recording}, which supports displaying the values in the IDE, for instance, like in Figure~\ref{FIG:bret}, inline~\cite{LiveLiterals,ExampleBasedGraalVM}, or using some dedicated UI affordance.
+%vRozen: Long sentence. Consider breaking into parts. ", which supports displaying the values in the IDE." 
+%vRozen: For example, Figure ...
+%vRozen 'some dedicated' -- vague -- replace 'some' by 'a'
 
@@
 \section{Using the Debugger ?}
+%vRozen: todo fixme heading title
 
@@
 \end{figure}
+%vRozen: todo fixme remove 'placeholder'
+%vRozen: perhaps a tikz image here
 
 A live environment must be able to react to two different events: a change in the code or a change in the test/input data. 
+%vRozen: 'live environment' -- 'live programming environment'
 If the code changes, we need to re-execute the code, and in the case of a compiled language, we need to compile the new code first. 
+%vRozen: who is 'we' ? Consider replacing 'we' by 'a programmer' Whenever a programmer changes the code, ...
 If the data changes, we need to be able to re-execute the programme with the new data. 
+%vRozen: Likewise, the program needs to re-execute immediately when the programmer adjusts the inputs.
 In addition, it is necessary to keep compilation and execution times low enough to maintain an interactive experience with the user.
+%vRozen: 'the user' -- suggest 'the programmer' or 'programmers'
 
@@
   \item The IDE or editor communicates with a server. The IDE sends changes to the code and function parameters, and the server sends back the stack trace of the method.
+%vRozen: 'or'. The editor is part of an IDE that ...
   \item The server initialises and communicates with the debugger to apply the changes received from the IDE, and generates a stack recording when the method is executed.
+%vRozen: US English: Initializes
+%vRozen: The server initializes the debugger, and continually applies the changes it receives from the IDE, generating a stack recording when a method is executed.
   \item The debugger runs an intermediate program, a keep alive agent. This keeps the debugger alive to reduce initialisation times.
+%vRozen: Note -- The acronym Keep Alive Agent (KAA) has been introduced already.
+%vRozen: The debugger runs the Keep Alive Agent, an intermediate program for reducing initialization times, which keeps the debugger alive between executions.
 \end{itemize}
@@
 \label{sec:stack-recording}
+%vRozen: fixme dangling label
 
@@
 \end{figure}
+%vRozen. The layout seems a bit off.
 
@@
 \end{table}
+
+\begin{figure}
+  \centering
+  \large  
+  \hspace{8pt} \color{darkgray} \textsf{\textbf{time}}  \color{black} $\rightarrow$ \raggedright
+  \vspace{2pt}
+
+  \rotatebox{90}{$\leftarrow$ \color{darkgray} \textsf{\textbf{space} \color{black}}}
+  \small
+  \begin{tabularx}{0.94\textwidth}{ l @{\hspace{4pt}} l @{\hspace{4pt}} X @{\hspace{4pt}} X @{\hspace{4pt}} X @{\hspace{4pt}} X @{\hspace{4pt}} X @{\hspace{4pt}} X @{\hspace{4pt}} X @{\hspace{4pt}} X @{\hspace{4pt}} X @{\hspace{4pt}} X @{\hspace{4pt}} l}
+  \toprule  
+  &
+  &
+  \textbf{$t_1$}&
+  \textbf{$t_2$}&
+  \textbf{$t_3$}&
+  \textbf{$t_4$}&
+  \textbf{$t_5$}&
+  \textbf{$t_6$}&
+  \textbf{$t_7$}&
+  \textbf{$t_8$}&
+  \textbf{$t_9$}\\
+  \midrule
+  &
+  &
+  \iC{n; i} &
+  \iC{n; i} &
+  \iC{n; i} &
+  \iC{n; i} &
+  \iC{n; i} &
+  \iC{n; i} &
+  \iC{n; i} &
+  \iC{n; i} &
+  \iC{n; i} \\
+  \midrule
+  1 & \iC{//foo(3)}\\
+  2 & \iC{int foo(int n)\{} \\
+  3 & \hspace{8pt}~\iC{int i = 0;}
+  & \iC{3; _} \\  
+  4 & \hspace{8pt}~\iC{while(i < n)\{} 
+  &
+  & \iC{3; 0}
+  &
+  & \iC{3; 1}
+  & 
+  & \iC{3; 2}\\
+  5 & \hspace{16pt}~\iC{i++;}
+  &
+  &
+  & \iC{3; 0}; 
+  &
+  & \iC{3; 1}\\
+  6 & \hspace{8pt}~\iC{\}}\\
+  7 & \hspace{8pt}~\iC{return i;}
+  &
+  &
+  &
+  &
+  &
+  &
+  &
+  &
+  \iC{3; 3}\\
+  8 & \iC{\}}
+    &
+  &
+  &
+  &
+  &
+  &
+  &
+  &
+  &
+  \iC{return 3}\\
+  \bottomrule
+  \end{tabularx}
+  \caption{Tracking variable values across time and space}
+\end{figure}
 

### 2023-09-15 paper/paper.tex
Continued reviewing. Added comments until Section 6.
Continued reviewing. Added comments until Section 7.
Parked ideas about the relationship between stack recordings, cause-and-effect chains and different live programming features.

@@
 %vRozen: who is 'we' ? Consider replacing 'we' by 'a programmer' Whenever a programmer changes the code, ...
+%vRozen: alternatively, 'we' could be software language engineers who are implementing probes for language X.
 If the data changes, we need to be able to re-execute the programme with the new data. 
@@
 %vRozen. The layout seems a bit off.
+%vRozen. I propose merging Figure 3 and Table 1 into a single figure, the tabularx version below.
 
@@
 
+%vRozen: Perhaps removing the time/space arrows would be better. it seems a bit cluttered
+%vRozen: Perhaps we can still add boxes around the stack recordings to make it more similar to figure 4. It gives the feeling of having data structures, not just values.
 \begin{figure}
@@
 In order to generate the data needed to probe a variable, we need to be able to retrieve the state of the variable during execution. 
+%vRozen: Repeats 'need' -- consider rephrasing
 This information is contained in the stackframe at the time of execution. 
+%vRozen: passive -- 'contains' is more active
 However, we need to add spatial and temporal information to this: we need to associate each stackframe state with the location in the code at which it was retrieved, and we also need to know the order in which these stackframes were retrieved.
+%vRozen: Repeats 'we need' -- Long sentence. Consider breaking the sentence into parts.
+%vRozen: Probing a variable requires both spacial and temporal information. ...
 
@@
 The stack record is presented as a chain of recorded stack frames, to which information has been added about the source code's location and height in the stack.
+%vRozen: Repeats 'stack record' and repeats the part of 'we need' in different words.
 This representation allows us to maintain a link between the spatial location (the reference to the source code) and the temporal location (the order of these stackframes) of the execution.
@@
 \end{figure}
+%vRozen: This is a UML sequence diagram.
+%vRozen: The loop may use an activation frame.
+%vRozen: The objects (or component instances) on the top of life lines appear in rectangular boxes.
 
 Figure \ref{fig:stack-recording-impl} shows how a stack recording is recorded.
+%vRozen: Repetitive 'stack recording' and 'recorded', consider rephrasing.
 To obtain the different states of the stackframe during the execution of a method, we place a breakpoint at the start of the function.
+%vRozen: method or function? -- pick one, at the beginning. Perhaps state that in this paper we can use these terms interchangeably.
 Once this has been reached, the method is executed step by step, using the debugger's step instruction.
+%vRozen: Repeats: step step step. Passive sentence. One the execution reaches this break point, it continues in a stepwise manner.
 Between each step, the state of the stackframe is recorded along with the source code's location information.
@@
 To enable stack recording, we developed a debugger leveraging the Java Debug Interface (JDI) library. 
+%vRozen: Past tense. Please use present tense whenever possible for simplicity. Use past perfect if necessary, stressing a solution has addressed a challenge.
 A simplified pseudo-code representation of this debugger is displayed in Figure \ref{fig:debugger-class}. 
+%vRozen. Passive. Suggest a more active sentence. Figure .. displays a simplified pseudo-code representation of this debugger.
 The class contains a method for starting the JVM (at line 2), a process to load or reload a class within the VM (at line 5), and a procedure for generating a method stack record (at line 10). 
 Although in practice, the implementation involves multiple interacting threads, these details have been omitted from this overview in favor of clarity and simplicity.
+%vRozen: High mental load for the reader. Suggest adding some signals to the text. For conciseness we omit <<thing>>. In practice, yada yada <<additional complexity>>.
 
@@
 Class loading is achieved using a dynamic ClassLoader, which gives us the ability to append class paths and load a class during execution. 
+%vRozen: passive sentence. A ClassLoader loads classes dynamically during the execution. By appending the class path ...
+
 Modifications to the code cause the class to be reloaded into the KeepAliveAgent.
+%vRozen: KAA, Keep Alive Agent or KeepAliveAgent?  
 However, there is a caveat to this procedure - it only works if the class signature remains consistent and only the method content is changed. 
@@
 It is to be noted that specific JVM implementations can bypass this limitation \cite{}.
+%vRozen: heavy it. missing citation
 
 To execute a function (line 10 in Debugger class, Figure \ref{fig:debugger-class}), the debugger creates the values associated with the arguments locally.
+%vRozen: Whenever a probing session executes a function (line 10), the debugger creates ...
+
 It then transmits these values to the JVM containing the KAA using the \code{mirrorOf} and \code{newInstance} methods of JDI's Reflection API (line 11).
+%vRozen: KAA, Keep Alive Agent or KeepAliveAgent? 
 The debugger then sets a breakpoint at the target method's input (line 12) and launches the target method using \code{invokeMethod}. 
@@
 Until the method is completed, the debugger registers the stack frame and takes a step forward in debugging.
+%vRozen: The above paragraph starts with a 'heavy it' followed by 'the debugger' several times. Perhaps use a "First, the debugger ... Next, it ... Finally, ..." temporal structure.
 
@@
 It is compatible with various editors, including VS Code, and provides a unified interface for all programming languages.
+%vRozen: heavy it. Suggest rephrasing: DAP is ...
 
 By exploiting this protocol, we have created a new language parametric backend, which we have implemented for the C, Python, Java and Javascript programming languages.
+%vRozen: Only here it becomes clear 'us' is we the authors -- creators of generic language technology.
+%vRozen: Long sentence.
 These languages are chosen to cover both compiled and interpreted languages. 
+%vRozen: Passive. We choose, ...
 This backend offers an interface common to all three languages, which includes methods for starting and initialising the debugging server, loading and reloading code in the debugger, and executing a method while performing stack recording.
+%vRozen: Long sentence, consider breaking into parts. 
 Table \ref{tab:dap-req} contains a list of the requests from the DAP that we have used to implement the interface.
@@
 These methods allow us to carry out stack recording independently of the chosen language and facilitate the future implementation of live programming interfaces. 
+%vRozen: Short paragraph. Consider merging.
 
 The implementation for each language includes a keep-alive agent and code to communicate with the debugger and keep-alive agent to provide the interface functionality.
+%vRozen: Keep Alive Agent, KAA, KeepAliveAgent or keep-alive agent?
 These methods depend on both the implementation of the debugging server and the keep-alive agent. 
+%vRozen: These methods --> Specific details 
 The differences between the implementations are shown in the Table \ref{tab:implementation-differences}.
+%vRozen: Passive. Figure .. highlights the differences between the implementations.
+%vRozen: Some connective sentence. We explain ...
 
@@
 \subsection{Implementation specifications}
+%vRozen: todo fixme -- chose different heading
 
@@
   \noindent\setlength\tabcolsep{4pt}%
-  \begin{tabular}{lrlrl}
+  \begin{tabularx}{\textwidth}{X r X r X}
       \toprule
@@
       \bottomrule
-  \end{tabular}
+  \end{tabularx}
   \caption{Size of the language specific components in SLOC and their implementation languages.}
@@
 The machine used to carry out the evaluations has an AMD Ryzen 5 2500U CPU and 8Gb of RAM.
+%vRozen: Avoid complicated past perfect tense 'was compiled', e.g., 'has been compiled' or 'we have compiled'.
 
@@
 We have developed a dynamic programming environment that integrates live probes via Java, C, Python and Javascript debugging servers. 
+%vRozen: Unnecessary switch between perfect tense and simple present tense.
 
 Each time changes are made in the code editor, the code is sent to the server, which then checks for parseability and syntax errors.
+%vRozen: 'parseability' ? The server parses the code and checks for syntax errors.
 In addition, if the code contains a comment beginning with "@", followed by a function call, the server attempts to create a stack recording of that function, with the parameters provided. 
+%vRozen: Long sentence. Consider breaking into parts.
+%vRozen: Repeats 'the server'. Perhaps use 'it'.
 
 The user interface consists of a web page that interacts with a local server. The demo features two display methods. 
+%vRozen: Short paragraph. Consider merging.
 
@@
 The first mode is inspired by Bret Victor's presentation. The screenshot in Figure \ref{fig:demo} shows a session with a C binary search function. 
+%vRozen: Figure X depicts screenshot of a session that executes a binary search function written in C.
 On the left side there is a code editor (based on CodeMirror 5), while on the right a panel displays live probes. 
+%vRozen: Consider rephrasing the part after the comma, specifically 'while' and 'live probes'.
 The display on the right is generated from source code and stackrecording with a tree sitter to display probes in different locations: 
+%vRozen: Long sentence. Consider breakin into parts. 'stack recording'
 
@@
 \end{itemize}
+%vRozen: repeats display(s). Consider saying it once, before the points.
 
@@
 However, using the tree sitter also allows us to be able to display the state of variables where they are used, such as in comparisons in loops.
+%vRozen. The significance of tree-sitter (or tree sitter?) has not yet been established. What is it and why is it needed?
 
@@
 The location in the source code corresponding to the state where the stackframe has been registered is highlighted in yellow in the Codemirror editor on the left.
+%vRozen: stack frame or stackframe.
 
@@
 In the context of a live programming environment, it is essential to have short response times after user interactions. 
+%vRozen: Short response times are essential for live programming environments.
 We therefore measured the performance of our approach in response to various questions:
@@
 \end{itemize}
+%vRozen: it helps to be precise on what is recorded, e.g., time in ms. How is 'performance' defined -- the average duration of each entire case?
+%vRozen: My advise is to make the 'protocol' or 'steps' more precise, such that it because straightforward to reproduce and compare the results. 
+%vRozen: Which programs are assessed? Is this all about the binary search function or is it about more programs?
 
 \subsubsection{How long does it take to compile and to load the code into the debugger, depending on the size of the programme?}
+%vRozen: todo fixme Replace heading by a descriptive title.
 
@@
 The summarized results can be found in Figure \ref{fig:compileload}.
+%vRozen: Paragrap is in past tense.
 
@@
 \subsubsection{How long does it need to run the program, depending on the number of steps to make the stack recording ?}
+%vRozen: todo fixme Replace heading by a descriptive title.
 
@@
 Next, we evaluated the execution time as a function of the number of steps in the debugger. 
+%vRozen: Past tense.
 Figure \ref{fig:execute-time} shows the total time taken as a function of the number of steps during stack registration.
@@
 In fact, for each implementation we obtain a constant time for each stack frame recorded, corresponding to the time taken to obtain the state of the stack frame and the time taken for the debugger to perform a step.
+%vRozen: Long sentence.
 The difference in time between the different DAP languages is explained by the implementation of the debug server. 
@@
 \end{itemize}
+%vRozen: Consider using paragraphs instead of bullet points.
 
 \subsubsection{What is the average performance during a live programming session?}
+%vRozen: todo fixme Replace heading by a descriptive title.
 
@@
 To test the performance of our approach under real conditions, we recreated the live programming scenario presented in the video by bret victor. 
+%vRozen: I would expect to see this earlier, where i commented on the 'protocol'.
+The structure could have a more clear division between describing the method, taking measurements and analyzing the results.
 The scenario describes the programming of a function performing a binary search on an array of characters.
@@
 
+%vRozen: Parking some ideas here about the relationship between stack recordings, cause-and-effect chains and different live programming features.
+It may be this fits better in related work or future work than the discussion.
+The LiveRec approach introduces the minimal data structure, the stack recording, for supporting probes.
+This live programming feature is limited to analyzing functions and methods.
+Of course, other liveness features exist that require a more elaborate structure and debugging instrumentation.
+In particular, omniscient debugging (also known as back in time debugging), supports exploratory programming with what-if scenarios through record, replay, undo and redo functionality.
+The modification of running programs requires an explicit representation of change that can augment and integrate with the stack recording for supporting these features.
+For instance, the Cascade framework represents change as cause-and-effect chains (based on edit operations) that can update both a live program's syntax and its run-time state.
+Future work can further investigate how the generic debugging instrumentation can be extended for supporting additional live programming features.
+
 While our approach to implement live probes on top of off-the-shelf debuggers provides a viable strategy to add live probes to many mainstream languages, some limitations and open questions remain.

### 2023-09-28 paper/paper.tex
minor comment
Section 3.1

@@
 
-\section{Using the Debugger ?}
-%vRozen: todo fixme heading title
-
-\subsection{Architecture}
+\section{Framing Debug Protocols}
+
+\subsection{High-level Architecture}
 
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{img/architecture.png}
-  \caption{General architecture (placeholder)}
+  \includegraphics[width=0.9\linewidth]{img/architecture.png}
+  \caption{High-level architecture}
   \label{fig:architecture}
 \end{figure}
-%vRozen: todo fixme remove 'placeholder'
-%vRozen: perhaps a tikz image here
-
-A live environment must be able to react to two different events: a change in the code or a change in the test/input data. 
-%vRozen: 'live environment' -- 'live programming environment'
-If the code changes, we need to re-execute the code, and in the case of a compiled language, we need to compile the new code first. 
-%vRozen: who is 'we' ? Consider replacing 'we' by 'a programmer' Whenever a programmer changes the code, ...
-%vRozen: alternatively, 'we' could be software language engineers who are implementing probes for language X.
-If the data changes, we need to be able to re-execute the programme with the new data. 
-%vRozen: Likewise, the program needs to re-execute immediately when the programmer adjusts the inputs.
-In addition, it is necessary to keep compilation and execution times low enough to maintain an interactive experience with the user.
-%vRozen: 'the user' -- suggest 'the programmer' or 'programmers'
-
-To meet these constraints, we propose the architecture shown in Figure \ref{fig:architecture} :
-\begin{itemize}
-  \item The IDE or editor communicates with a server. The IDE sends changes to the code and function parameters, and the server sends back the stack trace of the method.
-%vRozen: 'or'. The editor is part of an IDE that ...
-  \item The server initialises and communicates with the debugger to apply the changes received from the IDE, and generates a stack recording when the method is executed.
-%vRozen: US English: Initializes
-%vRozen: The server initializes the debugger, and continually applies the changes it receives from the IDE, generating a stack recording when a method is executed.
-  \item The debugger runs an intermediate program, a keep alive agent. This keeps the debugger alive to reduce initialisation times.
-%vRozen: Note -- The acronym Keep Alive Agent (KAA) has been introduced already.
-%vRozen: The debugger runs the Keep Alive Agent, an intermediate program for reducing initialization times, which keeps the debugger alive between executions.
-\end{itemize}
-
-\label{sec:stack-recording}
-%vRozen: fixme dangling label
+
+A programming environment with probes must be able to react to two different events: a change in the code or a change in the input data to the method. 
+If the code changes, it needs to be reexecuted, possible preceded by a compilation step in the case of compiled languages.
+Likewise, if one or more input values change, the probed method needs to be reexecuted, but in this case no recompilation is required. 
+Both compilation (if any) and reexecution of the probed method needs to be fast enough to avoid noticable delays in updating the display values of the method's variables.
+
+The high-level architecture of \coolname is shown in Figure \ref{fig:architecture}.
+The IDE is used by the programmer to edit source code.
+Since the programmer's focus (indicated by the cursor position) is always inside a single method\footnote{Of course, the cursor can be outside any method, but this is irrelevant to probes.}.
+If a method has been annotated with input values for its parameters (indicating that the programmer wants to probe the method), the code and the input values are sent to the Live Probes Server after every edit.
+
+The server communicates with a (custom) Debugger implementation through a debug protocol, such as JDI or DAP. 
+This debugger component has been initialized with a dummy program, called the Keep Alive Agent (KAA).
+The code for a probed method is dynamically loaded into the KAA, so that the method can be invoked through API calls of the debug protocol.
+By hotswapping the code into the KAA we avoid a significant performance penalty of restarting the Debugger itself every time a change is received by the Live Probes Server.
+
+During execution of the probed method, the Live Probes Server collects all intermediate stack frames in a stack recording, which is sent back to the IDE for display to the programmer.
 

### 2023-09-28 paper/paper.tex
Figures

@@
 \section{Framing Debug Protocols}
+\label{sec:stack-recording}
 
@@
     \begin{lstlisting}[language=C]
-//foo(3)
+//@foo(3)
 int foo(int n) {
@@
 
-\begin{table}[t]
-  \centering
-  \includegraphics[width=\textwidth]{img/matrix.pdf}
-  \caption{Tracking variable values across time and space}
-\end{table}
+% \begin{table}[t]
+%   \centering
+%   \includegraphics[width=\textwidth]{img/matrix.pdf}
+%   \caption{Tracking variable values across time and space}
+% \end{table}
 
@@
   \centering
-  \large  
-  \hspace{8pt} \color{darkgray} \textsf{\textbf{time}}  \color{black} $\rightarrow$ \raggedright
-  \vspace{2pt}
-
-  \rotatebox{90}{$\leftarrow$ \color{darkgray} \textsf{\textbf{space} \color{black}}}
   \small
@@
   \toprule  
-  &
+  & \textsf{\textbf{time}} $\rightarrow$
   &
@@
   \midrule
-  &
-  &
-  \iC{n; i} &
-  \iC{n; i} &
-  \iC{n; i} &
-  \iC{n; i} &
-  \iC{n; i} &
-  \iC{n; i} &
-  \iC{n; i} &
-  \iC{n; i} &
-  \iC{n; i} \\
-  \midrule
   1 & \iC{//foo(3)}\\
@@
   3 & \hspace{8pt}~\iC{int i = 0;}
-  & \iC{3; _} \\  
+  & \iC{n=3} \\  
   4 & \hspace{8pt}~\iC{while(i < n)\{} 
   &
-  & \iC{3; 0}
-  &
-  & \iC{3; 1}
-  & 
-  & \iC{3; 2}\\
+  & \iC{n=3; i=0}
+  &
+  & \iC{n=3; i=1}
+  & 
+  & \iC{n=3; i=2}\\
   5 & \hspace{16pt}~\iC{i++;}
@@
   &
-  & \iC{3; 0}; 
-  &
-  & \iC{3; 1}\\
+  & \iC{n=3; i=0}; 
+  &
+  & \iC{n=3; i=1}\\
   6 & \hspace{8pt}~\iC{\}}\\
@@
   &
-  \iC{3; 3}\\
+  \iC{n=3; n=3}\\
   8 & \iC{\}}
@@
   \end{tabularx}
-  \caption{Tracking variable values across time and space}
+  \caption{Tracking variable values across time and space (= source code)}
 \end{figure}

### 2023-09-28 paper/paper.tex
Change some figures

--- 2023-09-15 paper/paper.tex
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{img/architecture.png}
-  \caption{General architecture (placeholder)}
+  \includegraphics[width=0.8\linewidth]{img/architecture_2.png}
+  \caption{General architecture}
   \label{fig:architecture}
@@
   \centering
-  \includegraphics[width=0.6\linewidth]{img/stackrecording_impl.png}
-  \caption{Record Stackrecording (placeholder)}
+  \includegraphics[width=0.8\linewidth]{img/uml_sequence_record.png}
+  \caption{Record Stackrecording}
   \label{fig:stack-recording-impl}
 \end{figure}
+
+\begin{comment} %Mermaid
+sequenceDiagram
+    participant A as Live Probes server
+    participant B as Debugger (KAA)
+    A->>B: Execute method(args)
+    B->>A: Breakpoint at method entry
+    loop Until method end
+        A->>B: request stackframe
+        B->>A: stackframe
+        A->>B: request step
+        B->>A: hit breakpoint
+    end
+\end{comment}
 %vRozen: This is a UML sequence diagram.

### 2025-02-13 paper/paper.tex

--- 2023-09-28 paper/paper.tex
@@
   \centering
-  \includegraphics[width=0.9\linewidth]{img/architecture.png}
-  \caption{High-level architecture}
+  \includegraphics[width=0.8\linewidth]{img/architecture_2.png}
+  \caption{General architecture}
   \label{fig:architecture}
